### 2017-05-11 abstract.tex
First commit

\begin{abstract}
%\boldmath
This paper aims to show that knowing the core concepts related to a given parallel architecture is necessary to write correct code, regardless of the parallel programming paradigm used.
Programmers unaware of architecture concepts, such as beginners and students, often write parallel code that is slower than their sequential versions. It is also easy to write code that produces incorrect answers under specific conditions, which are hard to detect and correct.\\
The increasing popularization of multi-core architectures motivates the implementation of parallel programming frameworks and tools, such as OpenMP, that aim to lower the difficulty of parallel programming. OpenMP uses compilation directives, or \textit{pragmas}, to reduce the number of lines that the programmer needs to write. However, the programmer still has to know when and how to use each of these directives. 
The documentation and available tutorials for OpenMP give the idea that using compilation directives for parallel programming is easy. In this paper we show that this is not always the case by analysing a set of corrections of OpenMP programs made by students of a graduate course in Parallel and Distributed Computing, at University of São Paulo.
Several incorrect examples of OpenMP \textit{pragmas} were found in tutorials and official documents available in the Internet. The idea that OpenMP is easy to use can lead to superficial efforts in teaching fundamental parallel programming concepts. This can in its turn lead to code that does not develop the full potential of OpenMP, and could also crash inexplicably due to very specific and hard-to-detect conditions. Our main contribution is showing how important it is to teach core architecture and parallel programming concepts properly, even when you have powerful tools such as OpenMP available.
% Institution 2 = University of São Paulo
\end{abstract}

### 2017-05-11 acknowledgments.tex
First commit

\section*{Acknowledgments}
\label{sec:acknowledgments}

% Blind review. 
% Procurar no texto por Institution 2 = University of São Paulo.
% Retornar o texto abaixo com o agradecimento para as fundações.

The authors would like to thank Students who participated in this exercise, CNPq, CAPES (BEX 3401/15-4, CAPES-COFECUB Project No. 828/15), São Paulo Research Foundation (FAPESP) (procs. \#2012/23300-7) for the fellowships and the Araucária Foundation, Department of Science, Technology and High Education of State Government of Paraná (SETI-PR) and State Government of Paraná for the financial support.

% The authors would like to thank Students who participated in this exercise and financial support agencies.

### 2017-05-11 conclusions.tex
First commit

\section{Conclusions and next steps}
\label{sec:conclusions}

This paper's results reiterate that, even though \texttt{OpenMP} can be easy to use, programmers still need to know the details about parallel computing architectures and platforms. Teaching and learning the fundamental concepts about parallel programming is, and will continue to be, necessary to write accurate and high-performance code.

The experimental results show that although Internet tutorials transmit the idea of ease-of-use of the compilation directives approach, the core concepts of parallelism are required to write good parallel programs. Knowing the underlying platform and architectural details, and what effects are caused by directives and their interactions is also required.

Using \texttt{OpenMP} for trivial programs is easy, but for more advanced applications it is necessary to know how to correctly use locks and barriers, avoid deadlocks, and distribute the work evenly between threads. The programmer also has to be cautious when scheduling and accessing memory. To do all this correctly it is necessary to learn the fundamental concepts underlying parallel programming.

The main goal of parallel programming is writing programs that are efficient, portable and correct, leveraging contemporary hardware platforms~\cite{Asanovic2009}. These goals are still hard to reach even when using high-level approaches such as \texttt{OpenMP}.

It is important to teach Computer Science students how to program parallel applications from the beginning. Instead of teaching sequential programming first and parallel programming much later in the graduation course, Computer Science courses should start with \texttt{OpenMP} or other high-level tools, but not before providing solid fundamentals in parallel computing.

Considering the students' answers, a positive outcome of this experiment is that the students have learned that they should not rely on copying code from the Internet, because they can contain errors. Students also learned that they should test their code thoroughly, as errors in parallel programs are not easily detectable.

Future work will repeat the experiment in other instances of the graduate course, improving the experiment data set. Using a more systematic approach, future work will look for possible errors in more tutorials available on the Internet, and explore the difficulties faced by beginner and expert developers of parallel applications.

Finally, considering the results presented in this paper and the fact that even those who are teaching make mistakes in their tutorials and classes, more attention has to be devoted to teaching parallel and concurrent programming, its techniques, concepts and tools.

### 2017-05-11 concurrency.tex
First commit

\section{Concurrency and Multithreading Programming}
\label{sec:concurrency}

A program is composed of data declarations, assignments and control flow, following the syntax and abstractions of a programming language. A concurrent program is a set of sequential programs that can be executed in parallel~\cite{Ben-Ari:1990}.

In 2004, the first multi-core processor was launched~\cite{Sutter:2005} into the mainstream market. The increasing number of cores of current parallel machines has motivated concurrent execution of tasks in all scientific and commercial domains, although this paradigm may still face some resistance~\cite{FreeLunchIsOver} \cite{threadsProblem:2006} \cite{KISS:2004:article}. Two important concepts are \emph{concurrency} and \emph{parallelism}. Concurrency occurs when multiple tasks are logically active at the same time and parallelism occurs when multiple tasks are actually active at the same time. In this case parallelism is related to physically performing tasks in parallel. In a computer with a multi-core processor we can have physical parallelism.

In practically every language that supports multi-threading, the implementation of this concept follows the idea of having a function with the code to be executed by the thread when it is created. Programming multi-threaded applications is a complex task, independently of the language used. The non determinism that is introduced by the use of threads can be controlled using synchronization mechanisms such as locks \cite{threadsProblem:2006}, semaphores and barriers, but these mechanisms increase the complexity of the program even more.

In C/C++ languages the \texttt{pthreads} library can be used to control thread execution on GNU/Linux. This library implements the standard POSIX API (IEEE 1003.1c) and provides mechanisms to create and synchronize threads. The choice to use the \texttt{pthreads} library, and not \texttt{OpenMP}, can be motivated by the fact that \texttt{pthreads} is a common library, available on platforms with GNU/Linux, and operates in a lower level of abstraction in comparison with \texttt{OpenMP} threads. The Code~\ref{cod:sample:pthreads} presents a sample of code that uses the \texttt{pthreads} library.

\begin{lstlisting}[style=C, label=cod:sample:pthreads,caption=The Sample of pthreads code.]
int main(){
  pthread_t thread1;
  DATA_TYPE *params = 10;
  int iret1;

  iret1 = pthread_create( &thread1, NULL, task_function, (void*) params);
  printf("Thread 1 return: %d\n", iret1);
  pthread_join(thread1, NULL);

  return 0;
}

void *task_function(void *ptr) {
  DATA_TYPE *value;
  value = (DATA_TYPE *) ptr;
  /* Computing */
}
\end{lstlisting}

This sample code shows some important concepts. Until \texttt{line 6} only one thread is executing the main flow of instructions. When the \texttt{pthread\_create()} function is called, a new thread is created and then the application have two threads executing in parallel. For synchronization between threads barriers are necessary. The library implements barriers and they can be used in the code by calling \texttt{pthread\_join()}. In \texttt{line 8} we can see an example of this, where the thread that reaches that point will wait for the completion of other threads to continue the execution of the main flow.

To explain the use of \texttt{pthreads} we use the sample code in Code~\ref{cod:basic} to create a parallel version. The code have a reduction operation, in the multiplication of elements of $A$ and $B$ to the $dot$ variable. In this case, we want to parallelize the loop instances by dividing the amount of instances among a number of threads.

\begin{lstlisting}[style=C, label=cod:basic,caption=Sample of Basic Code.]
 for(i = 0; i < N; i++){
   dot += A[i] * B[i];
 }
\end{lstlisting}

The Code~\ref{cod:sample:vetdot:pthreads} shows a vector product operation that was implemented using \texttt{pthreads}. This example shows that we need to write several lines of code to perform a simple operation like sharing work between threads.

% float=ht 
\begin{lstlisting}[style=C, label=cod:sample:vetdot:pthreads,caption=The vetdot sample using pthreads.]
int A[SIZE], B[SIZE];
int dot = 0;

pthread_mutex_t lock;

int main( int argc, char *argv[] ){
  int num_threads, i, part_size;

  pthread_t tid[MAX_THREADS];

  pthread_attr_t attr;
  pthread_attr_init(&attr);

  /* Initialization vectors was supressed. */

  part_size = (SIZE + num_threads - 1) / num_threads;

  for (i=0; i < num_threads; i++) {
  	struct arg_struct *args = malloc(sizeof(struct arg_struct));

  	args->id = i;
  	args->i_part = (i * part_size);
  	args->e_part = MIN(((i + 1)* (part_size)), SIZE);

  	pthread_create(&(tid[i]), &attr, vecDot, (void *) args);
  }

  for (i=0; i<num_threads; i++) {
  	pthread_join(tid[i], NULL);
  }

  printf("Final result: %d.\n", dot);

  return 0;
}

void *vecDot(void *arguments){
  struct arg_struct *args = arguments;
  int i, partial_dot = 0;

  for(i = args->i_part; i < args->e_part; i++){ 
  	partial_dot += A[i] * B[i]; 
  }

  /* Critical region. */
  pthread_mutex_lock(&lock);
  dot += partial_dot;
  pthread_mutex_unlock(&lock);

  pthread_exit(0);
}
\end{lstlisting}

In the main function, the arrays are initialized and divided among the number of created threads. Each thread executes the \texttt{vetDot} function, saving the calculated $partial\_dot$ value in the global $dot$ variable. This is done inside a delimited critical region, defined by the use of using \textit{pthread} \textit{mutexes}.

### 2017-05-11 directives.tex
First commit

\section{Compilation Directives}
\label{sec:directives}

Since the introduction of heterogeneous parallel computing platforms tools have been developed to provide interfaces between compilers, runtimes and execution environments. The compilation directives is one of the approaches that has excelled in the context of parallel computing, because annotating code is usually easier than rewriting it. These directives are commonly implemented using the pre-processing directives \texttt{\#pragma}, in \texttt{C/C++}, and sentinels \texttt{!\$}, in the Fortran language.

Directives are used as annotations that provide tips about the original code and guide the compiler in the parallelization process of candidate regions. During the pre-processing, the \textit{tokens} that compose the directive can be replaced using macros, for example. The parallel version of the code will use features provided by the \texttt{OpenMP} runtime.

%\url{http://www.google.com/trends/explore#cat=0-5&q=openmp%2C%20hpf&cmpt=q&tz=Etc%2FGMT%2B3}
The use of directives was not introduced by \texttt{OpenMP}. Directives have been used as a macro specification language used in the pre-processing of languages such as C/C++ and Fortran. The macros transform the source code before the compilation. Languages such as \texttt{HPF} \cite{Richardson96highperformance} \cite{hpf:1997} had compiler directives for parallelism before the rise of \texttt{OpenMP}. However, with the popularity of multi-core processors and \texttt{OpenMP} supporting C/C++ and Fortran applications, \texttt{OpenMP} has become a robust runtime for multi-core processors and an interesting application of compilation directives. Figure~\ref{fig:google:trends:openmp} presents a comparison between the search results for the terms \lq\lq\texttt{OpenMP}\rq\rq{}, \lq\lq\texttt{HPF}\rq\rq{} (High Performance Fortran) and \lq\lq\texttt{Cilk}\rq\rq. The constantly higher number of searches for \lq\lq\texttt{OpenMP}\rq\rq{} can be taken as confirmation of its popularity.

\begin{figure}[htpb]
\centering
\includegraphics[width=1\columnwidth,height=0.35\columnwidth]{figures/google-trends-openmp.pdf}
\caption{Comparison of the Google Trends results for \lq\lq\texttt{OpenMP}\rq\rq, \lq\lq\texttt{HPF}\rq\rq{} and \lq\lq\texttt{Cilk}\rq\rq.}
\label{fig:google:trends:openmp}
\end{figure}

Using compilation directives is interesting because legacy code applications can be annotated and then parallelized by the runtime, to achieve performance improvements on multi-core processors. If the annotated code is analyzed by a compiler that supports \texttt{OpenMP} \cite{Dagum1998a} \cite{openmp:api:2011} \cite{openmp:api:2013} \cite{Chapman:2007} \cite{openmp:site}, a parallel version will be generated. Otherwise, the directives will be ignored and the code will remain sequential. Using the \texttt{OpenMP} runtime, parallel C/C++ or Fortran programming requires only the learning the directives' usage.

\texttt{OpenMP} is a set of compilation directives, functions libraries and environment variables that are based on the C/C++ and Fortran languages. The high level directives are able to expose the low level parallelism resources available on native implementations like the \texttt{pthreads} interface on GNU/Linux. Using the \texttt{OpenMP} directives much of the work is done automatically and implicitly, parallel regions have implicit barriers and the declaration of shared or private variables is simple.

\texttt{OpenMP} implements the \texttt{fork-join} model, in which multiple threads of execution perform tasks defined implicitly or explicitly by \texttt{OpenMP} directives~\cite{openmp:api:2011}~\cite{openmp:api:2013}. Threads are created from annotated code blocks. The \textit{master thread} executes sequentially until it finds the first parallel region. The similar \texttt{fork} operation creates a team of threads to execute a parallel region, the \textit{master thread} is part of the team. Nested parallel regions are allowed. When all threads in a team reach the barrier that marks the end of the parallel region, the master thread continues until it finds a new parallel region. From the annotated blocks, the runtime can generate implicit threads to execute the parallel regions or assign to some other thread the execution of explicit tasks defined by the programmer using the \texttt{task} directive.

Multiple compilers implement \texttt{OpenMP}, such as \texttt{GCC} \cite{libgomp:2015} and LLVM \texttt{clang} \cite{LLVM:CGO04} \cite{clang:site} \cite{openmp:llvm:site}, for example. \texttt{OpenMP} is certainly easier to use than \texttt{pthreads}~\cite{CPE:CPE529}. The directives allow a first contact with parallel programming that uses basic examples, letting the programmer experience an effortless application of parallel computing. However, complex parallel programming require more than simple \texttt{OpenMP} directives. Code~\ref{cod:sample:openmp} presents an example of the \texttt{OpenMP} compilation directives used to compute the dot product, which can be compared to a \texttt{pthreads} version presented in Code~\ref{cod:sample:vetdot:pthreads}.

\begin{lstlisting}[style=C, label=cod:sample:openmp,caption=Sample of OpenMP code.]
#pragma omp parallel firstprivate(aux_dot)
{
  #pragma omp single
  printf("Begin of the parallel region, number of threads: %d\n", omp_get_num_threads());

  #pragma omp for schedule(runtime)
  for(i = 0; i < SIZE; i++){ 
    aux_dot += A[i] * B[i];
  }

  #pragma omp critical
  dot += aux_dot;

  #pragma omp master
  printf("Final result: %d.\n", dot);
}
\end{lstlisting}

Expanding the code with directives allows to see how \texttt{OpenMP} reduces the amount of code that the programmer needs to write. The code with the expansions is shown in Code~\ref{cod:expanded:openmp}.

\begin{lstlisting}[style=C, label=cod:expanded:openmp,caption=OpenMP expanded code]
long long A[SIZE], B[SIZE], dot = 0;

void subfunction0 (void *data) {
  long long aux_dot = 0;

  if (GOMP_single_start ())
    printf("Begin of the parallel region, number of threads: %d\n", omp_get_num_threads());
  GOMP_barrier ();

  {
    long i, _s0, _e0;
    if (GOMP_loop_runtime_start (0, SIZE, 1, &_s0, &_e0))
      do {
        long _e1 = _e0;
        for (i = _s0; i < _e0; i++){
          aux_dot += A[i] * B[i];
        }
      } while (GOMP_loop_runtime_next (&_s0, &_e0));
    GOMP_loop_end ();
  }

  GOMP_critical_start();
  dot += aux_dot;
  GOMP_critical_end();

  if (omp_get_thread_num () == 0)
    printf("Final result: %d.\n", dot);
}

int main() {
  /* Initializations were supressed. */

  GOMP_parallel_start (subfunction0, NULL, 2);
  subfunction0 (NULL);
  GOMP_parallel_end ();

  return 0;
}
\end{lstlisting}

The Code~\ref{cod:expanded:openmp} shows calls to \texttt{GOMP\_parallel\_start()} \texttt{(line 33)} and \texttt{GOMP\_parallel\_end()} \texttt{(line 35)}. These functions are responsible to start and finish the team of threads that run \texttt{subfunction0(...)}. In the code it is also possible to see the functions that are mapped to barriers and synchronization. The calls to \texttt{GOMP\_barrier()} in \texttt{line 8}, \texttt{GOMP\_critical\_start()} and \texttt{GOMP\_critical\_end()} in \texttt{lines 22} and \texttt{24}, are mapped to semaphore and barrier primitives that are implemented in specific \texttt{OpenMP} libraries.

The \texttt{OpenMP} runtime uses the standard thread library of the system, in this case, \texttt{pthreads}. The library is used together with other native system resources to control scheduling and synchronization. This shows that \texttt{OpenMP} is simpler comparing to code using \texttt{pthreads}, but the programmer still needs to master fundamental concepts about parallelism.

Tutorials present content in a structured way, showing the syntax of directives and exemplifying the use of annotations. The examples are simple in most cases, as can be seen in Code~\ref{cod:sample:openmp}. This kind of example brings to student the idea that \texttt{OpenMP} is always very simple. They simply annotate the code and make the program parallel.
In simple applications, we write one or two annotations in the code and generate a parallel version without difficulty. However, when we have more complex applications, it is necessary to know more concepts of parallel computing to apply the correct modifications in the source code.

Another common situation when using compilation directives is obtaining a better performance for a parallel version of the code. It is necessary to perform large efforts to include the annotations and their proper restrictions. That turns the process into something like \lq\lq{}annotation or directive oriented programming\rq\rq. The time gained by not rewriting the application code is spent declaring directives and their restrictions.

Krawezik and Cappello~\cite{CPE:CPE905} evidenced this difficulty by comparing MPI and three OpenMP versions of benchmarks. The most popular use of \texttt{OpenMP} directives is at the loop level. In this approach the programmer discovers potentially parallel loops by profiling and checking the dependencies. The programmer then annotates the code with directives starting and closing parallel regions at each discovered parallelizable loop. However, good results were obtained only when SPMD was implemented on \texttt{OpenMP}, and achieving good performance required a significant programming effort~\cite{CPE:CPE905}.

### 2017-05-11 graduate-students-perception.tex
First commit

\section{Graduate Students Perception}
\label{sec:students:perception}

The \texttt{OpenMP} exercise was proposed just after basic concepts of parallelism with shared memory were taught to the students.
The wording of the exercise was: \emph{\lq\lq{}Search for a sample code in \texttt{OpenMP} tutorials and manuals on the Internet that uses compilation directives and has some kind of error. Write a report introducing the code, pointing the problems and describing which are the corrections made to eliminate error.\rq\rq{}}.

After that we provided an anonymous questionnaire using \emph{Google Forms}. The question was: \emph{\lq\lq{}What is your opinion about your \texttt{OpenMP} experience and the \texttt{OpenMP} exercise?\lq\lq{}}. We got 24 answers, all of them with positive feedback, and selected some of them below.

Several students commented on how easy it was to use \texttt{OpenMP}: \lq\lq{}... I appreciate the idea of annotating an already existing program...\rq\rq{}; \lq\lq{}... with \texttt{OpenMP} it is possible to make an application parallel with low effort.\rq\rq{}; \lq\lq{}Why, when I learned concurrent programming, I did not learn it (\texttt{OpenMP})?\rq\rq{}. The students appreciated learning and using \texttt{OpenMP} in a simple example.

Several students also commented on the lack of a more explicit vision: \lq\lq.{}..When we use threads explicitly, I have the impression that it is easier to understand possible problems.\rq\rq{}; \lq\lq{}When I use \texttt{OpenMP}, I am always afraid about the correctness, and if I used the correct optimization flags\rq\rq{}. There were also some complaints about how barriers work in \texttt{OpenMP}: \lq\lq{}I had little experience with \texttt{OpenMP} and \texttt{pthreads}, but I believe that the understanding of \texttt{pthreads} is easier for those who already have experience in C. The (\texttt{OpenMP}) implicit barriers make me confused, since I never know when there is or not a barrier.\rq\rq{}.

There were also some more interesting ideas: "This experience encouraged me to write programs easier to parallelize with \texttt{OpenMP}.".
The following answer summarizes the opinions on the ease of use of the API, and also on the possible difficulty of working with \texttt{OpenMP}: \lq\lq{}... \texttt{OpenMP} appears to be a good framework for productivity, that is, it allows the creation of concurrent programs in less time and with less code rather than with \texttt{pthreads}. But it does not make parallel programming a trivial or error-free task.\rq\rq{}.

In another question we asked the students about their experience with programming and about the difficulties they had learning \texttt{OpenMP} directives. Figure~\ref{fig:DifOpenMP} shows the results.

Many students reported that they had difficulties to learn \texttt{OpenMP}. The majority had no previous experience, as shown in Figure~\ref{fig:students:experience}. In Figure~\ref{fig:DifOpenMP} it can be seen that most students also had less than ten years of programming experience. There was only $31$ days between the \texttt{OpenMP} introductory classes and the due date of the assignment, and this may have contributed to their difficulties.

These answers provide some evidence that, despite being considered an easy approach, \texttt{OpenMP} also requires knowledge and experience.

\begin{figure}[ht]
\centering
\includegraphics[width=0.95\columnwidth,height=0.9\columnwidth]{figures/experience.eps}
\caption{Graphic of dispersion of the experience time versus
    difficulty in the understanding of OpenMP}
\label{fig:DifOpenMP}
\end{figure}

\begin{comment}
Talvez seja importante falar também que os alunos tiveram aulas sobre conceitos de paralelismo antes.

Qual opinião você pode dar sobre sua experiência com \texttt{OpenMP} e sobre o Exercício de Programação 1?

1. Legal.

2. Achei interessante a ideia de anotar um programa que já funciona sequencialmente, mas esse método dificulta um pouco pensar em paralelo. Quando implementamos explicitamente as threads, tenho a impressão de que fica mais fácil entender os possíveis problemas.

3. Não conhecia o \texttt{OpenMP}, aparentemente é uma boa ferramenta, com alguma facilidade de aprendizado e já tenho utilizado para processamento de dados GPS sem maiores dificuldades. Sobre o EP1, aparentemente não trouxe grandes problemas técnicos. Apenas acho que, para efeito de aprendizado, ao invés de utilizar um grande problema, fosse mais conveniente propor mais pequenos problemas que servissem para salientar tópicos específicos.

4. Gostei bastante do EP1 por que o foco estava mais na compreensão de paralelismo e concorrência e menos num algoritmo complexo. Testes de desempenho sempre são divertidos e interessantes (comparação, interpretação).

5. Acredito que o EP1 foi bastante útil para o aprendizado das API do \texttt{OpenMP} para a utilização na prática.

6. Foi uma experiência interessante. Nunca tinha usado uma API de alto nivel. Só tinha usado semáforos e mutex. Gostei o fato que o \texttt{OpenMP} faz que, com pouco esforço, uma aplicação se torne paralela.

7. A implementação foi bem fácil (porque o problema era bem simples) e deu espaço para experimentarmos um pouco com as comparações e gráficos dos resultados; funcionou muito bem.

8. A utilização do \texttt{OpenMP} é bastante simples comparada com o uso de threads ou troca de mensagens como MPI (as duas outras ferramentas para programação paralela que eu tinha experiência anteriormente). Porém, os mesmos cuidados com deadlocks, granularidade, barreiras e etc. valem para o \texttt{OpenMP} também.

9. Foi instrutivo utilizar o \texttt{OpenMP}, apredner um outro método de fazer paralelismo

10. Enquanto estudava \texttt{OpenMP} fiquei pensando "por que raios isso não foi visto depois de pthread em Concorrente?".

11. Os exercícios foram simples de entender e nos permitiram avaliar a razão dos resultados obtidos.

12. Achei legal a experiência de ver a comparação de desempenho por processador e por número de threads. Outra coisa que gosto é a parte de focar mais no conceito e seu entendimento. EP simples que tocava exatamente no conceito me parece melhor que um EP super trabalhoso que talvez desse trabalho em pontos não relevantes à disciplina.

13. Quanto à \texttt{OpenMP} as dificuldades são relacionadas a: - corretude: será que nenhum caso foi esquecido (algo comum em paralelismo)? - otimização: será que a escolha foi a melhor possível para a implementação?
Quanto ao EP1, as dificuldades foram:
A maior dificuldade foi no exercício 1. Foi difícil definir uma estratégia para avaliar os códigos encontrados: compilar e executá-los poderia trazer problemas mais rápidos mas tomaria mais tempo um por um enquanto que avaliar somente analisando o código era mais difícil encontrar erros em um dado exemplo mas era possível analisar mais códigos. Quanto ao exercício 2, a dificuldade foi encontrar máquinas com arquiteturas diferentes. A menos que vc esteja envolvido em alguma área de computação científica ou paralela, a grande maioria dos computadores são x86.

14. Foi uma experiência positiva. Uma boa introdução a paralelização, que mostrou o quanto é prático usar \texttt{OpenMP}, pelo menos para uma aplicação simples como a do EP1. O problema em questão era suficientemente interessante para pensar nas questões do paralelismo por permitir várias formas de abordagem, e as ferramentas básicas do \texttt{OpenMP} foram suficientes para os experimentos e discussões.

15. Essa experiência me encorajou a usar o \texttt{OpenMP} em programas que eu venha a escrever e que sejam fáceis de paralelizar."

16. Tive contanto com conceitos de paralelismo no meu curso de graduação inclusive utilizando o \texttt{OpenMP}, porém a experiência da utilização do \texttt{OpenMP} neste trabalho foi muito melhor. Acho que senti bastantante a diferença neste EP1 por partirmos de uma versão sequencial, diferentemento do trabalho que fiz na graduação e o \texttt{OpenMP} se mostrou muito bem nesta situação.

17. O \texttt{OpenMP} é muito fácil de aprender mas há códigos com erros na internet, geralmente esses códigos tinham erros na operação "cout" do c++, geralmente usam ele para mostrar dados dentro de uma região paralela, pareceria que não sabem que essa operação não é atómica, outros tipos de erros que achei foram erros no código, por exemplo o erro mais comum que eu vi foram que eles não verificam se o indice de um arranjo é valido ou não; quer dizer que alguns tutoriais tem erros no mesmo algoritmo não verificam se o algoritmo é correcto ou não.

18. Foi o primeiro contato que tive com \texttt{OpenMP}. Achei bastante proveitoso aprender a utilizar o \texttt{OpenMP}. Além disso, foi divertido fazer experimentos para avaliar o desempenho do programa usando diferentes implementações.

19. O \texttt{OpenMP} me parece um bom framework para produtividade, ou seja, permite a criação de programas concorrentes em menos tempo e com menos código do que com pthreads, por exemplo, mas não torna a programação paralela uma tarefa trivial ou livre de erros.

20. "O \texttt{OpenMP} oferece uma sintaxe concisa para resolver um problema complexo. Dominar o poder que a interface proporciona requer um tempo considerável de estudo e reflexão para se chegar ao resultado esperado, portanto trata-se muito mais de um exercício intelectual do que mecânico.
A sintaxe tem uma carga conceitual muito grande. Tenho experiência em trabalhar com outras linguagens que têm características semelhantes de concisão e carga conceitual (por exemplo, Ruby) e acredito que leva um tempo relativamente mais alto para se obter proficiência nesse tipo de tecnologia, quando comparado com tecnologias de mais baixo nível. Mas o ganho de produtividade é indiscutível.
Sobre o EP1, achei que foi bem dimensionado para o conteúdo apresentado em aula. O que acredito que poderia ajudar é colocar exemplos de código para análise de desempenho e de scripts para coletar os dados obtidos, de forma a proporcionar foco no \texttt{OpenMP} para os estudantes desenvolverem o trabalho."

21. Gostei do desenvolvimento do programa sugerido(exerc.2 programa mult.c). Mas não gostei do exercício de procurar códigos errados na internet(exerc. 1). Gostaria de ter tido mais exercícios programando, analisando performance, etc... Ou seja, acho que poderíamos ter explorado melhor a tecnologia, fazendo mais códigos e análise de performance.

22. Achei muito interessante o desenvolvimento do EP1. Ajudou a mostrar a facilidade do uso do \texttt{OpenMP}, conseguindo melhoras significativas de performance com pequenas mudanças no código.

23. Toma tempo acostumbrar ao troco de programas sequenciais para os paralelos, mas com \texttt{OpenMP} não foi muito dificil porque só tem que colocar algumas linhas de código e o programa se faz paralelo (mas igual tem que ter cuidado algumas vezes).

24. Tive pouca experiência com \texttt{OpenMP} e pthreads, mas acredito que o entendimento de pthreads é mais fácil para quem já tem experiência em C. As barreiras implícitas me confundem porque nunca tenho certeza se existe barreira ou não.

\end{comment}

### 2017-05-11 introduction.tex
First commit

\section{Introduction}
\label{sec:introduction}

There is an ongoing evolution of parallel processing platforms. The pursuit for high performance in general-purpose and scientific applications have turned to the exploration of the parallel processing capabilities provided by multi-core processors and arrangements of coprocessors like the Xeon Phi~\cite{xeonphiRuntime}. The machine currently at the top of the \textit{TOP500} list~\cite{dongarra1997top500} uses multi-core processors (12 cores) and Xeon Phi coprocessors~\cite{top500:site}.
Multi-core processors have become popular on personal computers and mobile devices. This popularization increases the need for specialized developers able to work in parallel applications and apply parallel computing models~\cite{Chen:2009:ChallengesMIC}~\cite{okur2012developers}.

Today there is a wide gap between legacy code applications that could benefit from parallel computing power and the new parallel computing platforms that could provide that power. However, legacy applications are generally sequential and therefore not easily scalable to multi-core architectures and parallel computing platforms.

Asanovic \textit{et al.} \cite{Asanovic2009} describes this scenario, and defends the need for Structural and Computational Patterns in Parallel Computing. These patterns allow the portability of parallel programming solutions, while also allowing specialization of applications with the same behavior. In some situations it is not a matter of achieving the highest possible performance, but instead of making the application achieve acceptable performance, benefiting from a new parallel computing platform~\cite{Asanovic2009}. A similar approach is proposed by Pllana \textit{et al.} in a development environment that uses a combination of Model-Driven Development and Parallel Building Blocks \cite{Pllana:2009}. The approach consists in creating the model and then generating or transforming the code using \emph{model-to-source} tools.

In this context, the approach of using Compilation Directives to guide the compiler in the parallelization process has gained popularity. These directives are implemented using the compiler's pre-processing directives and are utilized as annotations that provide tips about the sequential code. Among the tools and extensions that use compilation directives are the \texttt{OpenMP} API \cite{Dagum1998a} \cite{Chapman:2007} used for writing programs for multi-core architectures, and the \texttt{OpenACC} programming standard~\cite{openacc:api}, used for writing programs for heterogeneous CPU/GPU architectures. The \texttt{OpenMP 4.0} specification supports offloading capabilities \cite{openmp:api:2013} such as \texttt{OpenACC} and \texttt{AMD HSA}~\cite{amd:hsa:site}. These tools aim to make easier the task of designing and implementing efficient and provably correct parallel algorithms. However, the abstraction layers built by the extensions also have the potential to conceal architectural details and to generate new parallel programming challenges.

When comparing \texttt{OpenMP} to other available tools for developing multi-threaded applications such as the \texttt{pthreads} interface, the API can be considered almost costless. However, when writing more complex applications this may not be true. The design and implementation of such complex applications require more knowledge about platforms and execution models than what can be shown by tutorials, that present high level and trivial examples. These tutorials try to convince the user that programming with directives is simple and easy to do, which is not always true. For instance, to achieve good performance in some cases it is necessary to describe many restrictions on annotations~\cite{OpenMPTasks2009}~\cite{mattson2003good}.

Krawezik and Cappello~\cite{CPE:CPE905} evidenced this difficulty by comparing MPI and three OpenMP versions of benchmarks. The most popular use of \texttt{OpenMP} directives is at the loop level. The programmer discovers potentially parallel loops and annotates the code with directives starting and closing parallel regions. Good results were obtained only when SPMD was implemented on OpenMP, and achieving good performance required a significant programming effort~\cite{CPE:CPE905}.

In this work we have gathered data from assignments made by graduate students for the \emph{Introduction to Parallel and Distributed Computing} course. Among other exercises, the students were asked to find examples of \texttt{OpenMP} code in tutorials and official manuals that failed under special conditions. These conditions could be created by, for example, placing barriers or forcing racing conditions by exposing inaccurate thread management.

We argue that it is necessary to consider fundamental concepts of architectures and parallel programming techniques when teaching students new to the subject, so that they can learn when and how to apply the programming models~\cite{ben1999thinking}. The abstractions provided by parallel programming tools over concepts such as threads, barriers and loop parallelization should be carefully studied, analysed and conceptually built in the classroom before they are used by the students in their assignments.

We found eight tutorials with different kinds of \texttt{OpenMP} errors in the experiment with the students' assignments. Many of these errors where similar to the most common \texttt{OpenMP} mistakes~\cite{SuB:2005:CMO:1892830.1892863}. The most common type of error identified in our experiment was the management of shared or private variables. This problem is common across multiple programming paradigms.

The \textbf{main contribution of this work} is to show the importance of teaching core architecture and parallel programming concepts thoroughly, despite the availability of powerful tools such as \texttt{OpenMP}. We also aim to show that the \texttt{OpenMP} API is not as easy to use as it might seem.

The rest of the paper is organized as follows. The following section discusses related work. Section~\ref{sec:concurrency} discusses concurrency and other concepts in parallel computing. Section~\ref{sec:directives} presents the idea of compilation directives. Section~\ref{sec:methodology} discusses the methodology applied in the experiments, and Section~\ref{sec:results} presents our results. Section~\ref{sec:students:perception} presents the perceptions of the graduate students regarding the exercise and learning \texttt{OpenMP}. Finally, Section~\ref{sec:conclusions} presents the conclusions and next steps.

% Institution 2 = University of São Paulo

### 2017-05-11 methodology.tex
First commit

\section{Methodology}
\label{sec:methodology}

% Institution 2 = University of São Paulo

The motivation to write this paper comes from experiences on teaching. During the classes of Distributed Systems and Parallel Programming, a graduate course at University of São Paulo, a problem was detected in an \texttt{OpenMP} code sample.

To explain the idea we use as a basis the code sample showed in Code~\ref{cod:basic} of section \ref{sec:concurrency}. If we need to parallelize the loop instances, we can use the \texttt{OpenMP} \emph{reduction} clause because the code has a \emph{reduction} operation, but the idea of the tutorial was to show the use of the \emph{critical} clause.

The code written on the tutorial sample to solve using \emph{critical} clause is shown in Code~\ref{cod:tutorial:sample:with:problem}. However, this code has a problem.

\begin{lstlisting}[style=C, label=cod:tutorial:sample:with:problem,caption=Code of original OpenMP tutorial with problem.]
#pragma omp parallel
{
  #pragma omp for
  for(i = 0; i < N; i++){
   aux_dot += A[i] * B[i];
  }
  #pragma omp critical
  dot += aux_dot;
}
\end{lstlisting}

The idea that was used to parallelize the computing on \texttt{line 5} was to put the for loop inside a parallel region and parallelize the instances of the loop using the directive \texttt{\#pragma omp for}. An auxiliary variable $aux\_dot$ was used to keep the partial results in each thread, but $aux\_dot$ was not declared as private to each thread. This generated a race condition, in which threads can read the same value and can be interrupted before storing the updated value.

This situation was identified as a possible problem. In the second tutorial the authors thought the problem was in the initialization of the variable $dot$. The code was \lq\lq{}fixed\rq\rq{} using the \texttt{firstprivate(dot)} clause. The semantic of this clause is: \lq\lq{}all variables in the list are initialized with the original value before entering the parallel region\rq\rq{}. Code~\ref{cod:tutorial:sample:proposed:correction:1} presents the first attempt to fix the code in the class.

\begin{lstlisting}[style=C, float=htpb, label=cod:tutorial:sample:proposed:correction:1,caption=The first version of "fixed" code.]
#pragma omp parallel firstprivate(dot)
{
  #pragma omp single
  printf("Begin of the parallel region, number of threads: %d\n", omp_get_num_threads());
  #pragma omp for
  for(i = 0; i < SIZE; i++){
    aux_dot += A[i] * B[i];
    printf("Thread %d executes the loop iteration %02d: %lld * %lld = %lld -> %lld \n", omp_get_thread_num(), i, A[i], B[i], (A[i] * B[i]), aux_dot);
  }
  #pragma omp critical
  dot += aux_dot;

  #pragma omp master
  printf("Final result: %lld.\n", dot);
}
\end{lstlisting}

% Critical sections don't have barriers, neither at their beginnings nor at their ends. A critical section is a synchornisation construct on its own that prevents multiple threads from accessing the same data concurrently. You need an additional barrier after the critical section if you'd like to have the correct global minimum before you exit the parallel region. As was already said the parallel region has an implicit barrier at the end.

The author added lines on the code to show the final result, but with these modifications a new error was inserted. Only by the master thread prints the final result message, and this can occur before the completion of other threads, because the \emph{critical} clause does not have barriers, it only prevents concurrent accesses. The corrected code still had problems. One team of students found another problem in the \lq\lq{}fixed\rq\rq{} version of sample code. Code~\ref{cod:tutorial:sample:proposed:correction:2} presents the fixed and final code version.

\begin{lstlisting}[style=C, float=htpb, label=cod:tutorial:sample:proposed:correction:2,caption=The final version of code sample.]
#pragma omp parallel firstprivate(aux_dot)
{
  #pragma omp single
  printf("Begin of the parallel region, number of threads: %d\n", omp_get_num_threads());
  #pragma omp for
  for(i = 0; i < SIZE; i++){
    aux_dot += A[i] * B[i];
    printf("Thread %d executes the loop iteration %02d: %lld * %lld = %lld -> %lld \n", omp_get_thread_num(), i, A[i], B[i], (A[i] * B[i]), aux_dot);
  }
  #pragma omp critical
  dot += aux_dot;
}
printf("Final result: %lld.\n", dot);
\end{lstlisting}

The last correction puts the final message out of the parallel region to avoid printing incomplete results. This occurs because at the end of all parallel regions in OpenMP there is one implicit barrier that joins the work flow and only the \texttt{master thread} continues the execution in this case.

% Para nossa surpresa, não apenas um dos nossos exemplos em sala estava com problemas, mas erros foram encontrados em diversos tutoriais disponíveis na internet.

This simple example illustrates how mistakes can be made. The example motivated the proposition of an exercise for the students. It also reinforced the importance of fundamental parallelism concepts even in simpler approaches. A challenge was then proposed for the students, as one of the activities of the course. 
After the introductory lecture on \texttt{OpenMP}, the groups should search available \texttt{OpenMP} tutorials on the Internet for errors, and fix them. This exercise had $27$ participants, the group of students was divided in $6$ teams of two students and other students preferred to work alone. Most students used the \texttt{GCC} compiler and the \texttt{GNU libgomp} library in their assignments.

We applied a questionnaire to obtain more information about the students. Most students already knew threads, but few students had experience with \texttt{OpenMP}. Figure~\ref{fig:students:experience} presents how many students had contact with each parallel programming paradigms before the course. Since this is a graduate level course, we had students of various ages. This is reflected by the large differences in experience time between students, showed in Figure~\ref{fig:DifOpenMP}.

\begin{figure}[htpb]
\centering
% \includegraphics[width=0.95\columnwidth,height=0.65\columnwidth]{figures/experiencePP.eps}
\includegraphics[width=1.0\columnwidth,height=0.70\columnwidth, trim={0 1.0cm 0 1.5cm},clip]{figures/experiencePP.pdf}
\caption{APIs for Parallel Programming previously used by the students}
\label{fig:students:experience}
\end{figure}

The other question was about the programming languages known by the students. The objective of this information was to measure how easy it is to learn \texttt{OpenMP} when you know the basic concepts and what is the influence that the knowledge of other languages can have on the process. Figure~\ref{fig:LangMoreUsed} presents the distribution of programming languages known for the students involved in the experiment.

\begin{figure}[htpb]
\centering
% \includegraphics[width=0.95\columnwidth,height=0.65\columnwidth]{figures/experienceLP.eps}
\includegraphics[width=1.0\columnwidth,height=0.70\columnwidth, trim={0 1.0cm 0 1.5cm},clip]{figures/experienceLP.pdf}
\caption{Programming Languages more used by the students}
\label{fig:LangMoreUsed}
\end{figure}

### 2017-05-11 relatedwork.tex
First commit

\section{Related Work}
\label{sec:relatedwork}

There has been a large body of works that have presented results of studies and experiments in teaching parallel computing courses. S\"{u}\ss~and Leopold~\cite{SuB:2005:CMO:1892830.1892863} discuss the teaching of parallel computing, in particular the use of \texttt{OpenMP}. The authors present the most common errors identified in assignments given in parallel computing classes. The errors were classified in two categories: \emph{correctness mistakes} and \emph{performance mistakes}. Errors that impact the correctness of a program are errors of the first category. Errors which make programs perform slower are classified in the second category. The work also provides a checklist containing tips to avoid the most common mistakes.

Falcao~\cite{6565518} defends that teaching a parallel programming model or interface such as \texttt{OpenMP} in the beginning of undergraduate courses is not much harder than teaching a language like C or C++. The benefits of teaching these skills early on in the course would be interesting, considering that the majority of computer architectures are now are multi-core. Parallel computing should be taught with the use of higher level tools. One of the approaches proposed by Ferner \textit{et al.}~\cite{6651019} is to use compiler directives to describe how a program should be parallelized.

The need for Structural and Computational Patterns to Parallel Computing is defended by Asanovic \textit{et al.}~\cite{Asanovic2009} as the way to allow portability of solutions and the specialization of applications with the same behavior. Using patterns to design applications make it easy to change the target platform. The applications will be bound to platforms in low levels of architecture and probably inside a specialized function which can be easily changed.

A similar approach is proposed by Pllana \textit{et al.}~\cite{Pllana:2009} in a programming environment that uses a combination of Model-Driven Development and Parallel Building Blocks. The idea is to create the model and then generate or transform the code using model-to-source tools. The proposed environment uses intelligent agents to assist the programmer, this is done during the high level program composition and when the programmer loads a BLAS function for some dense linear algebra operation, the tool suggests the use of the appropriate parallel basic block. The parallelism is inside the parallel building blocks and the programmer is not exposed directly to the complexity of parallel programming problems.

Speyer \textit{et al.}~\cite{4755913} defend that implementing applications with high performance is harder than ever, specially due to the emergence of multi-core architectures. This leads to the development of hybrid applications that use both the \texttt{OpenMP} and \texttt{MPI} (\emph{Message Passing Interface}) APIs. This presents even greater challenges to the programmers, since they now need to know how to use both code annotations and message passing abstractions to achieve high performance. The paper then compares different parallel programming tools with regard to their ease of use. There is a substantial effort to start writing parallel and high performance code even with these tools.

Some problems can be solved using different directive combinations, and the chosen combination influences the code that is generated. The \texttt{Parallware}~\cite{Parallware:2014} source-to-source compiler showed how different strategies and uses of directive combinations can lead to different performance results. The use of the wrong \texttt{OpenMP} directives and strategies can significantly reduce the performance over a sequential version.

The \texttt{OmpVerify}~\cite{Basupalli:2011} system implements a static analysis tool based on the polyhedral model to extract the parallel portions of a program automatically and to detect important classes of common data-race errors in \texttt{OpenMP} \texttt{parallel for} directives. Saillard \textit{et al.}~\cite{Saillard:2014} statically validates barriers and work sharing constructs to avoid the improper use of directives that could cause deadlocks. S\"{u}\ss~and Leopold~\cite{Leopold:userOpenMP}, present an user's experience with parallel sorting and OpenMP. They have used several versions of a simple parallel sorting algorithm to show some weaknesses of OpenMP.

Substantial efforts have been made to achieve the simplest possible way to parallelize code, and the task is not easy. In this paper we present an exploratory analysis of common \texttt{OpenMP} errors, confirming that programming with high-level APIs and abstractions can still be difficult.

### 2017-05-11 results.tex
First commit

\section{Results}
\label{sec:results}

After the students submitted the assignment for evaluation, we analyzed the reports and tutorials where errors were found. One of the examples shown in class presented an error, and multiple errors were found in several tutorials available on the Internet. Table~\ref{table:tutorials:with:problems} summarizes the errors found in tutorials from the Internet and their corrections by the students.
\begin{table}[ht]
\renewcommand{\arraystretch}{1.3}
\caption{Tutorials that presented some problem}
\label{table:tutorials:with:problems}
\centering
\begin{tabular}{|>{\centering\arraybackslash} m{.2cm}|>{\centering\arraybackslash} m{1.7cm}|  >{\centering\arraybackslash} m{2.75cm}|>{\centering\arraybackslash} m{2.75cm} |}
\hline
\textbf{Id}&\textbf{Tutorial} & \textbf{Identified Problem} & \textbf{Correction}\\
\hline

\begin{minipage}[m]{0.02\columnwidth}%
\centering
1
\end{minipage}&
\begin{minipage}[t]{0.22\columnwidth}%
\texttt{Tutorial de OpenMP C/C++~\cite{LCCV:UFAL:Tutorial}}
\end{minipage}&
\begin{minipage}[t]{0.33\columnwidth}%
Incorrect result. Race condition in results variables was detected.
\end{minipage}& 
\begin{minipage}[t]{0.33\columnwidth}%
The clause firstprivate was used to transform $aux\_dot$ in private variable in threads. \tiny{\texttt{\#pragma omp parallel firstprivate(aux\_dot)}}
\end{minipage}
\tabularnewline
\hline

\begin{minipage}[m]{0.02\columnwidth}%
\centering
2
\end{minipage}&
\begin{minipage}[t]{0.22\columnwidth}%
\texttt{Guide into OpenMP~\cite{OpenmpAndFork}}
\end{minipage}&
\begin{minipage}[t]{0.33\columnwidth}%
\texttt{OpenMP} interaction with Linux fork implementation using the GCC. The created thread pool are passed to child processes, but the fork() do not duplicate the threads causing deadlock in the child processes.
\end{minipage}&
\begin{minipage}[t]{0.33\columnwidth}%
Make fork() before application of pragmas in the parent process. Organize the code for child processes create their own thread pool.
\end{minipage}
\tabularnewline
\hline

\begin{minipage}[m]{0.02\columnwidth}%
\centering
3
\end{minipage}&
\begin{minipage}[t]{0.22\columnwidth}%
\texttt{Code Project~\cite{CodeProject}}
\end{minipage}&
\begin{minipage}[t]{0.33\columnwidth}%
Sintax error. Missing the~\lq\lq\{\rq\rq~after directive for.
\end{minipage}&
\begin{minipage}[t]{0.33\columnwidth}%
The students tested take off~\lq\lq\}\rq\rq~at the final causing segmentation fault error.
So they put the~\lq\lq\{\rq\rq~after the directive for.
\end{minipage}
\tabularnewline
\hline

\begin{minipage}[m]{0.02\columnwidth}%
\centering
4
\end{minipage}&
\begin{minipage}[t]{0.22\columnwidth}%
\texttt{Parallel Processing with OpenMP~\cite{BostonUniversity}}
\end{minipage}&
\begin{minipage}[t]{0.33\columnwidth}%
Error on use of ordered clause. It is necessary to indicate that the ordered clause will be used inside of loop.
\end{minipage}& 
\begin{minipage}[t]{0.33\columnwidth}%
The ordered clause was put.
\end{minipage}
\tabularnewline
\hline
\begin{minipage}[m]{0.02\columnwidth}%
\centering
5
\end{minipage}&
\begin{minipage}[t]{0.22\columnwidth}%
\texttt{Task Parallelism in OpenMP~\cite{UniversityUtah}}
\end{minipage}&
\begin{minipage}[t]{0.33\columnwidth}%
The $tid$ is not defined as private. It is shared by all threads the vector $A$ values are not calculated.
\end{minipage}&
\begin{minipage}[t]{0.33\columnwidth}%
Declare $tid$ as private using the \texttt{private(tid)} clause.
\end{minipage}
\tabularnewline
\hline
\begin{minipage}[m]{0.02\columnwidth}%
\centering
6
\end{minipage}&
\begin{minipage}[t]{0.22\columnwidth}%
\texttt{Introduction to Parallel Processing. OpenMP~\cite{UniversityEmory}}
\end{minipage}&
\begin{minipage}[t]{0.33\columnwidth}%
Use of \texttt{cout} function inside parallel region without critical clause. The threads are interrupted during the printing, messages are concatenated.
\end{minipage}&
\begin{minipage}[t]{0.33\columnwidth}%
The code was modified using the critical clause.
\end{minipage}
\tabularnewline
\hline
\begin{minipage}[m]{0.02\columnwidth}%
\centering
7
\end{minipage}&
\begin{minipage}[t]{0.22\columnwidth}%
\texttt{Introduction to Parallel and Distribution Systems. OpenMP Overview~\cite{UniversityWayne}}
\end{minipage}&
\begin{minipage}[t]{0.33\columnwidth}%
In the matrix multiplication code the variables $j$ and $k$ are shared by all threads generating wrong result.
\end{minipage}&
\begin{minipage}[t]{0.33\columnwidth}%
Addition of private clause on \tiny{\texttt{\#pragma omp parallel for private(j, k)}}
\end{minipage}
\tabularnewline
\hline
\begin{minipage}[m]{0.02\columnwidth}%
\centering
8
\end{minipage} &
\begin{minipage}[t]{0.22\columnwidth}%
\texttt{Introduction to OpenMP. Dijkstra sample~\cite{UniversityCalifornia}}
\end{minipage} & 
\begin{minipage}[t]{0.33\columnwidth}%
When the code is executed with more of 3 threads produces wrong results. There is an error in assigning values for $startv$ and $endv$ variables that keep the start and end positions of the shortest path. With this problem the algorithm does not reach all graph vertices. On line 88 a critical section is created for updating the minimum value and the vertex ($md$ and $mv$) but the code is not avoiding that updates occurs before another thread update the global vector, it is printing the values set in the initialization vector.
\end{minipage}&
\begin{minipage}[t]{0.33\columnwidth}%
The definition of variables $startv$ and $endv$ was modified using the original idea of algorithm. The \texttt{\#pragma omp barrier} directive was added after the \texttt{\#pragma omp critical}.
\end{minipage}
\tabularnewline
\hline

\end{tabular}
\end{table}

% Esses sites são os seguintes:
% http://bisqwit.iki.fi/story/howto/openmp/#OpenmpAndFork, 
% http://www.codeproject.com/Articles/60176/A-Beginner-s-Primer-to-OpenMP
% Universidade de Boston - www.bu.edu/tech/files/2011/09/openmp.ppt
% Tutorial disponiblizado pela Universidade de Utah, dado pelo seguinte link: http://www.cs.utah.edu/~mhall/cs4961f11/CS4961-L9.pdf.
% Tutorial da universidade de Emory (USA) OpenMP  http://www.mathcs.emory.edu/~cheung/Courses/355/Syllabus/91-pthreads/openMP.html
% Universidade do Estado de Wayne (USA) Apresentação de Cheng-Zhong Xu. site: http://www.ece.eng.wayne.edu/ czxu/ece561/lnotes/OpenMP.pdf
% Site de Question, Answer. http://askubuntu.com/questions/590755/openmp­unable­to­use­multiple­threads­no­error­a ssociated­with­compilation. Já tinha sido corregido.
% Universidade de California, Davis. Introduction to OpenMP.  http://heather.cs.ucdavis.edu/~matloff/openmp.html#tutorials. http://heather.cs.ucdavis.edu/~matloff/OpenMP/Examples/NM/Dijkstra.c

The \emph{first tutorial} in Table~\ref{table:tutorials:with:problems} has a problem caused by a confusion about which variables are shared and which should be private. This problem is similar to the ones identified by S\"{u}\ss~and Leopold~\cite{SuB:2005:CMO:1892830.1892863} as common correctness mistakes (\lq\lq{}Forgetting to mark private variables as such\rq\rq{}). It was the fourth most frequent mistake in that study.

\emph{Tutorial 2} presents a problem with the interaction between \texttt{OpenMP} and the system's OS and compiler. Specifically, the \texttt{GCC} \texttt{fork()} implementation. When we use the fork primitive, the Operating System creates other processes (child processes) duplicating all resources used by processes that execute the \texttt{fork()} call. Using \texttt{OpenMP} and creating child processes makes the thread pool created by \texttt{OpenMP} to be passed to child processes, but \texttt{fork()} does not duplicate the threads, which causes deadlock in the child processes.

The problem presented in \emph{tutorial 3} was a syntax error, there was a missing ~\lq\lq\{\rq\rq{} after the directive.

\emph{Tutorial 4}, presented in Table~\ref{table:tutorials:with:problems}, was found by students because of an error using the \texttt{ordered} clause. When using this clause it is necessary to indicate that it will be used inside a loop. A similar mistake involving the clause was identified as common by S\"{u}\ss~and Leopold~\cite{SuB:2005:CMO:1892830.1892863}. The opposite situation happened when an \texttt{ordered} clause was put into a \texttt{for} construct without being specified by a separate \texttt{ordered} clause inside the loop. This is a restriction on the nesting of regions, an \texttt{ordered} region must be closely nested inside a loop or parallel loop region with an \texttt{ordered} clause~\cite{openmp:api:2011}.

Other problems in private variable definition were found in \emph{tutorials 5} and \emph{7}. In \emph{tutorial 5} the \texttt{tid} variable is not defined as private, causing some values in the array not being calculated. In \emph{tutorial 7} the variables \texttt{j} and \texttt{k} in the matrix multiplication are shared by all threads, generating wrong results.

In \emph{tutorial 6} the \texttt{cout} function is called inside a parallel region without a protecting \texttt{critical} clause. The call becomes a critical section, because the threads are interrupted during the printing and messages are concatenated generating incorrect results. The code was modified by students using the \texttt{critical} clause to ensure that the call is executed by one thread at a time.

\emph{Tutorial 8} presents a synchronization problem that causes an error in the assignment of values for the \texttt{startv} and \texttt{endv} variables that keep the start and end positions of the shortest path. Because of this problem the algorithm does not reach all graph vertices. The critical section is created to update the minimum value and the vertex (\texttt{md} and \texttt{mv}), but the code is not avoiding the occurrence of updates before another thread updates the global array. If all vertices are not visited some initialized values are not updated during the program execution, which prints the incorrect values. The code was modified using the original algorithm idea and a barrier was added to ensure synchronization using the \texttt{\#pragma omp barrier} directive after the \texttt{\#pragma omp critical}.

Table~\ref{table:summary:of:errors} summarizes the type of errors found by the students in Internet tutorials.

\begin{table}[htpb]
\renewcommand{\arraystretch}{1.3}
\caption{Summary of errors}
\label{table:summary:of:errors}
\centering
\begin{tabular}{|@{$~$}l@{ }|@{$~$}l@{ }|@{$~$}l@{ }|}
\hline
\textbf{Kind} &\textbf{Quantity} & \textbf{Tutorial} \\
\hline

\begin{minipage}[m]{0.35\columnwidth}%
private-shared
\end{minipage} &
\begin{minipage}[t]{0.10\columnwidth}%
\centering
3
\end{minipage} & 
\begin{minipage}[t]{0.20\columnwidth}%
1, 5 and 7
\end{minipage}
\tabularnewline
\hline

\begin{minipage}[m]{0.35\columnwidth}%
syntax error
\end{minipage} &
\begin{minipage}[t]{0.10\columnwidth}%
\centering
1
\end{minipage} & 
\begin{minipage}[t]{0.20\columnwidth}%
3
\end{minipage}
\tabularnewline
\hline

\begin{minipage}[m]{0.35\columnwidth}%
wrong use of directive
\end{minipage} &
\begin{minipage}[t]{0.10\columnwidth}%
\centering
1
\end{minipage} & 
\begin{minipage}[t]{0.20\columnwidth}%
4
\end{minipage}
\tabularnewline
\hline

\begin{minipage}[m]{0.35\columnwidth}%
synchronization problems
\end{minipage} &
\begin{minipage}[t]{0.10\columnwidth}%
\centering
2
\end{minipage} & 
\begin{minipage}[t]{0.20\columnwidth}%
6 and 8
\end{minipage}
\tabularnewline
\hline

\end{tabular}
\end{table}

Comparing the problems found in the experiment to the most common \texttt{OpenMP} mistakes~\cite{SuB:2005:CMO:1892830.1892863} we can see how some errors are recurrent, for instance, the decision about isolating data using declarations of shared or private variables.

### 2017-07-13 sections/01-introduction.tex
Re-organizing and fixing the used template

--- 2017-05-11 introduction.tex
@@
 
-There is an ongoing evolution of parallel processing platforms.
-The pursuit for high performance in general-purpose and scientific applications have turned to the exploration of the parallel processing capabilities provided by multi-core processors and arrangements of coprocessors like the Xeon Phi~\cite{xeonphiRuntime}.
-The machine currently at the top of the \textit{TOP500} list~\cite{dongarra1997top500} uses multi-core processors (12 cores) and Xeon Phi coprocessors~\cite{top500:site}.
-Multi-core processors have become popular on personal computers and mobile devices.
-This popularization increases the need for specialized developers able to work in parallel applications and apply parallel computing models~\cite{Chen:2009:ChallengesMIC}~\cite{okur2012developers}.
+...
 
-Today there is a wide gap between legacy code applications that could benefit from parallel computing power and the new parallel computing platforms that could provide that power.
-However, legacy applications are generally sequential and therefore not easily scalable to multi-core architectures and parallel computing platforms.
+In this context, the approach of using Compilation Directives to guide the
+compiler in the parallelization process has gained popularity. These directives
+are implemented using the compiler's pre-processing directives and are utilized
+as annotations that provide tips about the sequential code. Among the tools and
+extensions that use compilation directives are the \texttt{OpenMP} API
+\cite{Dagum1998a} \cite{Chapman:2007} used for writing programs for multi-core
+architectures, and the \texttt{OpenACC} programming
+standard~\cite{openacc:api}, used for writing programs for heterogeneous
+CPU/GPU architectures. The \texttt{OpenMP 4.0} specification supports
+offloading capabilities \cite{openmp:api:2013} such as \texttt{OpenACC} and
+\texttt{AMD HSA}~\cite{amd:hsa:site}. These tools aim to make easier the task
+of designing and implementing efficient and provably correct parallel
+algorithms. However, the abstraction layers built by the extensions also have
+the potential to conceal architectural details and to generate new parallel
+programming challenges.
 
-Asanovic \textit{et al.} \cite{Asanovic2009} describes this scenario, and defends the need for Structural and Computational Patterns in Parallel Computing.
-These patterns allow the portability of parallel programming solutions, while also allowing specialization of applications with the same behavior.
-In some situations it is not a matter of achieving the highest possible performance, but instead of making the application achieve acceptable performance, benefiting from a new parallel computing platform~\cite{Asanovic2009}.
-A similar approach is proposed by Pllana \textit{et al.} in a development environment that uses a combination of Model-Driven Development and Parallel Building Blocks \cite{Pllana:2009}.
-The approach consists in creating the model and then generating or transforming the code using \emph{model-to-source} tools.
+When comparing \texttt{OpenMP} to other available tools for developing
+multi-threaded applications such as the \texttt{pthreads} interface, the API
+can be considered almost costless. However, when writing more complex
+applications this may not be true. The design and implementation of such
+complex applications require more knowledge about platforms and execution
+models than what can be shown by tutorials, that present high level and trivial
+examples. These tutorials try to convince the user that programming with
+directives is simple and easy to do, which is not always true. For instance, to
+achieve good performance in some cases it is necessary to describe many
+restrictions on annotations~\cite{OpenMPTasks2009}~\cite{mattson2003good}.
 
-In this context, the approach of using Compilation Directives to guide the compiler in the parallelization process has gained popularity.
-These directives are implemented using the compiler's pre-processing directives and are utilized as annotations that provide tips about the sequential code.
-Among the tools and extensions that use compilation directives are the \texttt{OpenMP} API \cite{Dagum1998a} \cite{Chapman:2007} used for writing programs for multi-core architectures, and the \texttt{OpenACC} programming standard~\cite{openacc:api}, used for writing programs for heterogeneous CPU/GPU architectures.
-The \texttt{OpenMP 4.0} specification supports offloading capabilities \cite{openmp:api:2013} such as \texttt{OpenACC} and \texttt{AMD HSA}~\cite{amd:hsa:site}.
-These tools aim to make easier the task of designing and implementing efficient and provably correct parallel algorithms.
-However, the abstraction layers built by the extensions also have the potential to conceal architectural details and to generate new parallel programming challenges.
+Krawezik and Cappello~\cite{CPE:CPE905} evidenced this difficulty by comparing
+MPI and three OpenMP versions of benchmarks. The most popular use of
+\texttt{OpenMP} directives is at the loop level. The programmer discovers
+potentially parallel loops and annotates the code with directives starting and
+closing parallel regions. Good results were obtained only when SPMD was
+implemented on OpenMP, and achieving good performance required a significant
+programming effort~\cite{CPE:CPE905}.
 
-When comparing \texttt{OpenMP} to other available tools for developing multi-threaded applications such as the \texttt{pthreads} interface, the API can be considered almost costless.
-However, when writing more complex applications this may not be true.
-The design and implementation of such complex applications require more knowledge about platforms and execution models than what can be shown by tutorials, that present high level and trivial examples.
-These tutorials try to convince the user that programming with directives is simple and easy to do, which is not always true.
-For instance, to achieve good performance in some cases it is necessary to describe many restrictions on annotations~\cite{OpenMPTasks2009}~\cite{mattson2003good}.
+In this work we have gathered data from assignments made by graduate students
+for the \emph{Introduction to Parallel and Distributed Computing} course. Among
+other exercises, the students were asked to find examples of \texttt{OpenMP}
+code in tutorials and official manuals that failed under special conditions.
+These conditions could be created by, for example, placing barriers or forcing
+racing conditions by exposing inaccurate thread management.
 
-Krawezik and Cappello~\cite{CPE:CPE905} evidenced this difficulty by comparing MPI and three OpenMP versions of benchmarks.
-The most popular use of \texttt{OpenMP} directives is at the loop level.
-The programmer discovers potentially parallel loops and annotates the code with directives starting and closing parallel regions.
-Good results were obtained only when SPMD was implemented on OpenMP, and achieving good performance required a significant programming effort~\cite{CPE:CPE905}.
-
-In this work we have gathered data from assignments made by graduate students for the \emph{Introduction to Parallel and Distributed Computing} course.
-Among other exercises, the students were asked to find examples of \texttt{OpenMP} code in tutorials and official manuals that failed under special conditions.
-These conditions could be created by, for example, placing barriers or forcing racing conditions by exposing inaccurate thread management.
-
-We argue that it is necessary to consider fundamental concepts of architectures and parallel programming techniques when teaching students new to the subject, so that they can learn when and how to apply the programming models~\cite{ben1999thinking}.
-The abstractions provided by parallel programming tools over concepts such as threads, barriers and loop parallelization should be carefully studied, analysed and conceptually built in the classroom before they are used by the students in their assignments.
-
-We found eight tutorials with different kinds of \texttt{OpenMP} errors in the experiment with the students' assignments.
-Many of these errors where similar to the most common \texttt{OpenMP} mistakes~\cite{SuB:2005:CMO:1892830.1892863}.
-The most common type of error identified in our experiment was the management of shared or private variables.
-This problem is common across multiple programming paradigms.
-
-The \textbf{main contribution of this work} is to show the importance of teaching core architecture and parallel programming concepts thoroughly, despite the availability of powerful tools such as \texttt{OpenMP}.
-We also aim to show that the \texttt{OpenMP} API is not as easy to use as it might seem.
-
-The rest of the paper is organized as follows.
-The following section discusses related work.
-Section~\ref{sec:concurrency} discusses concurrency and other concepts in parallel computing.
-Section~\ref{sec:directives} presents the idea of compilation directives.
-Section~\ref{sec:methodology} discusses the methodology applied in the experiments, and Section~\ref{sec:results} presents our results.
-Section~\ref{sec:students:perception} presents the perceptions of the graduate students regarding the exercise and learning \texttt{OpenMP}.
-Finally, Section~\ref{sec:conclusions} presents the conclusions and next steps.
-
-% Institution 2 = University of São Paulo
+The rest of the paper is organized as follows. The following section discusses
+related work...

### 2017-07-13 sections/02-background.tex
Re-organizing and fixing the used template

\section{Background}
\label{sec:background}

TO-DO ...

### 2017-07-13 sections/03-relatedwork.tex
Re-organizing and fixing the used template

--- 2017-05-11 relatedwork.tex
@@
 
-There has been a large body of works that have presented results of studies and experiments in teaching parallel computing courses.
-S\"{u}\ss~and Leopold~\cite{SuB:2005:CMO:1892830.1892863} discuss the teaching of parallel computing, in particular the use of \texttt{OpenMP}.
-The authors present the most common errors identified in assignments given in parallel computing classes.
-The errors were classified in two categories: \emph{correctness mistakes} and \emph{performance mistakes}.
-Errors that impact the correctness of a program are errors of the first category.
-Errors which make programs perform slower are classified in the second category.
-The work also provides a checklist containing tips to avoid the most common mistakes.
-
-Falcao~\cite{6565518} defends that teaching a parallel programming model or interface such as \texttt{OpenMP} in the beginning of undergraduate courses is not much harder than teaching a language like C or C++.
-The benefits of teaching these skills early on in the course would be interesting, considering that the majority of computer architectures are now are multi-core.
-Parallel computing should be taught with the use of higher level tools.
-One of the approaches proposed by Ferner \textit{et al.}~\cite{6651019} is to use compiler directives to describe how a program should be parallelized.
-
-The need for Structural and Computational Patterns to Parallel Computing is defended by Asanovic \textit{et al.}~\cite{Asanovic2009} as the way to allow portability of solutions and the specialization of applications with the same behavior.
-Using patterns to design applications make it easy to change the target platform.
-The applications will be bound to platforms in low levels of architecture and probably inside a specialized function which can be easily changed.
-
-A similar approach is proposed by Pllana \textit{et al.}~\cite{Pllana:2009} in a programming environment that uses a combination of Model-Driven Development and Parallel Building Blocks.
-The idea is to create the model and then generate or transform the code using model-to-source tools.
-The proposed environment uses intelligent agents to assist the programmer, this is done during the high level program composition and when the programmer loads a BLAS function for some dense linear algebra operation, the tool suggests the use of the appropriate parallel basic block.
-The parallelism is inside the parallel building blocks and the programmer is not exposed directly to the complexity of parallel programming problems.
-
-Speyer \textit{et al.}~\cite{4755913} defend that implementing applications with high performance is harder than ever, specially due to the emergence of multi-core architectures.
-This leads to the development of hybrid applications that use both the \texttt{OpenMP} and \texttt{MPI} (\emph{Message Passing Interface}) APIs.
-This presents even greater challenges to the programmers, since they now need to know how to use both code annotations and message passing abstractions to achieve high performance.
-The paper then compares different parallel programming tools with regard to their ease of use.
-There is a substantial effort to start writing parallel and high performance code even with these tools.
-
-Some problems can be solved using different directive combinations, and the chosen combination influences the code that is generated.
-The \texttt{Parallware}~\cite{Parallware:2014} source-to-source compiler showed how different strategies and uses of directive combinations can lead to different performance results.
-The use of the wrong \texttt{OpenMP} directives and strategies can significantly reduce the performance over a sequential version.
-
-The \texttt{OmpVerify}~\cite{Basupalli:2011} system implements a static analysis tool based on the polyhedral model to extract the parallel portions of a program automatically and to detect important classes of common data-race errors in \texttt{OpenMP} \texttt{parallel for} directives.
-Saillard \textit{et al.}~\cite{Saillard:2014} statically validates barriers and work sharing constructs to avoid the improper use of directives that could cause deadlocks.
-S\"{u}\ss~and Leopold~\cite{Leopold:userOpenMP}, present an user's experience with parallel sorting and OpenMP.
-They have used several versions of a simple parallel sorting algorithm to show some weaknesses of OpenMP.
-
-Substantial efforts have been made to achieve the simplest possible way to parallelize code, and the task is not easy.
-In this paper we present an exploratory analysis of common \texttt{OpenMP} errors, confirming that programming with high-level APIs and abstractions can still be difficult.
+TO-DO ...

### 2017-07-13 sections/04-researchdesign.tex
Re-organizing and fixing the used template

\section{Research Design}
\label{sec:researchdesing}

The motivation to write this paper comes from experiences on teaching. During the classes of Distributed Systems and Parallel Programming, a graduate course at University of São Paulo...

### 2017-07-13 seps-omp-paper.tex
Re-organizing and fixing the used template

\include{configurations}

\begin{document}
%---
\title{OpenMP is Not as Easy as It Appears, \\
      but It is the Easiest One}
%---
\author{
    \IEEEauthorblockN{Pedro Bruel\IEEEauthorrefmark{1},
                      Paulo Meirelles\IEEEauthorrefmark{1} and
                      Alfredo Goldman\IEEEauthorrefmark{1}}
    \IEEEauthorblockA{\IEEEauthorrefmark{1}Software Systems Laboratory -- LSS\\
    Institute of Mathematics and Statistics -- IME\\
    University of São Paulo -- USP\\
    São Paulo/SP, Brazil\\
    \{phrb,paulormm,gold\}@ime.usp.br}
}
%---
\maketitle
%---
\input{sections/00-abstract}
%---
\IEEEpeerreviewmaketitle
%---
\input{sections/01-introduction}
\input{sections/02-background}
\input{sections/03-relatedwork}
\input{sections/04-researchdesign}
\input{sections/05-dataanalysis}
\input{sections/06-lessons}
\input{sections/07-conclusions}
%---
\input{sections/acknowledgments}
%---
\bibliographystyle{IEEEtran}
\bibliography{IEEEabrv,seps-omp-paper}
\end{document}

### 2017-07-21 sections/00-abstract.tex
Migrating to ACM sigplan format

\begin{abstract}
This paper provides a sample of a \LaTeX\ document which conforms,
somewhat loosely, to the formatting guidelines for
ACM SIG Proceedings. 
\end{abstract}

%
% The code below should be generated by the tool at
% http://dl.acm.org/ccs.cfm
% Please copy and paste the code instead of the example below. 
%
\begin{CCSXML}
<ccs2012>
 <concept>
  <concept_id>10010520.10010553.10010562</concept_id>
  <concept_desc>Computer systems organization~Embedded systems</concept_desc>
  <concept_significance>500</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010575.10010755</concept_id>
  <concept_desc>Computer systems organization~Redundancy</concept_desc>
  <concept_significance>300</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010553.10010554</concept_id>
  <concept_desc>Computer systems organization~Robotics</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
 <concept>
  <concept_id>10003033.10003083.10003095</concept_id>
  <concept_desc>Networks~Network reliability</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
</ccs2012>  
\end{CCSXML}

\ccsdesc[500]{Computer systems organization~Embedded systems}
\ccsdesc[300]{Computer systems organization~Redundancy}
\ccsdesc{Computer systems organization~Robotics}
\ccsdesc[100]{Networks~Network reliability}

\keywords{ACM proceedings, \LaTeX, text tagging}

### 2017-07-21 seps-omp-paper.tex
Migrating to ACM sigplan format

@@
 %---
-\title{OpenMP is Not as Easy as It Appears, \\
-      but It is the Easiest One}
+\title{P-Threads strikes back}
+%\titlenote{...}
+\subtitle{OpenMP is Not as Useful as It Appers}
+%\subtitlenote{...}
 %---
-\author{
-    \IEEEauthorblockN{Pedro Bruel\IEEEauthorrefmark{1},
-                      Paulo Meirelles\IEEEauthorrefmark{1} and
-                      Alfredo Goldman\IEEEauthorrefmark{1}}
-    \IEEEauthorblockA{\IEEEauthorrefmark{1}Software Systems Laboratory -- LSS\\
-    Institute of Mathematics and Statistics -- IME\\
-    University of São Paulo -- USP\\
-    São Paulo/SP, Brazil\\
-    \{phrb,paulormm,gold\}@ime.usp.br}
+\author{Pedro Bruel}
+%\authornote{This author is the one who did all the really hard work.}
+\affiliation{%
+  \institution{Software Systems Laboratory (LSS)\\ Institute of Mathematics and Statistics\\ University of São Paulo}
+  \streetaddress{Rua do Matão, 1010}
+  \city{São Paulo} 
+  \country{Brazil}
+  \postcode{05508-090}
 }
+\email{phrb@ime.usp.br}
 %---
-\maketitle
+\author{Paulo Meirelles}
+%\authornote{Posdoc}
+\affiliation{%
+  \institution{FLOSS Competence Center (CCSL)\\ Institute of Mathematics and Statistics\\ University of São Paulo}
+  \streetaddress{Rua do Matão, 1010}
+  \city{São Paulo} 
+  \country{Brazil}
+  \postcode{05508-090}
+}
+\email{paulormm@ime.usp.br}
 %---
+\author{Alfredo Goldman}
+%\authornote{Professor: the head.}
+\affiliation{%
+  \institution{Software Systems Laboratory (LSS)\\ Institute of Mathematics and Statistics\\ University of São Paulo}
+  \streetaddress{Rua do Matão, 1010}
+  \city{São Paulo} 
+  \country{Brazil}
+  \postcode{05508-090}
+}
+\email{gold@ime.usp.br}
+
+%---
+
+\renewcommand{\shortauthors}{Bruel et al.}
+
+%--
 \input{sections/00-abstract}
 %---
-\IEEEpeerreviewmaketitle
+
+\maketitle
+
 %---
@@
 %---
-\bibliographystyle{IEEEtran}
-\bibliography{IEEEabrv,seps-omp-paper}
+
+\bibliographystyle{ACM-Reference-Format}
+\bibliography{seps-omp-paper}
+
 \end{document}

### 2017-07-21 sections/04-researchdesign.tex
Adding initial research questions

@@
 The motivation to write this paper comes from experiences on teaching. During the classes of Distributed Systems and Parallel Programming, a graduate course at University of São Paulo...
+
+\begin{description}
+\item[RQ1:] \textit{Which API is easier to learn and use for beginning students?}
+%
+...
+\item[RQ2:] \textit{Which API is better to improve the performance for beginning students?}
+%
+...
+\item[RQ3:] \textit{Which API is easier to learn and use for advanced students?}
+%
+...
+\item[RQ4:] \textit{Which API is better to improve the performance for advanced students?}
+%
+...
+
+\end{description}

### 2017-07-25 data/plot.py
Add images for data

@@ -1,0 +1,359 @@
+#! /usr/bin/python
+
+import re
+import os
+import matplotlib as mpl
+
+import math
+import numpy
+
+import csv
+
+mpl.use('agg')
+
+import matplotlib.pyplot as plt
+import numpy as np
+
+def config_matplotlib():
+    plt.rc('text', usetex = True)
+    plt.rc('font', family = 'serif')
+
+    font = {'family' : 'serif',
+            'size'   : 28}
+
+    mpl.rc('font', **font)
+
+def autolabel(rects, ax):
+    for rect in rects:
+        height = rect.get_height()
+        ax.text(rect.get_x() + rect.get_width() / 2., height + .03,
+                '%.2f' % height,
+                ha = 'center', va = 'bottom', rotation = '0')
+
+def plot_bar_stacks(data,
+                    xlabel,
+                    ylabel,
+                    index_range,
+                    width,
+                    tick_labels,
+                    file_title,
+                    title,
+                    ymin,
+                    ymax,
+                    color_map):
+
+    indexes = np.arange(index_range)
+    fig     = plt.figure(1, figsize=(18, 10))
+    ax      = fig.add_subplot(111)
+
+    rects   = []
+
+    legend_names = []
+    legend_rects = []
+
+    previous_bottom = np.zeros(len(list(data.values())[0]))
+
+    cmap = mpl.cm.get_cmap(color_map)
+    colors = len(list(data.keys()))
+    c_index = 0
+
+    for key, value in data.items():
+        rect = ax.bar(indexes, value, width, color = cmap(1. * (c_index / colors)), edgecolor = 'lightgrey', bottom = previous_bottom)
+
+        c_index += 1
+
+        previous_bottom += value
+
+        legend_rects.append(rect[0])
+        legend_names.append(key)
+
+        rects.append(rects)
+
+    ax.set_title(title)
+    ax.set_xlabel(xlabel)
+    ax.set_xticks(indexes)
+    ax.set_ylabel(ylabel)
+    ax.set_xticklabels(tick_labels, rotation = '15', ha = 'right')
+
+    ax.set_ylim([ymin, ymax])
+
+    legend = ax.legend(tuple(legend_rects), tuple(legend_names),
+                        loc = 9, bbox_to_anchor = [0.5, 1.15], ncol = len(list(data.keys())), shadow = True,
+                        title = "", fancybox = True)
+
+    #autolabel(rects, ax)
+
[…]

### 2017-07-26 data/plot.py
Add new image, fix 'plot.py' permissions.

@@ -227,9 +227,29 @@ if __name__ == '__main__':
              "Technology",
              "Previous Knowledge",
              len(background_mean_data),
-             .9,
+             1.,
              [i.title() for i in background],
              "background_mean_knowledge",
+             "",
+             1,
+             5,
+             False,
+             "viridis")
+
+    learning_mean_data = []
+
+    for tech in learning:
+        learning_mean_data.append(np.mean([float(i) for i in data[tech]]))
+
+    print(learning_mean_data)
+
+    plot_bar(learning_mean_data,
+             "Technology",
+             "Difficulty to Learn",
+             len(learning_mean_data),
+             1.,
+             [i.replace("_", "\\_") for i in learning],
+             "learning_mean_difficulty",
              "",
              1,
              5,
@@ -254,7 +274,7 @@ if __name__ == '__main__':
              "",
              "Percentage of Students",
              len(classes_norm_data),
-             .9,
+             1.,
              [i.replace("_", "\\_") for i in classes],
              "yes_no_questions",
              "",

### 2017-07-26 data/plot.py
Add data on performance of student's assignments

@@ -40,7 +40,8 @@ def plot_bar_stacks(data,
                     title,
                     ymin,
                     ymax,
-                    color_map):
+                    color_map,
+                    xticks_rotation = '15'):
 
     indexes = np.arange(index_range)
     fig     = plt.figure(1, figsize=(18, 10))
@@ -73,7 +74,7 @@ def plot_bar_stacks(data,
     ax.set_xlabel(xlabel)
     ax.set_xticks(indexes)
     ax.set_ylabel(ylabel)
-    ax.set_xticklabels(tick_labels, rotation = '15', ha = 'right')
+    ax.set_xticklabels(tick_labels, rotation = xticks_rotation, ha = 'right')
 
     ax.set_ylim([ymin, ymax])
 
@@ -214,7 +215,7 @@ if __name__ == '__main__':
     classes    = ['had_concepts', 'knew_apis', 'helped_openmp_class', 'helped_pthreads_class']
     learning   = ['learn_openmp', 'learn_pthreads']
     using      = ['par_loop_openmp', 'par_loop_pthreads', 'par_nested_loop_openmp', 'par_nested_loops_pthreads', 'improve_seq_openmp', 'improve_seq_pthreads']
-    comparing  = ['harder_learn', 'simpler_tech', 'easier_tech', 'easier_par_loop', 'easier_nested_par_loop', 'easier_improve_seq']
+    comparing  = ['harder_learn', 'simpler_tech', 'easier_tech', 'easier_par_loop', 'easier_nested_par_loop', 'easier_improve_seq', 'had_best_performance']
 
     background_mean_data = []
 
@@ -335,7 +336,8 @@ if __name__ == '__main__':
     comparing_data = {}
 
     levels = {"OpenMP": "OpenMP",
-              "POSIX Threads": "Pthreads"}
+              "POSIX Threads": "Pthreads",
+              "Equivalent": "Equivalent"}
 
     for subject in comparing:
         sub_data = data[subject]

### 2017-07-29 sections/04-researchdesign.tex
Re-defining initial research questions
Add figures to text, start to compose the section

--- 2017-07-13 sections/04-researchdesign.tex
@@
 The motivation to write this paper comes from experiences on teaching. During the classes of Distributed Systems and Parallel Programming, a graduate course at University of São Paulo...
+
+\begin{description}
+\item[RQ1:] \textit{Which API is easier to learn and use according to the student perceptions?}
+%
+...
+\item[RQ2:] \textit{Which API is better to improve the performance according to the student perceptions?}
+%
+...
+
+\end{description}
+
+\subsection{Student Background}
+
+\begin{figure}[htpb]
+    \centering
+    \includegraphics[width=0.85\columnwidth]{background_mean_knowledge}
+    \caption{Student mean knowledge \textit{before} the assignment}
+    \label{fig:student-mean-knowledge}
+\end{figure}
+
+\begin{figure}[htpb]
+    \centering
+    \includegraphics[width=0.85\columnwidth]{yes_no_questions}
+    \caption{Student API knowledge \textit{before} the assignment and their relation to classes}
+    \label{fig:student-knowledge-yes-no}
+\end{figure}

### 2017-07-29 sections/05-results.tex
Add figures to text, start to compose the section

\section{Results}
\label{sec:results}

\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.85\columnwidth]{likert_questions}
    \caption{Student responses to Likert-scale questions}
    \label{fig:}
\end{figure}

\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.85\columnwidth]{learning_mean_difficulty}
    \caption{Student mean difficulty to learn the APIs}
    \label{fig:}
\end{figure}

\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.85\columnwidth]{comparisons}
    \caption{Student responses on comparisons of APIs, and the performance of their usage of the APIs}
    \label{fig:}
\end{figure}

### 2017-07-29 data/plot.py

@@ -19,7 +19,7 @@ def config_matplotlib():
     plt.rc('font', family = 'serif')
 
     font = {'family' : 'serif',
-            'size'   : 28}
+            'size'   : 40}
 
     mpl.rc('font', **font)
 
@@ -79,7 +79,7 @@ def plot_bar_stacks(data,
     ax.set_ylim([ymin, ymax])
 
     legend = ax.legend(tuple(legend_rects), tuple(legend_names),
-                        loc = 9, bbox_to_anchor = [0.5, 1.15], ncol = len(list(data.keys())), shadow = True,
+                        loc = 9, bbox_to_anchor = [0.5, 1.2], ncol = len(list(data.keys())), shadow = True,
                         title = "", fancybox = True)
 
     #autolabel(rects, ax)
@@ -102,7 +102,9 @@ def plot_bar(data,
              ymin,
              ymax,
              line,
-             color_map):
+             color_map,
+             xticks_rotation = '0',
+             align = 'center'):
 
     indexes = np.arange(index_range)
     fig     = plt.figure(1, figsize=(18, 10))
@@ -121,6 +123,7 @@ def plot_bar(data,
     ax.set_xticks(indexes)
     ax.set_ylabel(ylabel)
     ax.set_xticklabels(tick_labels, rotation = '0')
+    ax.set_xticklabels(tick_labels, rotation = xticks_rotation, ha = align)
 
     ax.set_ylim([ymin, ymax])
 
@@ -235,7 +238,7 @@ if __name__ == '__main__':
              1,
              5,
              False,
-             "viridis")
+             "Greys")
 
     learning_mean_data = []
 
@@ -255,7 +258,7 @@ if __name__ == '__main__':
              1,
              5,
              False,
-             "viridis")
+             "Greys")
 
     classes_norm_data = []
 
@@ -282,7 +285,9 @@ if __name__ == '__main__':
              0,
              100,
              False,
-             "viridis")
+             "Greys",
+             xticks_rotation = '15',
+             align = 'right')
 
     using_data = {}
 

### 2017-08-01 sections/03-apis.tex
Add text to responses section, improve figures

\section{OpenMP \& Pthreads}
\label{sec:apis}

\todo[inline,color=cyan,author=Pedro]{Provide brief descriptions of the APIs}

### 2017-08-01 sections/04-research-design.tex
Add text to responses section, improve figures

\section{Research Design}
\label{sec:researchdesign}

\todo[inline,color=cyan,author=Pedro]{Describe research design}
\todo[inline,color=cyan,author=Pedro]{Improve RQ descriptions}

\begin{description}
\item[RQ1:] \textit{Which API is easier to learn and use according to the student perceptions?}
\item[RQ2:] \textit{Which API is better to improve the performance according to the student perceptions?}
\end{description}

\subsection{Students' Background}

\todo[inline,color=cyan,author=Pedro]{Describe students' background}

\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.85\columnwidth]{background_questions}
    \caption{Student mean knowledge \textit{before} the assignment}
    \label{fig:background}
\end{figure}

\begin{table}[htpb]
    \centering
    \begin{tabular}{@{}p{0.9\columnwidth}p{0.08\columnwidth}@{}}
        \toprule
        \multicolumn{1}{c}{\scriptsize{Have you$\dots$}} & \textnumero \\ \midrule
        \scriptsize{Had contact with parallel and distributed concepts before?} & $(1)$ \\
        \scriptsize{Had contat with the APIs before?} & $(2)$ \\
        \scriptsize{Learned from the OpenMP class?} & $(3)$ \\
        \scriptsize{Learned from the Pthreads class?} & $(4)$ \\ \bottomrule
    \end{tabular}
    \caption{Questions for Figure \ref{fig:classes}}
    \label{tab:classes}
\end{table}

\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.85\columnwidth]{classes_questions}
    \caption{Student API knowledge \textit{before} the assignment and their relation to classes}
    \label{fig:classes}
\end{figure}

### 2017-08-01 sections/05-responses.tex
Add text to responses section, improve figures

\section{Students' Responses}
\label{sec:responses}

This section presents the responses given by the students to our
questionnaire. The questions are listed in Tables \ref{tab:likert}
and \ref{tab:comparisons}.

\todo[inline,color=cyan,author=Pedro]{Add reference for questionnaires from Herez}
\todo[inline,color=cyan,author=Pedro]{Improve figure colorscheme}

The first set of questions used a Likert-type Scale to measure the relationship
of the students with the two APIs in the context of the assignment. The Likert
Scale~\cite{likert1932technique} enables respondents to express their agreement
or disagreement with specific statements. Respondents choose between five
options: \textit{Strongly Agree} (SA), \textit{Agree} (A), \textit{Neutral}
(N), \textit{Disagree} (D) and \textit{Strongly Disagree}.

The questions of the first set are listed in Table \ref{tab:likert}. The
students expressed their agreement or disagreement with statements related to
the difficulty of parallelizing the first loop with independent iterations,
parallelizing the two nested loops with independent iterations and improving
the performance of the sequential code. Each question was repeated for
\textit{OpenMP} and \textit{Pthreads}, in a total of $6$ questions.

\begin{table}[htpb]
    \centering
    \begin{tabular}{@{}p{0.7\columnwidth}p{0.1\columnwidth}p{0.05\columnwidth}@{}}
        \toprule
        \multicolumn{2}{c}{\scriptsize{It is easy to$\dots$}} & \textnumero \\ \midrule
        \multirow{2}{*}{\parbox{0.7\columnwidth}{\scriptsize{Parallelize loops with independent iterations using:}}} & \scriptsize{OpenMP} & $(1)$ \\
        & \scriptsize{Pthreads} & $(2)$ \\
        \addlinespace{}
        \multirow{2}{*}{\parbox{0.7\columnwidth}{\scriptsize{Parallelize nested loops with independent iterations using:}}} & \scriptsize{OpenMP} & $(3)$ \\
        &  \scriptsize{Pthreads} & $(4)$ \\
        \addlinespace{}
        \multirow{2}{*}{\parbox{0.7\columnwidth}{\scriptsize{Improve the performance of sequential code using:}}} & \scriptsize{OpenMP} & $(5)$  \\
        &  \scriptsize{Pthreads} & $(6)$ \\ \bottomrule
    \end{tabular}
    \caption{Likert-like Scale questions for Figure \ref{fig:likert}}
    \label{tab:likert}
\end{table}

Figure \ref{fig:likert} shows a visualization of the students' responses.  The
height of each stacked sub-column represents the percentage of respondents that
selected each of the correspondent Likert scale agreement levels, which are
represented by different colors.

\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.85\columnwidth]{likert_questions}
    \caption{Student responses to the Likert Scale questions in Table \ref{tab:likert}}
    \label{fig:likert}
\end{figure}

The percentages of \textit{Strongly Agree} (SA) and \textit{Agree} (A)
responses are significantly larger for questions $(1)$, $(3)$ and $(5)$, which
are related to \textit{OpenMP}.  Likewise, the percentages of \textit{Disagree}
(D) responses are significantly larger for questions $(2)$, $(4)$ and $(6)$,
which are related to \textit{Pthreads}. These responses lead to the conclusion
that \textbf{the perception of the students was that it was easier to use
\textit{OpenMP} than \textit{Pthreads}} to parallelize and improve the
performance of the sequential code provided in the assignment.

The second set of questions used a five-point scale to measure the difficulty
the students had with the learning process of the two APIs for the assignment.
Respondents expressed their difficulty to learn \textit{OpenMP} and
\textit{Pthreads} in a scale of $1$ to $5$, choosing a point between \textit{No
Difficulty} $(1)$ and \textit{Much Difficulty} $(5)$.

Figure \ref{fig:learning} shows a visualization of the students' responses.
The height of each stacked sub-column represents the percentage of respondents
that selected each of the correspondent difficulty levels, which are
represented by different colors.

\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.85\columnwidth]{learning_difficulty_questions}
    \caption{Student difficulty to learn the APIs in a five-point scale}
    \label{fig:learning}
\end{figure}

The percentages of \textit{No Difficulty} $(1)$ and difficulty level $(2)$
responses are significantly larger for learning \textit{OpenMP}. Likewise, the
percentages of difficulty level $(4)$ and \textit{Much Difficulty} $(5)$
responses are significantly larger for learning \textit{Pthreads}.  We derive
from these responses the conclusion that \textbf{the students perceived less
difficulty to learn \textit{OpenMP} than \textit{Pthreads}} for the purposes of
the assignment.

The final set of questions regarding the usage and learning of \textit{OpenMP}
and \textit{Pthreads} in the context of the first assignment presented
comparison questions between the two APIs. Students chose which API they
perceived was the best answer to the questions in Table \ref{tab:comparisons}.

Responses to question $(13)$ in Table \ref{tab:comparisons} were not obtained
directly from the students. We analysed each assignment submission and
evaluated the performance of the each solution across the specified
experimental settings.

\begin{table}[htpb]
    \centering
    \begin{tabular}{@{}p{0.9\columnwidth}p{0.08\columnwidth}@{}}
        \toprule
        \multicolumn{1}{c}{\scriptsize{Which technology$\dots$}} & \textnumero \\ \midrule
        \scriptsize{Was more difficult to learn?} & $(7)$ \\
        \scriptsize{Was simpler to use?} & $(8)$ \\
        \scriptsize{Was easier to use?} & $(9)$ \\
        \scriptsize{Was easier to use for parallelizing loops with independent iterations?} & $(10)$ \\
        \scriptsize{Was easier to use for parallelizing nested loops with independent iterations?} & $(11)$  \\
        \scriptsize{Was easier to use for improving the performance of a sequential program?} & $(12)$  \\
        \scriptsize{Had the best performance on your assignment?} & $(13)$ \\ \bottomrule
    \end{tabular}
    \caption{Questions for Figure \ref{fig:comparisons}}
    \label{tab:comparisons}
\end{table}

Figure \ref{fig:comparisons} shows a visualization of the students' responses.
The height of each stacked sub-column represents the percentage of respondents
that selected each API, which are represented by different colors.

The percentage of \textit{OpenMP} selections is significantly larger for all
questions, except for questions $(7)$ and $(13)$. The responses for questions
$(7)$ -- $(12)$ strengthen the previous conclusion that the perception of the
students was that it was easier to use \textit{OpenMP} than \textit{Pthreads}
in the context of the assignment. The performance results that we obtained from
analysing the students' submissions challenge the perception of the students.
The responses to question $(13)$, extracted by our analysis, show that
\textbf{the students achieved better performance when using \textit{Pthreads}
than \textit{OpenMP}}.

\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.85\columnwidth]{comparisons}
    \caption{Student responses to the questions in Table \ref{tab:comparisons}}
    \label{fig:comparisons}
\end{figure}

\todo[inline,color=cyan,author=Pedro]{Improve these last paragraphs}

The responses presented in this section enable the conclusion that the students
perceived less difficulty using and learning \textit{OpenMP} in the context
of the assignment. They also reported the perception that using \textit{OpenMP}
made it easier to improve the performance of the sequential code we provided.
Our analysis of the submissions revealed that the majority of the students
($52.2\%$) achieved better performance in the assignment with either
\textit{OpenMP} ($17.4\%$) or \textit{Pthreads} ($34.8\%$). Contradicting the
perception of the students, the implementation using \textit{Pthreads} achieved
better performance in twice the submissions.

This shows that the perceptions of usability and performance improvement
regarding \textit{OpenMP} did not reflect in the actual implementations and
submissions. We believe that \textit{OpenMP} is still not as easy as it appears
to be, and that the students' responses strengthen the argument for teaching
lower-level technologies for parallel and distributed computing.

### 2017-08-01 data/plot.py

@@ -19,7 +19,7 @@ def config_matplotlib():
     plt.rc('font', family = 'serif')
 
     font = {'family' : 'serif',
-            'size'   : 40}
+            'size'   : 36}
 
     mpl.rc('font', **font)
 
@@ -41,7 +41,8 @@ def plot_bar_stacks(data,
                     ymin,
                     ymax,
                     color_map,
-                    xticks_rotation = '15'):
+                    xticks_rotation = '15',
+                    align = 'right'):
 
     indexes = np.arange(index_range)
     fig     = plt.figure(1, figsize=(18, 10))
@@ -74,7 +75,7 @@ def plot_bar_stacks(data,
     ax.set_xlabel(xlabel)
     ax.set_xticks(indexes)
     ax.set_ylabel(ylabel)
-    ax.set_xticklabels(tick_labels, rotation = xticks_rotation, ha = 'right')
+    ax.set_xticklabels(tick_labels, rotation = xticks_rotation, ha = align)
 
     ax.set_ylim([ymin, ymax])
 
@@ -229,7 +230,7 @@ if __name__ == '__main__':
 
     plot_bar(background_mean_data,
              "Technology",
-             "Previous Knowledge",
+             "Mean Previous Knowledge",
              len(background_mean_data),
              1.,
              [i.title() for i in background],
@@ -240,6 +241,57 @@ if __name__ == '__main__':
              False,
              "Greys")
 
+    background_data = {}
+
+    levels = {"1": "1",
+              "2": "2",
+              "3": "3",
+              "4": "4",
+              "5": "5"}
+
+    for subject in background:
+        sub_data = data[subject]
+        background_data[subject] = {}
+
+        for response in sub_data:
+            try:
+                background_data[subject][levels[response]] += 1
+            except KeyError:
+                background_data[subject][levels[response]] = 1
+
+        for key in background_data[subject].keys():
+            background_data[subject][key] *= 100 / len(sub_data)
+
+    print(background_data)
+
+    stacked_background_data = {}
+
+    for value in levels.values():
+        for key in background_data.keys():
+            if value not in background_data[key].keys():
+                background_data[key][value] = 0
+            try:
+                stacked_background_data[value].append(background_data[key][value])
+            except KeyError:
+                stacked_background_data[value] = [background_data[key][value]]
+
+
+    plot_bar_stacks(stacked_background_data,
+                    "",
+                    "Percentage of Responses",
+                    len(stacked_background_data["1"]),
+                    1.,
[…]

### 2017-08-01 sections/04-research-design.tex
Improve responses section.

@@
 \todo[inline,color=cyan,author=Pedro]{Describe research design}
-\todo[inline,color=cyan,author=Pedro]{Improve RQ descriptions}
+
+\subsection{The Concurrent, Parallel and Distributed Computing Course}
+
+\todo[inline,color=cyan,author=Pedro]{Describe the course}
+
+\subsection{Questionnaire}
+
+\todo[inline,color=cyan,author=Pedro]{Describe the questionnaire}
+\todo[inline,color=cyan,author=Pedro]{Add reference for questionnaires from Herez}
+
+\subsection{Research Questions}
+
+\todo[inline,color=cyan,author=Pedro]{Improve RQ discussion and descriptions}
+
+According to the students' perception, which API:
 
 \begin{description}
-\item[RQ1:] \textit{Which API is easier to learn and use according to the student perceptions?}
-\item[RQ2:] \textit{Which API is better to improve the performance according to the student perceptions?}
+    \item[RQ1:] \textit{Was easier to learn and use?}
+    \item[RQ2:] \textit{Improved performance the most?}
 \end{description}
 
-\subsection{Students' Background}
+After evaluation of the students' performance improvements
+in the assignment:
 
-\todo[inline,color=cyan,author=Pedro]{Describe students' background}
-
-\begin{figure}[htpb]
-    \centering
-    \includegraphics[width=0.85\columnwidth]{background_questions}
-    \caption{Student mean knowledge \textit{before} the assignment}
-    \label{fig:background}
-\end{figure}
-
-\begin{table}[htpb]
-    \centering
-    \begin{tabular}{@{}p{0.9\columnwidth}p{0.08\columnwidth}@{}}
-        \toprule
-        \multicolumn{1}{c}{\scriptsize{Have you$\dots$}} & \textnumero \\ \midrule
-        \scriptsize{Had contact with parallel and distributed concepts before?} & $(1)$ \\
-        \scriptsize{Had contat with the APIs before?} & $(2)$ \\
-        \scriptsize{Learned from the OpenMP class?} & $(3)$ \\
-        \scriptsize{Learned from the Pthreads class?} & $(4)$ \\ \bottomrule
-    \end{tabular}
-    \caption{Questions for Figure \ref{fig:classes}}
-    \label{tab:classes}
-\end{table}
-
-\begin{figure}[htpb]
-    \centering
-    \includegraphics[width=0.85\columnwidth]{classes_questions}
-    \caption{Student API knowledge \textit{before} the assignment and their relation to classes}
-    \label{fig:classes}
-\end{figure}
+\begin{description}
+    \item[RQ3:] \textit{Did the students' performance improvements match their
+        perception?}
+\end{description}

### 2017-08-01 sections/05-responses.tex
Improve responses section.

@@
-\section{Students' Responses}
+\section{Students' Background \& Responses}
 \label{sec:responses}
 
-This section presents the responses given by the students to our
-questionnaire. The questions are listed in Tables \ref{tab:likert}
-and \ref{tab:comparisons}.
+In this section we present the background of our students in the technologies
+relevant to the course and to the assignment. Then we present the responses
+given by the students to our questionnaire. The questions are listed in Tables
+\ref{tab:likert} and \ref{tab:comparisons}.
 
-\todo[inline,color=cyan,author=Pedro]{Add reference for questionnaires from Herez}
 \todo[inline,color=cyan,author=Pedro]{Improve figure colorscheme}
 
+\subsection{Background}
+
+The course of Concurrent, Parallel and Distributed Computing is offered to
+students of both the graduate and undergraduate programs in Computer Science at
+the University of São Paulo. To gauge the proficiency and knowledge of our
+students we elaborated a five-point scale from \textit{Do not Know} $(1)$ to
+\textit{Expert or Proficient} $(5)$.  The students were asked to express their
+proficiency level on this scale, and their responses are shown in Figure
+\ref{fig:background}.
+
+\begin{figure}[htpb]
+    \centering
+    \includegraphics[width=0.85\columnwidth]{background_questions}
+    \caption{Student knowledge \textit{before} the assignment}
+    \label{fig:background}
+\end{figure}
+
+It is common for Computer Science introductory and advanced courses in the
+graduate and undergraduate programs of the University of São Paulo to be given
+in programming languages such as C, Java and Python. Figure
+\ref{fig:background} shows that students reported higher proficiency in those
+languages. The C language was the one with which students reported most
+proficiency.
+
+The students were asked to evaluate their proficiency in \textit{OpenMP} and
+\textit{Pthreads}, \textit{before} completing the assignment, with two sets of
+questions. The first set used the same five-point scale as the other background
+questions and is also shown in Figure \ref{fig:background}. The students
+reported significantly larger proficiency in \textit{Pthreads} than
+\textit{OpenMP}, and the majority of the students reported they did not know
+\textit{OpenMP} before the course.
+
+\begin{table}[htpb]
+    \centering
+    \begin{tabular}{@{}p{0.9\columnwidth}p{0.08\columnwidth}@{}}
+        \toprule
+        \multicolumn{1}{c}{\scriptsize{Have you$\dots$}} & \textnumero \\ \midrule
+        \scriptsize{Had contact with parallel and distributed concepts before?} & $(1)$ \\
+        \scriptsize{Had contat with the APIs before?} & $(2)$ \\
+        \scriptsize{Learned from the OpenMP class?} & $(3)$ \\
+        \scriptsize{Learned from the Pthreads class?} & $(4)$ \\ \bottomrule
+    \end{tabular}
+    \caption{Questions for Figure \ref{fig:classes}}
+    \label{tab:classes}
+\end{table}
+
+The second set of questions regarding student proficiency with the APIs
+comprised the 4 \textit{Yes or No} questions listed in Table \ref{tab:classes}.
+This set of questions also asked the students to evaluate whether the classes
+on each subject helped their understanding and learning of the technologies.
+The student responses for this set are shown in Figure \ref{fig:classes}.  The
+majority of the students have had previous contact with the concepts related to
+concurrent, parallel and distributed programming, but the majority also have
+not had contact or experience with \textit{OpenMP} or \textit{Pthreads}.  Most
+students reported that the classes on each API helped their learning.
+
+\begin{figure}[htpb]
+    \centering
+    \includegraphics[width=0.85\columnwidth]{classes_questions}
+    \caption{Student API knowledge \textit{before} the assignment and their relation to classes}
+    \label{fig:classes}
+\end{figure}
+
+\subsection{Responses}
+
+The students were asked to answer 3 sets of questions regarding their
+experience with \textit{OpenMP} and \textit{Pthreads}, \textit{after}
+completing the assignment.
 The first set of questions used a Likert-type Scale to measure the relationship

### 2017-08-04 sections/02-background.tex
Add text to responses section, improve figures
Change related work section name.

@@
 
-TO-DO ...
+%\todo[inline,color=cyan,author=Pedro]{Cite our previous paper, making it clear that we are the authors}
+%\todo[inline,color=cyan,author=Pedro]{Describe related work}
+%\todo[inline,color=cyan,author=Pedro]{Describe related work since last paper}
+
+The discipline of Parallel and Distributed Computing has evolved from an
+elective topic into one of the main curricular components in a Computer Science
+courses, as pointed out in \cite{acmcurricula}, thus, deciding what
+technology/framework to teach students in Parallel Computing class has a great
+impact on the student's ability to develop efficient solutions for algorithmic
+problems.
+
+There are several chain of thoughts on what is the best level of abstraction to
+teach parallel programming courses. In \cite{6565518}, Falcao argues that
+introducing a higher level parallel programming interface such as OpenMP in the
+beginning of a CS course would cause very little overhead and would bring
+great benefits for the rest of the course. The approach of introducing mostly
+practical learning experiences spread all over a CS course - increasing in
+depth as the student advances in classes was also supported by the authors in
+\cite{FOLEY2017138}. In their work they introduce a tool named OnRamp, a web
+portal that aims to help students learn how to write parallel code in a
+top-down approach, i.e., from practical experiences to more conceptual ones.
+
+Pllana et al. use a similar approach as described in \cite{Pllana:2009} where
+the authors adopt a Model Driven Development to build parallel solutions by
+means of Parallel Building Blocks. The authors describe a programming
+environment that takes these high level model and transform them into source
+code. In that way, the tool goal is to hide the complexity of writing parallel
+code away from the students.
+
+In contrast to these top-down approaches for teaching parallel programming
+there is the bottom-up approach on which, learning a software tool is just one
+of the objectives of a parallel programming class and that would come after the
+student are presented the concepts of parallel programming.
+
+In a previous investigation \cite{goncalves:OpenMPNotEasy}, we showed that
+first teaching the conceptual aspects of parallel programming instead of
+jumping into teaching higher level parallel programming interfaces such as
+OpenMP is of extreme importance mainly because the available
+resources/tutorials for learning such APIs may contain errors that would be
+otherwise difficult to find. In our work we listed several sources of
+tutorials/hands-on materials retrieved from the web on which under-grad
+students were able to point parallel correctness mistakes
+\cite{SuB:2005:CMO:1892830.1892863}.
+
+Another point that should be taken into consideration is that, not all
+parallel solutions can be expressed using tools such as OpenMP. This is the
+main topic discussed by Sü\ss and Leopold in \cite{Leopold:userOpenMP}, where
+they show that, by using a simple sorting algorithm as example, we can identify
+patterns that are difficult to capture using OpenMP and that could be easily
+handled using a lower level programming interface such as POSIX Threads.
+
+In \cite{ADAMS201731}, Adams also showed a set of patterns implemented using
+PThreads that help students understand some aspects of parallel programming
+that would be harder to grasp by using only OpenMP examples. His catalog of
+patterns also contains many solutions implemented using OpenMP.
+
+In this paper we argue that teaching a lower level parallel programming
+interface such as PThreds is equally important to teaching higher level
+interfaces such as OpenMP and both should be part of a complete parallel
+programming class curriculum like the one described by Brown et al. in
+\cite{Brown:2010:SPC:1971681.1971689}, where it has been pointed that the
+choice of a software tool should be only one of the knowledge areas that should
+be explored during a parallel programming course, along with data structure and
+algorithms; software design; and parallel hardware platforms.
+
+\todo[inline,color=orange,author=Raphael]{Maybe we should make a better
+connection between the end of this section and the next one;}

### 2017-08-04 sections/02-background.tex
Review text for background section.

--- 2017-07-13 sections/02-background.tex
@@
 
-TO-DO ...
+%\todo[inline,color=cyan,author=Pedro]{Cite our previous paper, making it clear that we are the authors}
+%\todo[inline,color=cyan,author=Pedro]{Describe related work}
+%\todo[inline,color=cyan,author=Pedro]{Describe related work since last paper}
+
+The selection of teaching tools for Parallel and Distributed classes impacts
+the development of students' abilities to solve algorithmic problems
+efficiently. The importance of selecting good tools for teaching Parallel and
+Distributed Computing has been increasingly important, especially since the
+topic became a core component of the ACM undergraduate computer science
+curricula in 2013~\cite{acmcurricula}.
+
+Different approaches have been used to study the selection of the best level of
+abstraction for teaching Parallel and Distributed programming courses.
+Falcao~\cite{6565518} argues that introducing a high-level parallel programming
+interface such as \textit{OpenMP} in the beginning of an undergraduate computer
+science course would cause a small overhead and benefit the rest of the course.
+Foley \textit{et al.}~\cite{FOLEY2017138} support the introduction of practical
+learning experiences along an undergraduate course, of increasing depth as the
+course progresses. They introduce OnRamp, a web application to help students
+learn to write parallel code starting with practical and moving to conceptual
+experiences.  Pllana \textit{et al.}~\cite{Pllana:2009} adopt a Model Driven
+Development approach to develop parallel solutions using Parallel Building
+Blocks. They introduce a programming environment that transforms these
+high-level models in source code. The tool's goal is to hide from the students
+the complexity of writing parallel code.
+
+The approaches of the last paragraph introduce tools for parallel programming
+before the related concepts.  Instead, some learning approaches
+consider learning software tools as just one of the objectives of the course,
+and argue that their introduction should come after the introduction of
+concepts of Parallel and Distributed Computing.
+
+In previous work~\cite{goncalves:OpenMPNotEasy} we show that teaching the
+concepts of parallel programming first instead of high-level parallel
+programming APIs is important, especially because the available resources and
+tutorials for these APIs may contain difficult to find errors.  We listed
+several sources of tutorials and hands-on materials obtained in the internet in
+which undergraduate students were able to point parallel correctness
+errors~\cite{SuB:2005:CMO:1892830.1892863}.
+
+Sü\ss{} and Leopold~\cite{Leopold:userOpenMP} discuss the inability to express
+all parallel solutions in tools such as \textit{OpenMP}. They used a sorting
+algorithm to show that it is easy to find patterns that are difficult to
+implement in \textit{OpenMP}, but can be easily implemented in lower-level
+programming interfaces, such as POSIX Threads.
+
+Adams~\cite{ADAMS201731} presented a set of patterns implemented using
+\textit{Pthreads} that can help students understand aspects of parallel
+programming that would be difficult to grasp using only \textit{OpenMP}
+examples. His catalog also contained patterns implemented in \textit{OpenMP}.
+
+\todo[inline,color=cyan,author=Pedro]{This last paragraph maybe fits best in
+the end of the introduction}
+
+In this paper we present data that supports the importance of composing a
+curriculum for Parallel and Distributed Computing courses using low-level
+programming interfaces, such as \textit{Pthreads}, as well as high-level
+interfaces, such as \textit{OpenMP}.  Brown \textit{et
+al.}~\cite{Brown:2010:SPC:1971681.1971689}, points that the choice of a
+software tool should be only one of the topics explored during a parallel
+programming course. Topics such as data structures, algorithms, software design
+and parallel hardware platforms should also be explored.
+
+\todo[inline,color=orange,author=Raphael]{Maybe we should make a better
+connection between the end of this section and the next one;}

### 2017-08-07 sections/03-apis.tex
Improve responses section.
The Framework Sections

@@
 
-\todo[inline,color=cyan,author=Pedro]{Provide brief descriptions of the APIs}
+In this section we discuss the main concepts involved in both OpenMP and
+pthreads libraries as they can both be used to implement parallel programs by
+using the fork-join model, on which the sequential region execution is followed
+by parallel regions execution.
+
+\subsection{Pthreads}
+
+The pthreads library specification was design to solve the problem of having
+multiple hardware vendors delivering specific APIs and tools for developing
+multithread programs that were not portable across multiple platforms. The
+thread programming interface has been specified by the IEEE POSIX 1003.1c
+standard and any implementation adhering to this standard is called POSIX
+threads or pthreads.
+
+In C/C++ languages the pthreads library is used to control thread execution
+on GNU/Linux by providing mechanisms to create and synchronize threads. In this
+model, the execution starts in a sequential region and threads needed to be
+explicitly started in order to start a parallel region.
+
+Listing \ref{lst:listing-pthreads} presents a sample of code that uses the
+pthreads library. As explained before, the program starts its execution in a
+single thread and up to line 6 only one thread is executing the main flow of
+instructions.
+
+When the \texttt{pthread\_create()} function is called, a new thread is created
+and then the application enters a parallel region with two threads executing.
+The main mechanism available for synchroninzing thread executions are thread
+barrier. The pthread library implements barriers at the
+\texttt{pthread\_join()} where the thread that reaches that point will wait for
+the completion of other threads to continue the execution of the main flow.
+operation. In line 10 we can see an example of thread synchronization using the
+barrier definition.
+
+\begin{lstlisting}[language=C, basicstyle=\footnotesize, numbers=left, 
+                   frame=single, showspaces=false, showstringspaces=false,
+                   caption={Pthread Sample Program}, captionpos=b,
+                   label=lst:listing-pthreads]
+int main(){
+    pthread_t thread1 ;
+    DATA TYPE *params = 10;
+    int iret1;
+
+    iret1 = pthread_create(&thread1 , NULL, 
+                           task function , 
+                           (void*) params);
+    printf ("Thread 1 return : %d\n", iret1 ) ;
+    pthread_join(thread1 , NULL) ;
+
+    return 0;
+}
+void *task function(void *ptr) { 
+    DATATYPE *value;
+    value = (DATA TYPE *) ptr ;
+    /* Computing */
+}
+\end{lstlisting}
+
+The choice to use the pthreads library, and not OpenMP, can be motivated by the
+fact that pthreads is a common library, available on platforms with GNU/Linux,
+and operates in a lower level of abstraction in comparison with OpenMP threads.
+
+\subsection{OpenMP}
+
+OpenMP is an specification that defines a set of compiler directives, libraries
+and environment variables with the goal to deliver tools that help exploring
+the inherent program parallelism by means of threads. The OpenMP specification
+is kept by a group named OpenMP Architecture Review Board \footnote{} composed
+by several hardware manufacturers as well as parallel software developers and
+users.
+
+The code structure of an OpenMP program is composed by intertwined sequential
+and parallel regions that always start with a sequential one, i.e., the program
+starts with a sequential region - the master thread, then a series of parallel
+ones starts and ends - fork-join model.
+
+The OpenMP directives are implemented in C/C++ and Fortran compilers. At the
+C/C++ compilers, the specification of parallel executions is done by means of
+\textit{pragma} directives. The pragma directives are defined inside the code
+by means of the \texttt{\#pragma} keyword. In case the compiler does not
+recognize the keyword, the compilation procedure is not affected and the
+sequential version of the program is built.
+
+The directives format follow the rule \texttt{\#pragma directive [clauses]}.
+Each line has at least one directive and may contain one or more clauses. The
+main directives available at the specification are: 
+\begin{itemize}
+    \item Work sharing: this set of directives dictate how the work is shared
+    among threads;
+    \item Tasks: define independent code regions that can be executed by any
+    thread available. It also defines the dependencies between tasks;
+    \item Synchronization: used to synchronize the code regions that are being
+    executing within separated threads.
+\end{itemize}
+
+Listing \ref{lst:listing-omp} shows a basic OpenMP example. The example only
+shows the declaration of a parallel region named \texttt{firstprivate} (line
+1), then a region that should only be executed by a single thread (line 2),
+then a for that should be executed in parallel and that the work division will
+bu automatic (line 6). Then, in line 10 we define a critical region which can
+only be executed by a single thread at a time. In the end, at line 12 we define
+a region that should only be executed by the master thread.
+
+\begin{lstlisting}[language=C, basicstyle=\footnotesize, numbers=left, 
+                   frame=single, showspaces=false, showstringspaces=false,
+                   caption={OpenMP Sample Program}, captionpos=b,
+                   label=lst:listing-omp]
+#pragma omp parallel firstprivate (aux dot){ 
+    #pragma omp single
+    printf ("Begin of the parallel region, 
+            number of threads : %d\n", 
+            omp get num threads ()) ;
+    #pragma omp for schedule (auto) 
+    for(i = 0; i < SIZE; i++){
+        aux dot += A[i] * B[i]; 
+    }    
+    #pragma omp critical
+    dot += aux_dot ;
+    #pragma omp master
+    printf("Final result: %d.\n", dot);
+}
+\end{lstlisting}

### 2017-08-07 sections/01-introduction.tex
Add text to responses section, improve figures

@@
 
+\todo[inline,color=cyan,author=Pedro]{Write new introduction}
+
 ...
-
-In this context, the approach of using Compilation Directives to guide the
-compiler in the parallelization process has gained popularity. These directives
-are implemented using the compiler's pre-processing directives and are utilized
-as annotations that provide tips about the sequential code. Among the tools and
-extensions that use compilation directives are the \texttt{OpenMP} API
-\cite{Dagum1998a} \cite{Chapman:2007} used for writing programs for multi-core
-architectures, and the \texttt{OpenACC} programming
-standard~\cite{openacc:api}, used for writing programs for heterogeneous
-CPU/GPU architectures. The \texttt{OpenMP 4.0} specification supports
-offloading capabilities \cite{openmp:api:2013} such as \texttt{OpenACC} and
-\texttt{AMD HSA}~\cite{amd:hsa:site}. These tools aim to make easier the task
-of designing and implementing efficient and provably correct parallel
-algorithms. However, the abstraction layers built by the extensions also have
-the potential to conceal architectural details and to generate new parallel
-programming challenges.
-
-When comparing \texttt{OpenMP} to other available tools for developing
-multi-threaded applications such as the \texttt{pthreads} interface, the API
-can be considered almost costless. However, when writing more complex
-applications this may not be true. The design and implementation of such
-complex applications require more knowledge about platforms and execution
-models than what can be shown by tutorials, that present high level and trivial
-examples. These tutorials try to convince the user that programming with
-directives is simple and easy to do, which is not always true. For instance, to
-achieve good performance in some cases it is necessary to describe many
-restrictions on annotations~\cite{OpenMPTasks2009}~\cite{mattson2003good}.
-
-Krawezik and Cappello~\cite{CPE:CPE905} evidenced this difficulty by comparing
-MPI and three OpenMP versions of benchmarks. The most popular use of
-\texttt{OpenMP} directives is at the loop level. The programmer discovers
-potentially parallel loops and annotates the code with directives starting and
-closing parallel regions. Good results were obtained only when SPMD was
-implemented on OpenMP, and achieving good performance required a significant
-programming effort~\cite{CPE:CPE905}.
 
@@
 for the \emph{Introduction to Parallel and Distributed Computing} course. Among
-other exercises, the students were asked to find examples of \texttt{OpenMP}
-code in tutorials and official manuals that failed under special conditions.
-These conditions could be created by, for example, placing barriers or forcing
-racing conditions by exposing inaccurate thread management.
+other exercises, the students were asked to ...
 
-The rest of the paper is organized as follows. The following section discusses
-related work...
+\todo[inline,color=cyan,author=Pedro]{Describe paper structure}

### 2017-08-07 sections/02-background.tex

@@
 
-The approaches of the last paragraph introduce tools for parallel programming
-before the related concepts.  Instead, some learning approaches
-consider learning software tools as just one of the objectives of the course,
-and argue that their introduction should come after the introduction of
-concepts of Parallel and Distributed Computing.
+The approaches mentioned so far introduce tools for parallel programming before
+the related concepts.  Other approaches instead consider learning software
+tools as just one of the objectives of the course, and argue that their
+introduction should come after the introduction of concepts of Parallel and
+Distributed Computing.
 

### 2017-08-07 sections/02-background.tex
Re-organizing sections

@@
 
-%\todo[inline,color=cyan,author=Pedro]{Cite our previous paper, making it clear that we are the authors}
-%\todo[inline,color=cyan,author=Pedro]{Describe related work}
-%\todo[inline,color=cyan,author=Pedro]{Describe related work since last paper}
+In this section we discuss the main concepts involved in both OpenMP and
+pthreads libraries as they can both be used to implement parallel programs by
+using the fork-join model, on which the sequential region execution is followed
+by parallel regions execution.
 
-The selection of teaching tools for Parallel and Distributed classes impacts
-the development of students' abilities to solve algorithmic problems
-efficiently. The importance of selecting good tools for teaching Parallel and
-Distributed Computing has been increasingly important, especially since the
-topic became a core component of the ACM undergraduate computer science
-curricula in 2013~\cite{acmcurricula}.
+\subsection{Pthreads}
 
-Different approaches have been used to study the selection of the best level of
-abstraction for teaching Parallel and Distributed programming courses.
-Falcao~\cite{6565518} argues that introducing a high-level parallel programming
-interface such as \textit{OpenMP} in the beginning of an undergraduate computer
-science course would cause a small overhead and benefit the rest of the course.
-Foley \textit{et al.}~\cite{FOLEY2017138} support the introduction of practical
-learning experiences along an undergraduate course, of increasing depth as the
-course progresses. They introduce OnRamp, a web application to help students
-learn to write parallel code starting with practical and moving to conceptual
-experiences.  Pllana \textit{et al.}~\cite{Pllana:2009} adopt a Model Driven
-Development approach to develop parallel solutions using Parallel Building
-Blocks. They introduce a programming environment that transforms these
-high-level models in source code. The tool's goal is to hide from the students
-the complexity of writing parallel code.
+The pthreads library specification was design to solve the problem of having
+multiple hardware vendors delivering specific APIs and tools for developing
+multithread programs that were not portable across multiple platforms. The
+thread programming interface has been specified by the IEEE POSIX 1003.1c
+standard and any implementation adhering to this standard is called POSIX
+threads or pthreads.
 
-The approaches mentioned so far introduce tools for parallel programming before
-the related concepts.  Other approaches instead consider learning software
-tools as just one of the objectives of the course, and argue that their
-introduction should come after the introduction of concepts of Parallel and
-Distributed Computing.
+In C/C++ languages the pthreads library is used to control thread execution
+on GNU/Linux by providing mechanisms to create and synchronize threads. In this
+model, the execution starts in a sequential region and threads needed to be
+explicitly started in order to start a parallel region.
 
-In previous work~\cite{goncalves:OpenMPNotEasy} we show that teaching the
-concepts of parallel programming first instead of high-level parallel
-programming APIs is important, especially because the available resources and
-tutorials for these APIs may contain difficult to find errors.  We listed
-several sources of tutorials and hands-on materials obtained in the internet in
-which undergraduate students were able to point parallel correctness
-errors~\cite{SuB:2005:CMO:1892830.1892863}.
+Listing \ref{lst:listing-pthreads} presents a sample of code that uses the
+pthreads library. As explained before, the program starts its execution in a
+single thread and up to line 6 only one thread is executing the main flow of
+instructions.
 
-Sü\ss{} and Leopold~\cite{Leopold:userOpenMP} discuss the inability to express
-all parallel solutions in tools such as \textit{OpenMP}. They used a sorting
-algorithm to show that it is easy to find patterns that are difficult to
-implement in \textit{OpenMP}, but can be easily implemented in lower-level
-programming interfaces, such as POSIX Threads.
+When the \texttt{pthread\_create()} function is called, a new thread is created
+and then the application enters a parallel region with two threads executing.
+The main mechanism available for synchroninzing thread executions are thread
+barrier. The pthread library implements barriers at the
+\texttt{pthread\_join()} where the thread that reaches that point will wait for
+the completion of other threads to continue the execution of the main flow.
+operation. In line 10 we can see an example of thread synchronization using the
+barrier definition.
 
-Adams~\cite{ADAMS201731} presented a set of patterns implemented using
-\textit{Pthreads} that can help students understand aspects of parallel
-programming that would be difficult to grasp using only \textit{OpenMP}
-examples. His catalog also contained patterns implemented in \textit{OpenMP}.
+\begin{lstlisting}[language=C, basicstyle=\footnotesize, numbers=left, 
+                   frame=single, showspaces=false, showstringspaces=false,
+                   caption={Pthread Sample Program}, captionpos=b,
+                   label=lst:listing-pthreads]
+int main(){
+    pthread_t thread1 ;
+    DATA TYPE *params = 10;
+    int iret1;
 
-\todo[inline,color=cyan,author=Pedro]{This last paragraph maybe fits best in
-the end of the introduction}
+    iret1 = pthread_create(&thread1 , NULL, 
+                           task function , 
+                           (void*) params);
+    printf ("Thread 1 return : %d\n", iret1 ) ;
+    pthread_join(thread1 , NULL) ;
 
-In this paper we present data that supports the importance of composing a
-curriculum for Parallel and Distributed Computing courses using low-level
-programming interfaces, such as \textit{Pthreads}, as well as high-level
-interfaces, such as \textit{OpenMP}.  Brown \textit{et
-al.}~\cite{Brown:2010:SPC:1971681.1971689}, points that the choice of a
-software tool should be only one of the topics explored during a parallel
-programming course. Topics such as data structures, algorithms, software design
-and parallel hardware platforms should also be explored.
+    return 0;
+}
+void *task function(void *ptr) { 
+    DATATYPE *value;
+    value = (DATA TYPE *) ptr ;
+    /* Computing */
+}
+\end{lstlisting}
 
-\todo[inline,color=orange,author=Raphael]{Maybe we should make a better
-connection between the end of this section and the next one;}
+The choice to use the pthreads library, and not OpenMP, can be motivated by the
+fact that pthreads is a common library, available on platforms with GNU/Linux,
+and operates in a lower level of abstraction in comparison with OpenMP threads.
+
+\subsection{OpenMP}
+
+OpenMP is an specification that defines a set of compiler directives, libraries
+and environment variables with the goal to deliver tools that help exploring
+the inherent program parallelism by means of threads. The OpenMP specification
+is kept by a group named OpenMP Architecture Review Board \footnote{} composed
+by several hardware manufacturers as well as parallel software developers and
+users.
+
+The code structure of an OpenMP program is composed by intertwined sequential
+and parallel regions that always start with a sequential one, i.e., the program
+starts with a sequential region - the master thread, then a series of parallel
+ones starts and ends - fork-join model.
+
+The OpenMP directives are implemented in C/C++ and Fortran compilers. At the
+C/C++ compilers, the specification of parallel executions is done by means of
+\textit{pragma} directives. The pragma directives are defined inside the code
+by means of the \texttt{\#pragma} keyword. In case the compiler does not
+recognize the keyword, the compilation procedure is not affected and the
+sequential version of the program is built.
+
+The directives format follow the rule \texttt{\#pragma directive [clauses]}.
+Each line has at least one directive and may contain one or more clauses. The
+main directives available at the specification are: 
+\begin{itemize}
+    \item Work sharing: this set of directives dictate how the work is shared
+    among threads;
+    \item Tasks: define independent code regions that can be executed by any
+    thread available. It also defines the dependencies between tasks;
+    \item Synchronization: used to synchronize the code regions that are being
+    executing within separated threads.
+\end{itemize}
+
+Listing \ref{lst:listing-omp} shows a basic OpenMP example. The example only
+shows the declaration of a parallel region named \texttt{firstprivate} (line
+1), then a region that should only be executed by a single thread (line 2),
+then a for that should be executed in parallel and that the work division will
+bu automatic (line 6). Then, in line 10 we define a critical region which can
+only be executed by a single thread at a time. In the end, at line 12 we define
+a region that should only be executed by the master thread.
+
+\begin{lstlisting}[language=C, basicstyle=\footnotesize, numbers=left, 
+                   frame=single, showspaces=false, showstringspaces=false,
+                   caption={OpenMP Sample Program}, captionpos=b,
+                   label=lst:listing-omp]
+#pragma omp parallel firstprivate (aux dot){ 
+    #pragma omp single
+    printf ("Begin of the parallel region, 
+            number of threads : %d\n", 
+            omp get num threads ()) ;
+    #pragma omp for schedule (auto) 
+    for(i = 0; i < SIZE; i++){
+        aux dot += A[i] * B[i]; 
+    }    
+    #pragma omp critical
+    dot += aux_dot ;
+    #pragma omp master
+    printf("Final result: %d.\n", dot);
+}
+\end{lstlisting}

### 2017-08-07 sections/03-relatedwork.tex
Re-organizing sections

@@
 
-TO-DO ...
+%\todo[inline,color=cyan,author=Pedro]{Cite our previous paper, making it clear that we are the authors}
+%\todo[inline,color=cyan,author=Pedro]{Describe related work}
+%\todo[inline,color=cyan,author=Pedro]{Describe related work since last paper}
+
+The selection of teaching tools for Parallel and Distributed classes impacts
+the development of students' abilities to solve algorithmic problems
+efficiently. The importance of selecting good tools for teaching Parallel and
+Distributed Computing has been increasingly important, especially since the
+topic became a core component of the ACM undergraduate computer science
+curricula in 2013~\cite{acmcurricula}.
+
+Different approaches have been used to study the selection of the best level of
+abstraction for teaching Parallel and Distributed programming courses.
+Falcao~\cite{6565518} argues that introducing a high-level parallel programming
+interface such as \textit{OpenMP} in the beginning of an undergraduate computer
+science course would cause a small overhead and benefit the rest of the course.
+Foley \textit{et al.}~\cite{FOLEY2017138} support the introduction of practical
+learning experiences along an undergraduate course, of increasing depth as the
+course progresses. They introduce OnRamp, a web application to help students
+learn to write parallel code starting with practical and moving to conceptual
+experiences.  Pllana \textit{et al.}~\cite{Pllana:2009} adopt a Model Driven
+Development approach to develop parallel solutions using Parallel Building
+Blocks. They introduce a programming environment that transforms these
+high-level models in source code. The tool's goal is to hide from the students
+the complexity of writing parallel code.
+
+The approaches mentioned so far introduce tools for parallel programming before
+the related concepts.  Other approaches instead consider learning software
+tools as just one of the objectives of the course, and argue that their
+introduction should come after the introduction of concepts of Parallel and
+Distributed Computing.
+
+In previous work~\cite{goncalves:OpenMPNotEasy} we show that teaching the
+concepts of parallel programming first instead of high-level parallel
+programming APIs is important, especially because the available resources and
+tutorials for these APIs may contain difficult to find errors.  We listed
+several sources of tutorials and hands-on materials obtained in the internet in
+which undergraduate students were able to point parallel correctness
+errors~\cite{SuB:2005:CMO:1892830.1892863}.
+
+Sü\ss{} and Leopold~\cite{Leopold:userOpenMP} discuss the inability to express
+all parallel solutions in tools such as \textit{OpenMP}. They used a sorting
+algorithm to show that it is easy to find patterns that are difficult to
+implement in \textit{OpenMP}, but can be easily implemented in lower-level
+programming interfaces, such as POSIX Threads.
+
+Adams~\cite{ADAMS201731} presented a set of patterns implemented using
+\textit{Pthreads} that can help students understand aspects of parallel
+programming that would be difficult to grasp using only \textit{OpenMP}
+examples. His catalog also contained patterns implemented in \textit{OpenMP}.
+
+\todo[inline,color=cyan,author=Pedro]{This last paragraph maybe fits best in
+the end of the introduction}
+
+In this paper we present data that supports the importance of composing a
+curriculum for Parallel and Distributed Computing courses using low-level
+programming interfaces, such as \textit{Pthreads}, as well as high-level
+interfaces, such as \textit{OpenMP}.  Brown \textit{et
+al.}~\cite{Brown:2010:SPC:1971681.1971689}, points that the choice of a
+software tool should be only one of the topics explored during a parallel
+programming course. Topics such as data structures, algorithms, software design
+and parallel hardware platforms should also be explored.
+
+\todo[inline,color=orange,author=Raphael]{Maybe we should make a better
+connection between the end of this section and the next one;}

### 2017-08-08 sections/05-responses.tex
Minor edits to responses

@@
 students we elaborated a five-point scale from \textit{Do not Know} $(1)$ to
-\textit{Expert or Proficient} $(5)$.  The students were asked to express their
+\textit{Expert or Proficient} $(5)$.  We asked the students to express their
 proficiency level on this scale, and their responses are shown in Figure
@@
 
-The students were asked to evaluate their proficiency in \textit{OpenMP} and
+We asked the students to evaluate their proficiency in \textit{OpenMP} and
 \textit{Pthreads}, \textit{before} completing the assignment, with two sets of
@@
 
-The students were asked to answer 3 sets of questions regarding their
+We asked the students to answer 3 sets of questions regarding their
 experience with \textit{OpenMP} and \textit{Pthreads}, \textit{after}

### 2017-08-08 sections/02-apis.tex
Edit section on APIs

\section{Pthreads \& OpenMP}
\label{sec:apis}

In this section we briefly describe \textit{OpenMP} and \textit{Pthreads}.

\subsection{Pthreads}

The \textit{Pthreads} library specification was designed to provide a standard
and portable API for developing multi-threaded programs for multiple vendor
platforms.  The programming interface was specified by the \textit{IEEE POSIX
1003.1c} standard. Any implementation adhering to this standard is said to be
POSIX threads, or \textit{Pthreads}.

The C/C++ \textit{Pthreads} library controls thread execution in GNU/Linux by
creating and synchronizing threads. In this model, the execution begins in a
sequential region and threads must be explicitly started to begin
a parallel region.
Listing \ref{lst:listing-pthreads} presents a sample code using the
\textit{Pthreads} library. The program begins execution in a
single thread and until line 6 only one thread is running.

When \texttt{pthread\_create} is called a thread is created and the application
enters a parallel region with two simultaneous threads.  The main mechanism
available for synchronizing thread executions is the barrier. The
\textit{Pthreads} library implements barriers as calls to
\texttt{pthread\_join}. A thread that reaches a barrier will wait for other
threads. Line 10 shows an example of thread synchronization using a barrier.

\begin{lstlisting}[language=C, basicstyle=\ttfamily\scriptsize, numbers=left,
                   frame=no, showspaces=false, showstringspaces=false,
                   caption={\textit{Pthreads} Sample Program}, captionpos=b,
                   numberstyle=\tiny,
                   xleftmargin=0.5cm,
                   label=lst:listing-pthreads, keywords={%
                       DATATYPE, pthread_t, pthread_create,
                       pthread_join, task_function, NULL, int, main,
                       void, printf, return%
                       },
                   otherkeywords={::, \#pragma, \#include, <<<,>>>, \&, \*, +, -, /, [, ], >, <}
                   ]
int main(){
    pthread_t thread_id;
    DATATYPE *parameters = 10;
    int return_code;

    return_code = pthread_create(&thread_id, NULL,
                                 task_function,
                                 (void *) parameters);

    printf("Return Code: %d\n", return_code );
    pthread_join(thread_id , NULL);

    return 0;
}

void *task_function(void *data){
    DATATYPE *value;
    value = (DATATYPE *) data;
}
\end{lstlisting}

\subsection{OpenMP}

\textit{OpenMP} is an specification that defines a set of compiler directives,
libraries and environment variables that help exploring program parallelism.
The \textit{OpenMP} specification is kept by the \textit{OpenMP Architecture
Review Board}, composed by hardware manufacturers, and parallel software
developers and users.

\textit{OpenMP} implements the \textit{fork-join} model. An \textit{OpenMP}
program contains weaved sequential and parallel regions and always starts with
a sequential region, or master thread.  \textit{OpenMP} directives are
implemented in C/C++ and Fortran compilers. In C/C++ compilers the
specification of parallel executions is done using \textit{pragmas}, defined in
by using the \texttt{\#pragma} keyword.
The compilation process is not affected by the removal of the pragma,
which will result in the generation of a sequential version of the program.

\textit{OpenMP} directives follow the format \texttt{\#pragma directive
[clauses]}.  Each line contains at least one directive and may contain one or
more clauses. The main directives defined by the specification are related
to work sharing, task definition and dependencies, and synchronization.

Listing \ref{lst:listing-omp} presents an \textit{OpenMP} example. The
declaration of a parallel region with the \texttt{firstprivate} pragma on line
1 indicates the variable \texttt{aux\_dot} should be private to every thread
and initialized with the enclosing scope's value.  Line 3 will be executed by
only one thread, because of the \texttt{single} pragma.
The \texttt{for} in line 7 will be executed in parallel, with automatic work
division.  The pragma at line 12 defines a critical region which can only be
executed by a single thread at a time. The pragma at line 15 defines a
region that should only be executed by the master thread.

\begin{lstlisting}[language=C, basicstyle=\ttfamily\scriptsize, numbers=left,
                   frame=no, showspaces=false, showstringspaces=false,
                   caption={\textit{OpenMP} Sample Program}, captionpos=b,
                   numberstyle=\tiny,
                   xleftmargin=0.6cm,
                   label=lst:listing-omp, keywords={%
                       \#pragma,
                       omp, parallel, firstprivate,
                       single, omp_get_num_threads,
                       for, schedule, auto,
                       critical, master,
                       NULL, int, main,
                       void, printf, return%
                   },
                   otherkeywords={::, \#pragma, \#include, <<<,>>>, \&, \*, +, -, /, [, ], >, <}
                       ]
#pragma omp parallel firstprivate(aux_dot){
    #pragma omp single
    printf("Start of parallel region,
            number of threads: %d\n",
            omp_get_num_threads ());

    #pragma omp for schedule(auto)
    for(i = 0; i < SIZE; i++){
        aux_dot += A[i] * B[i];
    }

    #pragma omp critical
    dot += aux_dot;

    #pragma omp master
    printf("Result: %d.\n", dot);
}
\end{lstlisting}

### 2017-08-08 sections/05-responses.tex
Edit section on APIs

@@
 regarding \textit{OpenMP} did not reflect in the actual implementations and
-submissions. We believe that \textit{OpenMP} is still not as easy as it appears
+submissions. We assert that \textit{OpenMP} is still not as easy as it appears
 to be, and that the students' responses strengthen the argument for teaching

### 2017-08-08 sections/01-introduction.tex
Moved a few sections from the background the the introduction

@@
 
-...
+The selection of teaching tools for Parallel and Distributed classes impacts
+the development of students' abilities to solve algorithmic problems
+efficiently. The importance of selecting good tools for teaching Parallel and
+Distributed Computing has been increasingly important, especially since the
+topic became a core component of the ACM undergraduate computer science
+curricula in 2013~\cite{acmcurricula}.
 
@@
 
-\todo[inline,color=cyan,author=Pedro]{Describe paper structure}
+In this paper we present data that supports the importance of composing a
+curriculum for Parallel and Distributed Computing courses using low-level
+programming interfaces, such as \textit{Pthreads}, as well as high-level
+interfaces, such as \textit{OpenMP}.  Brown \textit{et
+al.}~\cite{Brown:2010:SPC:1971681.1971689}, points that the choice of a
+software tool should be only one of the topics explored during a parallel
+programming course. Topics such as data structures, algorithms, software design
+and parallel hardware platforms should also be explored.

### 2017-08-08 sections/03-background.tex
Moved a few sections from the background the the introduction

--- 2017-08-07 sections/03-relatedwork.tex
@@
-\section{Related Work}
-\label{sec:relatedwork}
-
-%\todo[inline,color=cyan,author=Pedro]{Cite our previous paper, making it clear that we are the authors}
-%\todo[inline,color=cyan,author=Pedro]{Describe related work}
-%\todo[inline,color=cyan,author=Pedro]{Describe related work since last paper}
-
-The selection of teaching tools for Parallel and Distributed classes impacts
-the development of students' abilities to solve algorithmic problems
-efficiently. The importance of selecting good tools for teaching Parallel and
-Distributed Computing has been increasingly important, especially since the
-topic became a core component of the ACM undergraduate computer science
-curricula in 2013~\cite{acmcurricula}.
+\section{Background}
+\label{sec:background}
 
@@
 
-\todo[inline,color=cyan,author=Pedro]{This last paragraph maybe fits best in
-the end of the introduction}
-
-In this paper we present data that supports the importance of composing a
-curriculum for Parallel and Distributed Computing courses using low-level
-programming interfaces, such as \textit{Pthreads}, as well as high-level
-interfaces, such as \textit{OpenMP}.  Brown \textit{et
-al.}~\cite{Brown:2010:SPC:1971681.1971689}, points that the choice of a
-software tool should be only one of the topics explored during a parallel
-programming course. Topics such as data structures, algorithms, software design
-and parallel hardware platforms should also be explored.
-
 \todo[inline,color=orange,author=Raphael]{Maybe we should make a better

### 2017-08-08 sections/01-introduction.tex
Add text to research design. Minor edits.

@@
 
-The selection of teaching tools for Parallel and Distributed classes impacts
+The selection of teaching tools for Parallel and Distributed courses impacts
 the development of students' abilities to solve algorithmic problems

### 2017-08-08 sections/04-research-design.tex
Add text to research design. Minor edits.

@@
 
-\todo[inline,color=cyan,author=Pedro]{Describe research design}
+This section describes our research design. We devised a questionnaire
+regarding student experience with solving the same problem using
+\textit{Pthreads} and \textit{OpenMP} during the Concurrent, Parallel and
+Distributed Computing Course at the Computer Science program of the University
+of São Paulo. We answered 3 research questions using analyses of the students'
+responses.
 
@@
 
-\todo[inline,color=cyan,author=Pedro]{Describe the course}
+The Computer Science program of the University of São Paulo offers Concurrent,
+Parallel and Distributed Computing courses for graduate and undergraduate
+students in a single class. The course aims to promote meaningful exchanges of
+knowledge and experience between undergraduate and graduate students.
+
+The course presents students with a series of software engineering challenges
+in the form of assignments. The final is composed of an essay and a
+presentation on a relevant topic related to Concurrent, Parallel and
+Distributed Computing.
+
+We devised an assignment to measure the students' difficulties to
+learn and use \textit{Pthreads} and \textit{OpenMP} to solve
+a simple problem. Students had to implement parallel versions
+of a sequential code using \textit{Pthreads} and \textit{OpenMP}.
+We provided a sequential implementation for the calculation of
+the Mandelbrot Set. The sequential code presented an embarrassingly
+parallel problem with two nested loops, and a third loop with
+dependent iterations that could not be parallelized easily.
+
+The assignment required that the students measure the performance of the
+sequential code and the performance improvements they achieved using
+\textit{Pthreads} and \textit{OpenMP}.  After completing the assignment the
+students were asked to answer a questionnaire.
 
@@
 
-\todo[inline,color=cyan,author=Pedro]{Describe the questionnaire}
-\todo[inline,color=cyan,author=Pedro]{Add reference for questionnaires from Herez}
+The questionnaire had two parts. The first part asked questions
+about the students' previous experiences with common programming languages and
+parallel and distributed programming concepts and tools. We were also
+interested in students' perception of the impact of our classes on
+\textit{Pthreads} and \textit{OpenMP} on their learning difficulty.
+
+The questions of the second part were related to the students'
+experience with \textit{Pthreads} and \textit{OpenMP} \textit{after}
+the completion of the assignment.
+
+We obtained responses, and consent to use them in our research, from 38 of the
+54 students enrolled in the Concurrent, Parallel and Distributed Computing
+course. Section \ref{sec:background} presents a detailed description of
+each set of questions our analysis of the students' responses.
 
@@
 
-\todo[inline,color=cyan,author=Pedro]{Improve RQ discussion and descriptions}
+In previous work~\cite{goncalves:OpenMPNotEasy} we showed that
+teaching \textit{OpenMP} requires teaching important parallel
+and distributed computing concepts first, since it is easy
+to find errors on commonly used \textit{OpenMP} learning material.
 
-According to the students' perception, which API:
+In this paper we argue that students are capable of learning and using
+\textit{Pthreads} to obtain good performance improvements, despite their
+perception that \textit{OpenMP} is much easier to learn and use.
+The research questions that led the development of the questionnaire
+and our analysis of the students' responses were the following:
 
 \begin{description}
-    \item[RQ1:] \textit{Was easier to learn and use?}
-    \item[RQ2:] \textit{Improved performance the most?}
-\end{description}
-
-After evaluation of the students' performance improvements
-in the assignment:
-
-\begin{description}
+    \item[RQ1:] \textit{According to the students' perception, which API was
+        easier to learn and use?}
+    \item[RQ2:] \textit{With which API did the students improve performance the
+        most?}
     \item[RQ3:] \textit{Did the students' performance improvements match their
@@
 \end{description}
+
+To answer the first two questions we used students' responses to
+our questionnaire. To answer the third question we analysed the
+students' submissions and parallel implementations. The
+students' responses and our analysis are presented next.

### 2017-08-08 sections/00-abstract.tex
Add text to responses section, improve figures

@@
 \begin{abstract}
-This paper provides a sample of a \LaTeX\ document which conforms,
-somewhat loosely, to the formatting guidelines for
-ACM SIG Proceedings. 
+\todo[inline,color=cyan,author=Pedro]{Summarize results}
 \end{abstract}
@@
 %
+
 \begin{CCSXML}
@@
  <concept>
-  <concept_id>10010520.10010553.10010562</concept_id>
-  <concept_desc>Computer systems organization~Embedded systems</concept_desc>
+  <concept_id>10010147.10010169</concept_id>
+  <concept_desc>Computing methodologies~Parallel computing methodologies</concept_desc>
   <concept_significance>500</concept_significance>
@@
  <concept>
-  <concept_id>10010520.10010575.10010755</concept_id>
-  <concept_desc>Computer systems organization~Redundancy</concept_desc>
-  <concept_significance>300</concept_significance>
- </concept>
- <concept>
-  <concept_id>10010520.10010553.10010554</concept_id>
-  <concept_desc>Computer systems organization~Robotics</concept_desc>
-  <concept_significance>100</concept_significance>
- </concept>
- <concept>
-  <concept_id>10003033.10003083.10003095</concept_id>
-  <concept_desc>Networks~Network reliability</concept_desc>
-  <concept_significance>100</concept_significance>
+  <concept_id>10010147.10010919</concept_id>
+  <concept_desc>Computing methodologies~Distributed computing methodologies</concept_desc>
+  <concept_significance>500</concept_significance>
  </concept>
@@
 
-\ccsdesc[500]{Computer systems organization~Embedded systems}
-\ccsdesc[300]{Computer systems organization~Redundancy}
-\ccsdesc{Computer systems organization~Robotics}
-\ccsdesc[100]{Networks~Network reliability}
+\ccsdesc[500]{Computing methodologies~Parallel computing methodologies}
+\ccsdesc[500]{Computing methodologies~Distributed computing methodologies}
 
-\keywords{ACM proceedings, \LaTeX, text tagging}
+\keywords{Parallel and Distributed Computing, OpenMP, Pthreads}

### 2017-08-08 sections/00-abstract.tex
Add text for the abstract, introduction and conclusion.

@@
 \begin{abstract}
-\todo[inline,color=cyan,author=Pedro]{Summarize results}
+
+    Selecting the most effective tools and languages for teaching parallel and
+    distributed computing presents a challenge for teachers and students. Often
+    a high-level solution seems easier, but it comes with the cost of hiding
+    underlying concepts and problems. We present a comparison between the
+    \textit{Pthreads} and \textit{OpenMP} APIs considering the experiences and
+    performance results of undergraduate and graduate Computer Science students
+    from the University of São Paulo.
+    We show that students were able to achieve better performance results using
+    \textit{Pthreads} than \textit{OpenMP}, despite their perception that
+    \textit{Pthreads} is harder to learn and use.
+
 \end{abstract}

### 2017-08-08 sections/01-introduction.tex
Add text for the abstract, introduction and conclusion.

--- 2017-08-07 sections/01-introduction.tex
@@
 
-\todo[inline,color=cyan,author=Pedro]{Write new introduction}
+Multi-core architectures became essential and ubiquitous for High-Performance
+Computing in the last decade, their relevance increased by recent slowdowns on
+Moore's law and the end of Dennard scaling.  Selecting the appropriate tool for
+programming multi-core architectures is also essential, since it is necessary
+to program a multi-core architecture effectively to leverage its performance.
 
-...
+Programming multi-core architectures demand different skills and tools from
+what is needed for traditional sequential programming. Selecting an effective
+language, framework or API for parallel and distributed programming is a
+challenge for teachers and students, especially due to the quantity and
+heterogeneity of available resources.
 
-In this work we have gathered data from assignments made by graduate students
-for the \emph{Introduction to Parallel and Distributed Computing} course. Among
-other exercises, the students were asked to ...
+Tools with a higher level of abstraction may provide a smoother transition from
+sequential to parallel and distributed programming, making it easier for
+students to develop working code. The cost of high-level abstractions often is
+the hiding of underlying concepts, problems and solutions, as well as code
+performance.
 
-\todo[inline,color=cyan,author=Pedro]{Describe paper structure}
+The contribution of this paper is a comparison between \textit{Pthreads} and
+\textit{OpenMP}, two commonly used APIs for parallel and distributed
+programming.  The comparison considers the perceptions and performance results
+from undergraduate and graduate students enrolled in the course of Concurrent,
+Parallel and Distributed Computing of the University of São Paulo, Brazil.  We
+show that students were able to achieve better performance results using
+\textit{Pthreads} than \textit{OpenMP}, despite their perception that
+\textit{Pthreads} is harder to learn and use.
+
+The paper is organized as follows.
+Section \ref{sec:apis} describes \textit{OpenMP} and \textit{Pthreads}.
+Section \ref{sec:background} discusses related work.
+Section \ref{sec:researchdesign} discusses the course of
+Concurrent, Parallel and Distribute Computing of the University
+of São Paulo, and presents the questionnaire and our research
+questions.
+Section \ref{sec:responses} presents, discusses and analyses
+the students' responses to the questionnaire.
+Section \ref{sec:conclusion} summarizes the results and
+presents future work.

### 2017-08-08 sections/02-apis.tex
Add text to research design. Minor edits.
Add text for the abstract, introduction and conclusion.

@@
 
-In this section we briefly describe \textit{OpenMP} and \textit{Pthreads}.
+The selection of teaching tools for Parallel and Distributed courses impacts
+the development of students' abilities to solve algorithmic problems
+efficiently. The importance of selecting good tools for teaching Parallel and
+Distributed Computing has been increasingly important, especially since the
+topic became a core component of the ACM undergraduate computer science
+curricula in 2013~\cite{acmcurricula}.  In this section we briefly describe
+\textit{OpenMP} and \textit{Pthreads}, two commonly used tools for parallel and
+distributed programming.
 
@@
 \texttt{pthread\_join}. A thread that reaches a barrier will wait for other
-threads. Line 10 shows an example of thread synchronization using a barrier.
+threads. Line 11 shows an example of thread synchronization using a barrier.
 

### 2017-08-08 sections/03-background.tex
Add text for the abstract, introduction and conclusion.

@@
 
-Different approaches have been used to study the selection of the best level of
+Different approaches have been proposed to study the selection of the best level of
 abstraction for teaching Parallel and Distributed programming courses.
@@
 
-\todo[inline,color=orange,author=Raphael]{Maybe we should make a better
-connection between the end of this section and the next one;}
+In this paper we present data that supports the importance of composing a
+curriculum for Parallel and Distributed Computing courses using low-level
+programming interfaces, such as \textit{Pthreads}, as well as high-level
+interfaces, such as \textit{OpenMP}.  Brown \textit{et
+al.}~\cite{Brown:2010:SPC:1971681.1971689}, points that the choice of a
+software tool should be only one of the topics explored during a parallel
+programming course. Topics such as data structures, algorithms, software design
+and parallel hardware platforms should also be explored.

### 2017-08-08 sections/05-responses.tex
Add text for the abstract, introduction and conclusion.

@@
 
+\begin{figure}[htpb]
+    \centering
+    \includegraphics[width=0.85\columnwidth]{classes_questions}
+    \caption{Students' responses to questions in Table \ref{tab:classes}}
+    \label{fig:classes}
+\end{figure}
+
 The second set of questions regarding student proficiency with the APIs
@@
 students reported that the classes on each API helped their learning.
-
-\begin{figure}[htpb]
-    \centering
-    \includegraphics[width=0.85\columnwidth]{classes_questions}
-    \caption{Student API knowledge \textit{before} the assignment and their relation to classes}
-    \label{fig:classes}
-\end{figure}
 
@@
 
+\begin{figure}[htpb]
+    \centering
+    \includegraphics[width=0.85\columnwidth]{likert_questions}
+    \caption{Students' responses to the Likert Scale questions in Table \ref{tab:likert}}
+    \label{fig:likert}
+\end{figure}
+
 Figure \ref{fig:likert} shows a visualization of the students' responses.  The
@@
 represented by different colors.
-
-\begin{figure}[htpb]
-    \centering
-    \includegraphics[width=0.85\columnwidth]{likert_questions}
-    \caption{Student responses to the Likert Scale questions in Table \ref{tab:likert}}
-    \label{fig:likert}
-\end{figure}
 
@@
 
+Figure \ref{fig:comparisons} shows a visualization of the students' responses.
+The height of each stacked sub-column represents the percentage of respondents
+that selected each API, which are represented by different colors.
+
+The percentage of \textit{OpenMP} selections is significantly larger for all
+questions, except for questions $(7)$ and $(13)$. The responses for questions
+$(7)$ -- $(12)$ strengthen the previous conclusion that the perception of the
+students was that it was easier to use \textit{OpenMP} than \textit{Pthreads}
+in the context of the assignment. The performance results that we obtained from
+analysing the students' submissions challenge the perception of the students.
+The responses to question $(13)$, extracted by our analysis, show that
+\textbf{the students achieved better performance when using \textit{Pthreads}
+than \textit{OpenMP}}.
+
 \begin{table}[htpb]
@@
 
-Figure \ref{fig:comparisons} shows a visualization of the students' responses.
-The height of each stacked sub-column represents the percentage of respondents
-that selected each API, which are represented by different colors.
-
-The percentage of \textit{OpenMP} selections is significantly larger for all
-questions, except for questions $(7)$ and $(13)$. The responses for questions
-$(7)$ -- $(12)$ strengthen the previous conclusion that the perception of the
-students was that it was easier to use \textit{OpenMP} than \textit{Pthreads}
-in the context of the assignment. The performance results that we obtained from
-analysing the students' submissions challenge the perception of the students.
-The responses to question $(13)$, extracted by our analysis, show that
-\textbf{the students achieved better performance when using \textit{Pthreads}
-than \textit{OpenMP}}.
-
 \begin{figure}[htpb]
@@
     \includegraphics[width=0.85\columnwidth]{comparisons}
-    \caption{Student responses to the questions in Table \ref{tab:comparisons}}
+    \caption{Students' responses to the questions in Table \ref{tab:comparisons}}
     \label{fig:comparisons}
 \end{figure}
-
-\todo[inline,color=cyan,author=Pedro]{Improve these last paragraphs}
-
-The responses presented in this section enable the conclusion that the students
-perceived less difficulty using and learning \textit{OpenMP} in the context
-of the assignment. They also reported the perception that using \textit{OpenMP}
-made it easier to improve the performance of the sequential code we provided.
-Our analysis of the submissions revealed that the majority of the students
-($52.2\%$) achieved better performance in the assignment with either
-\textit{OpenMP} ($17.4\%$) or \textit{Pthreads} ($34.8\%$). Contradicting the
-perception of the students, the implementation using \textit{Pthreads} achieved
-better performance in twice the submissions.
-
-This shows that the perceptions of usability and performance improvement
-regarding \textit{OpenMP} did not reflect in the actual implementations and
-submissions. We assert that \textit{OpenMP} is still not as easy as it appears
-to be, and that the students' responses strengthen the argument for teaching
-lower-level technologies for parallel and distributed computing.

### 2017-08-08 sections/07-conclusion.tex
Add text for the abstract, introduction and conclusion.

\section{Conclusion}
\label{sec:conclusion}

In this paper we elaborated a questionnaire to gauge students' perceptions and
measure their difficulty to learn \textit{OpenMP} and \textit{Pthreads}.  The
questionnaire contained questions that regarded each technology individually
and in comparison with the other.  The students were enrolled in the
Concurrent, Parallel and Distributed course from the University of São Paulo,
Brazil.

The students' responses to the questionnaire enabled conclude that the students
perceived less difficulty using and learning \textit{OpenMP}, in the context of
a course assignment, but that they also achieved greater performance
improvements using \textit{Pthreads}.

Contrary to our analysis, students reported the perception that using
\textit{OpenMP} made it easier to improve the performance of the sequential
code we provided.  Our analysis of their submissions revealed that the majority
of the students ($52.2\%$) achieved better performance in the assignment with
either \textit{OpenMP} ($17.4\%$) or \textit{Pthreads} ($34.8\%$).
Contradicting the perception of the students, the implementation using
\textit{Pthreads} achieved better performance in twice the submissions.

The perceptions of usability and performance improvement regarding
\textit{OpenMP} was not reflected in the actual implementations and
submissions.  We assert that \textit{OpenMP} is still not as easy as it appears
to be, and that the students' responses strengthen the argument for teaching
lower-level technologies for parallel and distributed computing.
The main contribution of this paper is the conclusion that students
are able to achieve better performance results using \textit{Pthreads}
than \textit{OpenMP}, despite their perception that \textit{Pthreads}
is harder to learn and use.

In future work we will compare the perception and performance of students
regarding learning and achieving performance improvements using different
languages and tools for parallel and distributed computing. We will also study
how attempting solving more difficult parallel and distributed problem impacts
the students' perceptions and performance.

### 2017-08-08 sections/02-apis.tex
Included missing reference

@@
 Listing \ref{lst:listing-pthreads} presents a sample code using the
-\textit{Pthreads} library. The program begins execution in a
-single thread and until line 6 only one thread is running.
+\textit{Pthreads} library. The code was taken from
+\cite{goncalves:OpenMPNotEasy} and shows a simple multiplication of two
+matrices element-wise and summation of this multiplication.
 
-When \texttt{pthread\_create} is called a thread is created and the application
-enters a parallel region with two simultaneous threads.  The main mechanism
-available for synchronizing thread executions is the barrier. The
-\textit{Pthreads} library implements barriers as calls to
+The program begins its execution in a single thread and until line 17 only one
+thread is running. When \texttt{pthread\_create} is called a thread is created
+and the application enters a parallel region with two simultaneous threads. As
+it happens inside a for loop, several threads are expected to be created.
+
+The strategy used for such parallelization is to devide the vectors in chunks
+and delegating the multiplication of the elements inside these chunks among
+threads, what can be seen in lines 14 and 15.
+
+The main mechanism available for synchronizing thread executions is the
+barrier. The \textit{Pthreads} library implements barriers as calls to
 \texttt{pthread\_join}. A thread that reaches a barrier will wait for other
-threads. Line 11 shows an example of thread synchronization using a barrier.
+threads. Line 21 shows an example of thread synchronization using a barrier.
+
+As the threads are supposed to write in a shared variable (\texttt{dot}), we
+have to explicitly take care of the concurrent access to the variable, as seen
+in Line 33 and 35, where the threads have to perform a lock at a mutex in order
+to enter the critical region.
 
@@
                    ]
-int main(){
-    pthread_t thread_id;
-    DATATYPE *parameters = 10;
-    int return_code;
-
-    return_code = pthread_create(&thread_id, NULL,
-                                 task_function,
-                                 (void *) parameters);
-
-    printf("Return Code: %d\n", return_code );
-    pthread_join(thread_id , NULL);
-
+int A[SIZE], B[SIZE]; int dot = 0;
+pthread_mutex_t lock ;
+int main(int argc, char *argv []){ 
+    int num_threads, i, part size ;
+    pthread_t tid [MAX_THREADS];
+    pthread_attr_t attr; 
+    pthread_attr_init(&attr);
+    /* Initialization vectors was supressed . */
+    part_size = (SIZE + num_threads - 1) / num_threads;
+    for (i=0; i < num_threads; i++){
+        struct arg_struct *args = 
+                malloc(sizeof(struct arg_struct));
+        args->id = i;
+        args->i_part = (i * part_size);
+        args->e_part = MIN(((i + 1)* (part_size)), 
+                           SIZE) ;
+        pthread_create(&(tid[i]) ,&attr, vecDot, 
+                       (void *) args) ;
+    }
+    for (i=0; i<num_threads; i++) { 
+        pthread_join(tid[i], NULL);
+    }
+    printf("Final result: %d.\n", dot); 
     return 0;
 }
-
-void *task_function(void *data){
-    DATATYPE *value;
-    value = (DATATYPE *) data;
+void *vecDot (void *arguments){
+    struct arg_struct *args = arguments;
+    int i, partial_dot=0;
+    for(i = args->i_part; i < args->e_part; i++){
+        partial_dot += A[i] * B[i]; 
+    }
+    /* Critical region . */ 
+    pthread_mutex_lock(&lock) ; 
+    dot += partial_dot;
+    pthread_mutex_unlock(&lock) ;
+    pthread_exit(0); 
 }

### 2017-08-09 sections/02-apis.tex
Minor revisions and spacing changes.

--- 2017-08-08 sections/02-apis.tex
@@
 creating and synchronizing threads. In this model, the execution begins in a
-sequential region and threads must be explicitly started to begin
-a parallel region.
-Listing \ref{lst:listing-pthreads} presents a sample code using the
-\textit{Pthreads} library. The program begins execution in a
-single thread and until line 6 only one thread is running.
+sequential region and threads must be explicitly started to begin a parallel
+region.  Listing~\ref{lst:pthreads} presents a \textit{Pthreads}
+example. The code shows the parallel computation of the dot product of two
+arrays.
 
-When \texttt{pthread\_create} is called a thread is created and the application
-enters a parallel region with two simultaneous threads.  The main mechanism
-available for synchronizing thread executions is the barrier. The
-\textit{Pthreads} library implements barriers as calls to
-\texttt{pthread\_join}. A thread that reaches a barrier will wait for other
-threads. Line 11 shows an example of thread synchronization using a barrier.
+The program in Listing~\ref{lst:pthreads} runs in a single thread until
+line 22, when \texttt{pthread\_create} is called.  Each call
+in the loop initializes a new thread.
+The parallelization in this example divides the arrays in chunks
+and shares multiplications between threads.
+One thread synchronization tool available in \textit{Pthreads} is the barrier,
+implemented as calls to \texttt{pthread\_join}. A thread that reaches a barrier
+waits for other threads. Line 27 shows an example of thread synchronization
+using a barrier.
+To avoid race conditions on the \texttt{dot} shared variable
+the example uses \textit{mutexes} to create a critical
+region, as seen in lines 42 and 44.
 
@@
                    xleftmargin=0.5cm,
-                   label=lst:listing-pthreads, keywords={%
+                   label=lst:pthreads, keywords={%
                        DATATYPE, pthread_t, pthread_create,
                        pthread_join, task_function, NULL, int, main,
-                       void, printf, return%
+                       void, printf, return, pthread_mutex_t,
+                       pthread_attr_t, pthread_attr_init,
+                       MAX_THREADS, SIZE, char, struct, malloc,
+                       MIN, pthread_mutex_lock, pthread_mutex_unlock,
+                       pthread_exit%
                        },
@@
                    ]
-int main(){
-    pthread_t thread_id;
-    DATATYPE *parameters = 10;
-    int return_code;
+int A[SIZE], B[SIZE]; int dot = 0;
+pthread_mutex_t lock;
 
-    return_code = pthread_create(&thread_id, NULL,
+int main(int argc, char *argv []){
+  int num_threads, i, part_size;
+  pthread_t tid [MAX_THREADS];
+  pthread_attr_t attr;
+  pthread_attr_init(&attr);
+  /* Array initialization supressed */
+
+  part_size = (SIZE + num_threads - 1) / num_threads;
+
+  for (i = 0; i < num_threads; i++){
+      struct arg_struct *args = \
+              malloc(sizeof(struct arg_struct));
+
+      args->id = i;
+      args->i_part = (i * part_size);
+      args->e_part = MIN(((i + 1) * (part_size)),
+                         SIZE);
+
+      pthread_create(&(tid[i]) ,&attr,
                      task_function,
-                                 (void *) parameters);
-
-    printf("Return Code: %d\n", return_code );
-    pthread_join(thread_id , NULL);
-
+                     (void *) args);
+  }
+  for (i = 0; i < num_threads; i++) {
+      pthread_join(tid[i], NULL);
+  }
+  printf("Result: %d.\n", dot);
   return 0;
@@
 
-void *task_function(void *data){
-    DATATYPE *value;
-    value = (DATATYPE *) data;
+void *task_function (void *arguments){
+  struct arg_struct *args = (arg_struct *) arguments;
+  int i, partial_dot = 0;
+
+  for(i = args->i_part; i < args->e_part; i++){
+      partial_dot += A[i] * B[i];
+  }
+
+  /* Critical region */
+  pthread_mutex_lock(&lock);
+  dot += partial_dot;
+  pthread_mutex_unlock(&lock);
+  pthread_exit(0);
 }
@@
 
-Listing \ref{lst:listing-omp} presents an \textit{OpenMP} example. The
+Listing \ref{lst:omp} presents an \textit{OpenMP} example for the
+same computation shown in Listing \ref{lst:pthreads}. The
 declaration of a parallel region with the \texttt{firstprivate} pragma on line
@@
 only one thread, because of the \texttt{single} pragma.
-The \texttt{for} in line 7 will be executed in parallel, with automatic work
+The loop in line 8 will be executed in parallel, with automatic work
 division.  The pragma at line 12 defines a critical region which can only be
@@
                    xleftmargin=0.6cm,
-                   label=lst:listing-omp, keywords={%
+                   label=lst:omp, keywords={%
                        \#pragma,

### 2017-08-09 sections/00-abstract.tex
Reviewing abstract

@@
 Selecting the most effective tools and languages for teaching parallel and
-    distributed computing presents a challenge for teachers and students. Often
-    a high-level solution seems easier, but it comes with the cost of hiding
-    underlying concepts and problems. We present a comparison between the
-    \textit{Pthreads} and \textit{OpenMP} APIs considering the experiences and
-    performance results of undergraduate and graduate Computer Science students
-    from the University of São Paulo.
+distributed computing presents a challenge for teachers and students. Often a
+high-level solution seems easier, but it comes with the cost of hiding
+underlying concepts and problems. In this paper we present a comparison between
+the \textit{Pthreads} and \textit{OpenMP} APIs considering the experiences and
+performance results of undergraduate and graduate Computer Science students.
 We show that students were able to achieve better performance results using

### 2017-08-09 sections/01-introduction.tex
Reviewing Introduction

@@
 what is needed for traditional sequential programming. Selecting an effective
-language, framework or API for parallel and distributed programming is a
+language, framework or API for parallel, and distributed programming is a
 challenge for teachers and students, especially due to the quantity and
@@
 students to develop working code. The cost of high-level abstractions often is
-the hiding of underlying concepts, problems and solutions, as well as code
+the hiding of underlying concepts, problems, and solutions, as well as code
 performance.

### 2017-08-09 sections/04-research-design.tex
Add text for the abstract, introduction and conclusion.
Reviewing Research Design

@@
 
-This section describes our research design. We devised a questionnaire
-regarding student experience with solving the same problem using
-\textit{Pthreads} and \textit{OpenMP} during the Concurrent, Parallel and
-Distributed Computing Course at the Computer Science program of the University
-of São Paulo. We answered 3 research questions using analyses of the students'
-responses.
+We devised a questionnaire regarding student experience with solving the same
+problem using \textit{Pthreads} and \textit{OpenMP} during the Concurrent,
+Parallel and Distributed Computing Course at the Computer Science program of
+the University of São Paulo. We answered 3 research questions using analyses of
+the students' responses.
 
@@
 sequential code and the performance improvements they achieved using
-\textit{Pthreads} and \textit{OpenMP}.  After completing the assignment the
-students were asked to answer a questionnaire.
+\textit{Pthreads} and \textit{OpenMP}. After completing the assignment we asked
+the students to answer a questionnaire.
 
@@
 The questions of the second part were related to the students'
-experience with \textit{Pthreads} and \textit{OpenMP} \textit{after}
+experience with \textit{Pthreads} and \textit{OpenMP} after
 the completion of the assignment.
@@
 54 students enrolled in the Concurrent, Parallel and Distributed Computing
-course. Section \ref{sec:background} presents a detailed description of
-each set of questions our analysis of the students' responses.
+course. Section \ref{sec:responses} presents a detailed description of each set
+of questions our analysis of the students' responses.
 
 \subsection{Research Questions}
+\label{sec:resques}
 

### 2017-08-09 sections/01-introduction.tex
More text Revisions.

--- 2017-08-08 sections/01-introduction.tex
@@
 Parallel and Distributed Computing of the University of São Paulo, Brazil.  We
-show that students were able to achieve better performance results using
-\textit{Pthreads} than \textit{OpenMP}, despite their perception that
-\textit{Pthreads} is harder to learn and use.
+show that students were able to achieve better performance results with
+\textit{Pthreads} than with \textit{OpenMP}, despite their perception
+that \textit{Pthreads} is harder to learn and use.
 
 The paper is organized as follows.
-Section \ref{sec:apis} describes \textit{OpenMP} and \textit{Pthreads}.
+Section \ref{sec:apis} briefly describes \textit{OpenMP} and \textit{Pthreads}.
 Section \ref{sec:background} discusses related work.
-Section \ref{sec:researchdesign} discusses the course of
+Section \ref{sec:researchdesign} presents the course of
 Concurrent, Parallel and Distribute Computing of the University
-of São Paulo, and presents the questionnaire and our research
+of São Paulo, and introduces the questionnaire and our research
 questions.

### 2017-08-09 sections/02-apis.tex
More text Revisions.

@@
 
-The C/C++ \textit{Pthreads} library controls thread execution in GNU/Linux by
+The \textit{Pthreads} library controls thread execution by
 creating and synchronizing threads. In this model, the execution begins in a
@@
 
+Note that the examples shown in Listings \ref{lst:pthreads} and
+\ref{lst:omp} solve the same problem. The example in \textit{Phthreads}
+contains much more code than the example in \textit{OpenMP}, and seems
+to be much more complex as well. It is therefore understandable that students might
+feel overwhelmed by \textit{Pthreads} example, especially when both examples are
+presented side-by-side.
+
 \begin{lstlisting}[language=C, basicstyle=\ttfamily\scriptsize, numbers=left,

### 2017-08-09 sections/04-research-design.tex
More text Revisions.

--- 2017-08-08 sections/04-research-design.tex
@@
 Distributed Computing Course at the Computer Science program of the University
-of São Paulo. We answered 3 research questions using analyses of the students'
+of São Paulo. We answered 3 research questions analysing the students'
 responses.
 
-\subsection{The Concurrent, Parallel and Distributed Computing Course}
+\subsection{The Course}
 
@@
 presentation on a relevant topic related to Concurrent, Parallel and
-Distributed Computing.
+Distributed Computing.  The course material is available
+online~\footnote{\url{https://phrb.github.io/MAC5742-0219} [Accessed in
+07/08/2017]}.
 
-We devised an assignment to measure the students' difficulties to
-learn and use \textit{Pthreads} and \textit{OpenMP} to solve
-a simple problem. Students had to implement parallel versions
-of a sequential code using \textit{Pthreads} and \textit{OpenMP}.
-We provided a sequential implementation for the calculation of
-the Mandelbrot Set. The sequential code presented an embarrassingly
-parallel problem with two nested loops, and a third loop with
+We devised an assignment to measure the students' difficulties to learn and use
+\textit{Pthreads} and \textit{OpenMP} to solve a simple problem. Students had
+to implement parallel versions of a sequential code using \textit{Pthreads} and
+\textit{OpenMP}.  We provided a sequential implementation for the calculation
+of the Mandelbrot Set~\cite{douady1984etude}. The sequential code presented an
+embarrassingly parallel problem with two nested loops, and a third loop with
 dependent iterations that could not be parallelized easily.
@@
 54 students enrolled in the Concurrent, Parallel and Distributed Computing
-course. Section \ref{sec:background} presents a detailed description of
-each set of questions our analysis of the students' responses.
+course. Section \ref{sec:responses} presents a detailed description of
+each set of questions and our analysis of the students' responses.
 
 \subsection{Research Questions}
+\label{sec:resques}
 
-In previous work~\cite{goncalves:OpenMPNotEasy} we showed that
-teaching \textit{OpenMP} requires teaching important parallel
-and distributed computing concepts first, since it is easy
-to find errors on commonly used \textit{OpenMP} learning material.
+In previous work~\cite{goncalves:OpenMPNotEasy} we showed that teaching
+\textit{OpenMP} requires the introduction of important parallel and distributed
+computing concepts first, since it is easy to find errors on commonly used
+\textit{OpenMP} learning material.
 

### 2017-08-09 sections/05-responses.tex
Change title and minor edits.
Minor revisions and spacing changes.
More text Revisions.

@@
 
-\todo[inline,color=cyan,author=Pedro]{Improve figure colorscheme}
+%\todo[inline,color=cyan,author=Pedro]{Improve figure colorscheme}
 
@@
 
-The course of Concurrent, Parallel and Distributed Computing is offered to
-students of both the graduate and undergraduate programs in Computer Science at
-the University of São Paulo. To gauge the proficiency and knowledge of our
-students we elaborated a five-point scale from \textit{Do not Know} $(1)$ to
-\textit{Expert or Proficient} $(5)$.  We asked the students to express their
-proficiency level on this scale, and their responses are shown in Figure
-\ref{fig:background}.
+To gauge the proficiency and knowledge of our students we elaborated a
+five-point scale from \textit{Do not Know} $(1)$ to \textit{Expert or
+Proficient} $(5)$.  We asked the students to express their proficiency level on
+this scale, and their responses are shown in Figure \ref{fig:background}.
 
@@
 questions and is also shown in Figure \ref{fig:background}. The students
-reported significantly larger proficiency in \textit{Pthreads} than
+reported significantly larger proficiency in \textit{Pthreads} than in
 \textit{OpenMP}, and the majority of the students reported they did not know
@@
     \label{tab:classes}
+    \vspace{-0.6cm}
 \end{table}
@@
 \begin{figure}[htpb]
+    \vspace{-0.7em}
     \centering
@@
     \label{fig:classes}
+    \vspace{-1em}
 \end{figure}
@@
     \label{tab:likert}
+    \vspace{-0.6em}
 \end{table}
@@
 \begin{figure}[htpb]
+    \vspace{-2em}
     \centering
@@
     \label{fig:likert}
+    \vspace{-1em}
 \end{figure}
@@
     \label{tab:comparisons}
+    \vspace{-0.6cm}
 \end{table}
@@
 \begin{figure}[htpb]
+    \vspace{-.7em}
     \centering
@@
     \label{fig:comparisons}
+    \vspace{-1em}
 \end{figure}

### 2017-08-09 sections/07-conclusion.tex
More text Revisions.

@@
 questionnaire contained questions that regarded each technology individually
-and in comparison with the other.  The students were enrolled in the
-Concurrent, Parallel and Distributed course from the University of São Paulo,
-Brazil.
+and in comparison with the other.
 
-The students' responses to the questionnaire enabled conclude that the students
-perceived less difficulty using and learning \textit{OpenMP}, in the context of
-a course assignment, but that they also achieved greater performance
+The students' responses to the questionnaire enable the conclusion that the
+students perceived less difficulty using and learning \textit{OpenMP}, in the
+context of a course assignment, but that they also achieved greater performance
 improvements using \textit{Pthreads}.
@@
 are able to achieve better performance results using \textit{Pthreads}
-than \textit{OpenMP}, despite their perception that \textit{Pthreads}
+than using \textit{OpenMP}, despite their perception that \textit{Pthreads}
 is harder to learn and use.
 
-In future work we will compare the perception and performance of students
-regarding learning and achieving performance improvements using different
-languages and tools for parallel and distributed computing. We will also study
-how attempting solving more difficult parallel and distributed problem impacts
-the students' perceptions and performance.
+Brown \textit{et al.}~\cite{Brown:2010:SPC:1971681.1971689} points that the
+choice of a software tool should be only one of the topics explored during a
+parallel programming course.  In future work we will compare the perception and
+performance of students regarding learning and achieving performance
+improvements using different languages and tools for parallel and distributed
+computing.  Since we only studied an embarrassing parallel problem which could
+give an initial advantage to \textit{OpenMP}, in future work we will explore
+how more difficult parallel and distributed problems impacts the students'
+perceptions and performance.

### 2017-08-09 sections/00-abstract.tex

@@
 the \textit{Pthreads} and \textit{OpenMP} APIs considering the experiences and
-performance results of undergraduate and graduate Computer Science students.
+performance results of undergraduate and graduate Computer Science students
+from the University of São Paulo.
 We show that students were able to achieve better performance results using

### 2017-09-08 rebuttal/rebuttal.md
Add rebuttal file

## Paper Reviews and Rebuttal

### Review Summary

| Paper Merit   | Reviewer Expertise |
| :------------ | :----------------- |
| Accept        | Knowledgeable      |
| Weak Reject   | Knowledgeable      |
| Strong Accept | Some Familiarity   |

### Reviewer Questions and Rebuttal

### Reviewer #9A

The following explicit questions need to be addressed:

1. What were the learning objectives of the course and research project?
2. Did you have ethical approval to do a study like this? You mention consent
   but not much more than that
3. You mention significant, but you don't have evidence to support this claim.
   You need to show statistical analysis for your data.
4. The graphs are hard to read. Consider using a different colour palette.
5. Need to add limitation or threats to validity for your research approach.
6. Do you have any comments on generalizing this kind of research approach to
   study other aspects of programming?

### Reviewer #9B

The following explicit questions need to be addressed:

#### Explicit Questions

##### Major Issues

1. The description of the programming task is somewhat vague and it is unclear
   if it went through any validation for appropriateness to measure students'
   ability. What was the exact task (the authors mention the Mandelbrot Set
   they provided as an example, but do not expand on their "simple problem").
   There is no way for the reader to determine how appropriate this task was
   without knowing the problem, so perhaps listing the question/problem itself
   would be best.
2. The reviewer assumes that the programming assignment and questionnaires were
   solo activities. However, this was not explicitly described. (Since the
   authors described this course as a place where undergraduate students and
   graduate students could share their knowledge).
3. How were the questions developed, and what measures did the authors take to
   make sure they were appropriate to measure what they wanted to measure?
4. Were the questions listed in Portuguese (the reviewer believes this might be
   an option as this study took place in Brazil) or English?
5. Some of the questions in Table 1 are vague and the students might interpret
   them in different ways. For example, what level of commitment/experience
   does "Had contact with the APIs before?" assume? What does "the API" mean?
   The same question goes for what "learned" means in the following two
   questions.
6. In Table 3, what is the difference between questions 2 and 3? "Was simpler
   to use" vs. "Was easier to use".
7. The reviewer does not believe that the major finding of this paper is
   accurately described and cannot agree with the author's statement in the
   conclusion that 52.2% of students performing better using Pthreads
   constitutes a "majority."

##### Minor Issues

1. Some of the statements in the introduction and the descriptions of OpenMP
   and Pthreads can use additional references (e.g., "slowdowns on Moore's law
   and the end of Dennard scaling").
2. Recommend changing "Listing X" to "Figure X".
3. Recommend displaying "Listing 1" in one column instead of splitting it in
   the middle.
4. Recommend using active voice instead of passive voice (e.g., "Different
   approaches have been proposed to study..."
5. Recommend not using colors for Figures since some readers might not be able
   to distinguish them (or someone might print it out in black-and-white).
6. Table 1. The second question has a typo "Had contat [sic] with the API///"

#### Other Comments

- Comment:
> The authors did answer their three research questions. However, the authors
> over-claim one of their major results (see #7, below), along with saying some
> of their results are "significant" without providing statistical measure.

### Reviewer #9C

#### Explicit Questions

1. Why do you call it a Likert-like scale?
2. How were the APIs introduced in the course?
3. What were the exercises?
4. Is it possible that differences in recognition depend on the presentation of
   the APIs or teaching material, etc.?

#### Other Comments

- Comment
> I personally think that a much larger discussion section is required that
> actually discusses the chosen method and the conclusions drawn from it.

- Comment:
> Since you have a Likert-Scale, you can apply non-parametric tests on it in
> order to check, whether group A or B is different. Descriptive data and raw
> measurement is fine, because it gives readers a change to get an idea about
> the data, but only relying on such things is not enough. Having said this, I
> do not think that you can conclude something from the present paper, because
> the conclusion is just based on "looking at descriptive data".

- Comment:
> Please fix the statistical part of the paper. Nothing would be more annoying
> than having a paper published whose interpretation depends on results that
> would be different in case appropriate analysis techniques would have been
> applied.

### 2017-09-08 rebuttal/rebuttal.md
Improve rebuttal file syntax

@@
 
-The following explicit questions need to be addressed:
+#### Explicit Questions
 
@@
 ### Reviewer #9B
-
-The following explicit questions need to be addressed:
 
@@
 
-- Comment:
-> The authors did answer their three research questions. However, the authors
-> over-claim one of their major results (see #7, below), along with saying some
-> of their results are "significant" without providing statistical measure.
+1. The authors did answer their three research questions. However, the authors
+   over-claim one of their major results (see #7, below), along with saying some
+   of their results are "significant" without providing statistical measure.
 
@@
 
-- Comment
-> I personally think that a much larger discussion section is required that
-> actually discusses the chosen method and the conclusions drawn from it.
-
-- Comment:
-> Since you have a Likert-Scale, you can apply non-parametric tests on it in
-> order to check, whether group A or B is different. Descriptive data and raw
-> measurement is fine, because it gives readers a change to get an idea about
-> the data, but only relying on such things is not enough. Having said this, I
-> do not think that you can conclude something from the present paper, because
-> the conclusion is just based on "looking at descriptive data".
-
-- Comment:
-> Please fix the statistical part of the paper. Nothing would be more annoying
-> than having a paper published whose interpretation depends on results that
-> would be different in case appropriate analysis techniques would have been
-> applied.
+1. I personally think that a much larger discussion section is required that
+  actually discusses the chosen method and the conclusions drawn from it.
+2. Since you have a Likert-Scale, you can apply non-parametric tests on it in
+   order to check, whether group A or B is different. Descriptive data and raw
+   measurement is fine, because it gives readers a change to get an idea about
+   the data, but only relying on such things is not enough. Having said this, I
+   do not think that you can conclude something from the present paper, because
+   the conclusion is just based on "looking at descriptive data".
+3. Please fix the statistical part of the paper. Nothing would be more annoying
+   than having a paper published whose interpretation depends on results that
+   would be different in case appropriate analysis techniques would have been
+   applied.

### 2017-09-21 rebuttal/rebuttal.md
Attributing tasks

@@
 
-1. What were the learning objectives of the course and research project?
+1. What were the learning objectives of the course and research project? (TODO: Alfredo)
 2. Did you have ethical approval to do a study like this? You mention consent
-   but not much more than that
+   but not much more than that (TODO: Paulo)
 3. You mention significant, but you don't have evidence to support this claim.
-   You need to show statistical analysis for your data.
-4. The graphs are hard to read. Consider using a different colour palette.
-5. Need to add limitation or threats to validity for your research approach.
+   You need to show statistical analysis for your data (TODO: Pedro)
+4. The graphs are hard to read. Consider using a different colour palette (TODO: Pedro)
+5. Need to add limitation or threats to validity for your research approach (TODO: Raphael)
 6. Do you have any comments on generalizing this kind of research approach to
-   study other aspects of programming?
+   study other aspects of programming? (TODO: Alfredo)
 
@@
    without knowing the problem, so perhaps listing the question/problem itself
-   would be best.
+   would be best. (TODO: Pedro)
 2. The reviewer assumes that the programming assignment and questionnaires were
@@
    authors described this course as a place where undergraduate students and
-   graduate students could share their knowledge).
+   graduate students could share their knowledge). (TODO: Alfredo)
 3. How were the questions developed, and what measures did the authors take to
-   make sure they were appropriate to measure what they wanted to measure?
+   make sure they were appropriate to measure what they wanted to measure? (DONE: ele pode ter razão, não temos o que fazer)
 4. Were the questions listed in Portuguese (the reviewer believes this might be
-   an option as this study took place in Brazil) or English?
+   an option as this study took place in Brazil) or English? (TODO: Pedro)
 5. Some of the questions in Table 1 are vague and the students might interpret
@@
    The same question goes for what "learned" means in the following two
-   questions.
+   questions.(DONE: ele pode ter razão, não temos o que fazer)
 6. In Table 3, what is the difference between questions 2 and 3? "Was simpler
-   to use" vs. "Was easier to use".
+   to use" vs. "Was easier to use". (TODO: Pedro)
 7. The reviewer does not believe that the major finding of this paper is
@@
    conclusion that 52.2% of students performing better using Pthreads
-   constitutes a "majority."
+   constitutes a "majority." (TODO: Pedro, mas acho que se pensarmos juntos mitigaremos)
 
@@
    and Pthreads can use additional references (e.g., "slowdowns on Moore's law
-   and the end of Dennard scaling").
-2. Recommend changing "Listing X" to "Figure X".
+   and the end of Dennard scaling"). (TODO: Raphael)
+2. Recommend changing "Listing X" to "Figure X". (TODO: Paulo)
 3. Recommend displaying "Listing 1" in one column instead of splitting it in
-   the middle.
+   the middle. (TODO: Paulo)
 4. Recommend using active voice instead of passive voice (e.g., "Different
-   approaches have been proposed to study..."
+   approaches have been proposed to study..." (TODO: Paulo)
 5. Recommend not using colors for Figures since some readers might not be able
-   to distinguish them (or someone might print it out in black-and-white).
-6. Table 1. The second question has a typo "Had contat [sic] with the API///"
+   to distinguish them (or someone might print it out in black-and-white). (TODO: Pedro, relacionado com item 4 do revisão 1)
+6. Table 1. The second question has a typo "Had contat [sic] with the API///" (TODO: Paulo)
 
@@
    over-claim one of their major results (see #7, below), along with saying some
-   of their results are "significant" without providing statistical measure.
+   of their results are "significant" without providing statistical measure. (TODO: Pedro, mas acho que se pensarmos juntos mitigaremos)
 
@@
 
-1. Why do you call it a Likert-like scale?
-2. How were the APIs introduced in the course?
-3. What were the exercises?
+1. Why do you call it a Likert-like scale? (TODO: Paulo)
+2. How were the APIs introduced in the course? (TODO: Alfredo)
+3. What were the exercises? (TODO: Pedro, mas do item 1 do revisor 2)
 4. Is it possible that differences in recognition depend on the presentation of
-   the APIs or teaching material, etc.?
+   the APIs or teaching material, etc.? (TODO: Raphael -- pode-se colocar nas limitações do trabalho, relacionado ao item 5 do revisor 1)
 
@@
 1. I personally think that a much larger discussion section is required that
-  actually discusses the chosen method and the conclusions drawn from it.
+  actually discusses the chosen method and the conclusions drawn from it. (TODO: Pedro)
 2. Since you have a Likert-Scale, you can apply non-parametric tests on it in
@@
    do not think that you can conclude something from the present paper, because
-   the conclusion is just based on "looking at descriptive data".
+   the conclusion is just based on "looking at descriptive data". (TODO: Paulo)
 3. Please fix the statistical part of the paper. Nothing would be more annoying
@@
    would be different in case appropriate analysis techniques would have been
-   applied.
+   applied. (TODO: Pedro, mas acho que se pensarmos juntos mitigaremos)

### 2017-10-11 sections/02-apis.tex
Listing to Figure

@@
 sequential region and threads must be explicitly started to begin a parallel
-region.  Listing~\ref{lst:pthreads} presents a \textit{Pthreads}
+region.  Figure~\ref{lst:pthreads} presents a \textit{Pthreads}
 example. The code shows the parallel computation of the dot product of two
@@
 
-The program in Listing~\ref{lst:pthreads} runs in a single thread until
+The program in Figure~\ref{lst:pthreads} runs in a single thread until
 line 22, when \texttt{pthread\_create} is called.  Each call
@@
 
+\begin{figure}[htb]
+\begin{minipage}{\linewidth}
 \begin{lstlisting}[language=C, basicstyle=\ttfamily\scriptsize, numbers=left,
                    frame=no, showspaces=false, showstringspaces=false,
-                   caption={\textit{Pthreads} Sample Program}, captionpos=b,
                    numberstyle=\tiny,
-                   xleftmargin=0.5cm,
-                   label=lst:pthreads, keywords={%
+                   xleftmargin=0.5cm, frame=trBL,
+                   keywords={%
                        DATATYPE, pthread_t, pthread_create,
@@
 \end{lstlisting}
+\end{minipage}
+\caption{\textit{Pthreads} Sample Program}
+\label{lst:pthreads}
+\end{figure}
 
@@
 
-Listing \ref{lst:omp} presents an \textit{OpenMP} example for the
-same computation shown in Listing \ref{lst:pthreads}. The
+Figure \ref{lst:omp} presents an \textit{OpenMP} example for the
+same computation shown in Figure \ref{lst:pthreads}. The
 declaration of a parallel region with the \texttt{firstprivate} pragma on line
@@
 
-Note that the examples shown in Listings \ref{lst:pthreads} and
+Note that the examples shown in Figures \ref{lst:pthreads} and
 \ref{lst:omp} solve the same problem. The example in \textit{Phthreads}
@@
 
+\begin{figure}[htb]
+\begin{minipage}{\linewidth}
 \begin{lstlisting}[language=C, basicstyle=\ttfamily\scriptsize, numbers=left,
                    frame=no, showspaces=false, showstringspaces=false,
-                   caption={\textit{OpenMP} Sample Program}, captionpos=b,
                    numberstyle=\tiny,
-                   xleftmargin=0.6cm,
-                   label=lst:omp, keywords={%
+                   xleftmargin=0.6cm, frame=trBL,
+                   keywords={%
                        \#pragma,
@@
 \end{lstlisting}
+\end{minipage}
+\caption{\textit{OpenMP} Sample Program}
+\label{lst:omp}
+\end{figure}

### 2017-10-11 sections/04-research-design.tex
Add sentence on CUDA and OpenMPI.
Expliciting ethical approval

@@
 The course presents students with a series of software engineering challenges
-in the form of assignments. The final is composed of an essay and a
-presentation on a relevant topic related to Concurrent, Parallel and
-Distributed Computing.  The course material is available
+in the form of assignments involving GPU programming with CUDA and distributed
+computing with OpenMPI. The final is composed of an essay and a presentation on
+a relevant topic related to Concurrent, Parallel and Distributed Computing.
+The course material is available
 online~\footnote{\url{https://phrb.github.io/MAC5742-0219} [Accessed in
@@
 
-We obtained responses, and consent to use them in our research, from 38 of the
-54 students enrolled in the Concurrent, Parallel and Distributed Computing
-course. Section \ref{sec:responses} presents a detailed description of
-each set of questions and our analysis of the students' responses.
+We obtained the ethical approval from the Department of Computer Science at the
+University of São Paulo to conduct this study based on the questionnaire. We
+collected responses, and consent to use them in our research, from 38 of the 54
+students enrolled in the Concurrent, Parallel and Distributed Computing course.
+Section \ref{sec:responses} presents a detailed description of each set of
+questions and our analysis of the students' responses.
 

### 2017-10-11 sections/03-background.tex
More text Revisions.
active voice

@@
 
-Different approaches have been proposed to study the selection of the best level of
-abstraction for teaching Parallel and Distributed programming courses.
+Several authors proposed different approaches to study the selection of the best
+level of abstraction for teaching Parallel and Distributed programming courses.
 Falcao~\cite{6565518} argues that introducing a high-level parallel programming
@@
 programming interfaces, such as \textit{Pthreads}, as well as high-level
-interfaces, such as \textit{OpenMP}.  Brown \textit{et
-al.}~\cite{Brown:2010:SPC:1971681.1971689}, points that the choice of a
-software tool should be only one of the topics explored during a parallel
-programming course. Topics such as data structures, algorithms, software design
-and parallel hardware platforms should also be explored.
+interfaces, such as \textit{OpenMP}.

### 2017-10-11 sections/05-responses.tex
Change author order, make sure tables are next to figures.
Fixing typo
Removing 'Likert-like' and using just 'scale'

@@
 
-We asked the students to evaluate their proficiency in \textit{OpenMP} and
-\textit{Pthreads}, \textit{before} completing the assignment, with two sets of
-questions. The first set used the same five-point scale as the other background
-questions and is also shown in Figure \ref{fig:background}. The students
-reported significantly larger proficiency in \textit{Pthreads} than in
-\textit{OpenMP}, and the majority of the students reported they did not know
-\textit{OpenMP} before the course.
-
 \begin{table}[htpb]
@@
         \scriptsize{Had contact with parallel and distributed concepts before?} & $(1)$ \\
-        \scriptsize{Had contat with the APIs before?} & $(2)$ \\
+        \scriptsize{Had contact with the APIs before?} & $(2)$ \\
         \scriptsize{Learned from the OpenMP class?} & $(3)$ \\
@@
 
+We asked the students to evaluate their proficiency in \textit{OpenMP} and
+\textit{Pthreads}, \textit{before} completing the assignment, with two sets of
+questions. The first set used the same five-point scale as the other background
+questions and is also shown in Figure \ref{fig:background}. The students
+reported significantly larger proficiency in \textit{Pthreads} than in
+\textit{OpenMP}, and the majority of the students reported they did not know
+\textit{OpenMP} before the course.
+
 The second set of questions regarding student proficiency with the APIs
@@
 completing the assignment.
-The first set of questions used a Likert-type Scale to measure the relationship
-of the students with the two APIs in the context of the assignment. The Likert
-Scale~\cite{likert1932technique} enables respondents to express their agreement
-or disagreement with specific statements. Respondents choose between five
+The first set of questions used a scale defined by us with 5 levels to measure the relationship
+of the students with the two APIs in the context of the assignment. This kind of scale enables respondents to express their agreement
+or disagreement with specific statements~\cite{likert1932technique}. Respondents choose between five
 options: \textit{Strongly Agree} (SA), \textit{Agree} (A), \textit{Neutral}
@@
 \textit{OpenMP} and \textit{Pthreads}, in a total of $6$ questions.
+
+Figure \ref{fig:likert} shows a visualization of the students' responses.  The
+height of each stacked sub-column represents the percentage of respondents that
+selected each of the correspondent scale agreement levels, which are
+represented by different colors.
+
+The percentages of \textit{Strongly Agree} (SA) and \textit{Agree} (A)
+responses are significantly larger for questions $(1)$, $(3)$ and $(5)$, which
+are related to \textit{OpenMP}.  Likewise, the percentages of \textit{Disagree}
+(D) responses are significantly larger for questions $(2)$, $(4)$ and $(6)$,
+which are related to \textit{Pthreads}. These responses lead to the conclusion
+that \textbf{the perception of the students was that it was easier to use
+\textit{OpenMP} than \textit{Pthreads}} to parallelize and improve the
+performance of the sequential code provided in the assignment.
 
@@
     \end{tabular}
-    \caption{Likert-like Scale questions for Figure \ref{fig:likert}}
+    \caption{Scale questions for Figure \ref{fig:likert}}
     \label{tab:likert}
@@
     \includegraphics[width=0.85\columnwidth]{likert_questions}
-    \caption{Students' responses to the Likert Scale questions in Table \ref{tab:likert}}
+    \caption{Students' responses to the scale questions in Table \ref{tab:likert}}
     \label{fig:likert}
@@
 \end{figure}
-
-Figure \ref{fig:likert} shows a visualization of the students' responses.  The
-height of each stacked sub-column represents the percentage of respondents that
-selected each of the correspondent Likert scale agreement levels, which are
-represented by different colors.
-
-The percentages of \textit{Strongly Agree} (SA) and \textit{Agree} (A)
-responses are significantly larger for questions $(1)$, $(3)$ and $(5)$, which
-are related to \textit{OpenMP}.  Likewise, the percentages of \textit{Disagree}
-(D) responses are significantly larger for questions $(2)$, $(4)$ and $(6)$,
-which are related to \textit{Pthreads}. These responses lead to the conclusion
-that \textbf{the perception of the students was that it was easier to use
-\textit{OpenMP} than \textit{Pthreads}} to parallelize and improve the
-performance of the sequential code provided in the assignment.
 

### 2017-10-11 rebuttal/rebuttal.md
Updating TO-DO list

@@
 2. Did you have ethical approval to do a study like this? You mention consent
-   but not much more than that (TODO: Paulo)
+   but not much more than that (DONE)
 3. You mention significant, but you don't have evidence to support this claim.
@@
 3. How were the questions developed, and what measures did the authors take to
-   make sure they were appropriate to measure what they wanted to measure? (DONE: ele pode ter razão, não temos o que fazer)
+   make sure they were appropriate to measure what they wanted to measure? (DONE)
 4. Were the questions listed in Portuguese (the reviewer believes this might be
@@
    The same question goes for what "learned" means in the following two
-   questions.(DONE: ele pode ter razão, não temos o que fazer)
+   questions.(DONE)
 6. In Table 3, what is the difference between questions 2 and 3? "Was simpler
@@
    and the end of Dennard scaling"). (TODO: Raphael)
-2. Recommend changing "Listing X" to "Figure X". (TODO: Paulo)
+2. Recommend changing "Listing X" to "Figure X". (DONE)
 3. Recommend displaying "Listing 1" in one column instead of splitting it in
-   the middle. (TODO: Paulo)
+   the middle. (DONE)
 4. Recommend using active voice instead of passive voice (e.g., "Different
-   approaches have been proposed to study..." (TODO: Paulo)
+   approaches have been proposed to study..." (DONE)
 5. Recommend not using colors for Figures since some readers might not be able
    to distinguish them (or someone might print it out in black-and-white). (TODO: Pedro, relacionado com item 4 do revisão 1)
-6. Table 1. The second question has a typo "Had contat [sic] with the API///" (TODO: Paulo)
+6. Table 1. The second question has a typo "Had contat [sic] with the API///" (DONE)
 
@@
 
-1. Why do you call it a Likert-like scale? (TODO: Paulo)
+1. Why do you call it a Likert-like scale? (DONE)
 2. How were the APIs introduced in the course? (TODO: Alfredo)

### 2017-10-16 sections/01-introduction.tex
Added reference requested

@@
 Computing in the last decade, their relevance increased by recent slowdowns on
-Moore's law and the end of Dennard scaling.  Selecting the appropriate tool for
-programming multi-core architectures is also essential, since it is necessary
-to program a multi-core architecture effectively to leverage its performance.
+Moore's law and the end of Dennard scaling \cite{esmaeilzadeh2012dark}.
+Selecting the appropriate tool for programming multi-core architectures is also
+essential, since it is necessary to program a multi-core architecture
+effectively to leverage its performance.
 
@@
 show that students were able to achieve better performance results with
-\textit{Pthreads} than with \textit{OpenMP}, despite their perception
-that \textit{Pthreads} is harder to learn and use.
+\textit{Pthreads} than with \textit{OpenMP}, despite their perception that
+\textit{Pthreads} is harder to learn and use.
 
-The paper is organized as follows.
-Section \ref{sec:apis} briefly describes \textit{OpenMP} and \textit{Pthreads}.
-Section \ref{sec:background} discusses related work.
-Section \ref{sec:researchdesign} presents the course of
-Concurrent, Parallel and Distribute Computing of the University
-of São Paulo, and introduces the questionnaire and our research
-questions.
-Section \ref{sec:responses} presents, discusses and analyses
-the students' responses to the questionnaire.
-Section \ref{sec:conclusion} summarizes the results and
+The paper is organized as follows.  Section \ref{sec:apis} briefly describes
+\textit{OpenMP} and \textit{Pthreads}.  Section \ref{sec:background} discusses
+related work.  Section \ref{sec:researchdesign} presents the course of
+Concurrent, Parallel and Distribute Computing of the University of São Paulo,
+and introduces the questionnaire and our research questions.  Section
+\ref{sec:responses} presents, discusses and analyses the students' responses to
+the questionnaire.  Section \ref{sec:conclusion} summarizes the results and
 presents future work.

### 2017-10-16 sections/07-conclusion.tex
Added paragraph discussing some threats to validity aspects

@@
 \textit{OpenMP} made it easier to improve the performance of the sequential
-code we provided.  Our analysis of their submissions revealed that the majority
+code we provided.  Our analysis of their submissions revealed that most 
 of the students ($52.2\%$) achieved better performance in the assignment with
@@
 
+In our study we have not compared the material used for teaching both
+technologies in depth. In particular, we have not made any comparison regarding
+the difficulty associated with the notes used for teaching both technologies.
+Thus, on the matter of threats to the study validity, the students' perceptions
+of the difficulty level of the references, used during the class, were not
+taken into account. Also, in order to provide a more in-depth statistical
+analysis, the study needs to be performed in several classes of the Concurrent,
+Parallel and Distributed Computing courses.  Nevertheless, this pilot study
+presents some great insights that will help devise statistical hypothesis that
+should be tested in later studies.
+
 Brown \textit{et al.}~\cite{Brown:2010:SPC:1971681.1971689} points that the

### 2017-10-16 sections/02-apis.tex
Add mandelbrot explanation and pseudocode.

@@
                    numberstyle=\tiny,
-                   xleftmargin=0.5cm, frame=trBL,
+                   xleftmargin=0.5cm,
                    keywords={%
@@
                    numberstyle=\tiny,
-                   xleftmargin=0.6cm, frame=trBL,
+                   xleftmargin=0.6cm,
                    keywords={%

### 2017-10-16 sections/04-research-design.tex
Add mandelbrot explanation and pseudocode.

@@
 dependent iterations that could not be parallelized easily.
+
+\begin{figure}[htb]
+\begin{minipage}{\linewidth}
+\begin{lstlisting}[language=C, basicstyle=\ttfamily\scriptsize, numbers=left,
+                   frame=no, showspaces=false, showstringspaces=false,
+                   numberstyle=\tiny,
+                   xleftmargin=0.5cm,
+                   keywords={%
+                       DATATYPE, pthread_t, pthread_create,
+                       pthread_join, task_function, NULL, int, main,
+                       void, printf, return, pthread_mutex_t,
+                       pthread_attr_t, pthread_attr_init,
+                       MAX_THREADS, SIZE, char, struct, malloc,
+                       MIN, pthread_mutex_lock, pthread_mutex_unlock,
+                       pthread_exit, from, to, and%
+                       },
+                   otherkeywords={::, \#pragma, \#include, <<<,>>>, \&, \*, +, -, /, [, ], >, <}
+                   ]
+z = 0;
+for(y from 0 to y_max - 1) {
+    /* Indepedent iterations */
+    for(x from 0 to x_max - 1) {
+        /* Independet iterations */
+        for (iteration from 0 to iteration_max,
+             and f_c(z) < limit) {
+            /* Dependent iterations */
+            z = f_c(z);
+        }
+    }
+}
+\end{lstlisting}
+\end{minipage}
+\caption{Pseudocode for the computation of the Mandelbrot set}
+\label{lst:mandelbrot-pseudo}
+\end{figure}
+
+The Mandelbrot set is informally defined as the set of complex numbers $c$ for
+which the function $f_c(z) = z^2 + x$ does not diverge when iterated starting
+in $z=0$, that is, the sequence the sequence $f_c(0), f_c(f_c(0)),
+f_c(f_c(f_c(0))),\dots$ is always limited. Figure~\ref{lst:mandelbrot-pseudo}
+shows pseudocode for the computation of the Mandelbrot set, listing which loops
+have dependent and independent iterations.
 

### 2017-10-16 sections/04-research-design.tex
Improve pseudocode.

--- 2017-10-11 sections/04-research-design.tex
@@
 dependent iterations that could not be parallelized easily.
+
+\begin{figure}[htb]
+\begin{minipage}{\linewidth}
+\begin{lstlisting}[language=C, basicstyle=\ttfamily\scriptsize, numbers=left,
+                   frame=no, showspaces=false, showstringspaces=false,
+                   numberstyle=\tiny,
+                   xleftmargin=0.5cm,
+                   keywords={%
+                       DATATYPE, pthread_t, pthread_create,
+                       pthread_join, task_function, NULL, int, main,
+                       void, printf, return, pthread_mutex_t,
+                       pthread_attr_t, pthread_attr_init,
+                       MAX_THREADS, SIZE, char, struct, malloc,
+                       MIN, pthread_mutex_lock, pthread_mutex_unlock,
+                       pthread_exit, from, to, and%
+                       },
+                   otherkeywords={::, \#pragma, \#include, <<<,>>>, \&, \*, +, -, /, [, ], >, <}
+                   ]
+z = 0;
+for(x from 0 to x_max - 1) {
+    /* Independent iterations */
+    for(y from 0 to y_max - 1) {
+        /* Independent iterations */
+        for (iteration from 0 to iteration_max,
+             and f_c(z) < limit) {
+            /* Iterations depend on
+               previous values of z */
+            z = f_c(z);
+        }
+    }
+}
+\end{lstlisting}
+\end{minipage}
+\caption{Pseudocode for the computation of the Mandelbrot set}
+\label{lst:mandelbrot-pseudo}
+\end{figure}
+
+The Mandelbrot set is informally defined as the set of complex numbers $c$ for
+which the function $f_c(z) = z^2 + x$ does not diverge when iterated starting
+in $z=0$, that is, the sequence the sequence $f_c(0), f_c(f_c(0)),
+f_c(f_c(f_c(0))),\dots$ is always limited. Figure~\ref{lst:mandelbrot-pseudo}
+shows pseudocode for the computation of the Mandelbrot set, listing which loops
+have dependent and independent iterations.
 

### 2017-10-16 sections/05-responses.tex
Change colors to grayscale.

@@
 questions and is also shown in Figure \ref{fig:background}. The students
-reported significantly larger proficiency in \textit{Pthreads} than in
+reported larger proficiency in \textit{Pthreads} than in
 \textit{OpenMP}, and the majority of the students reported they did not know
@@
 The percentages of \textit{Strongly Agree} (SA) and \textit{Agree} (A)
-responses are significantly larger for questions $(1)$, $(3)$ and $(5)$, which
+responses are larger for questions $(1)$, $(3)$ and $(5)$, which
 are related to \textit{OpenMP}.  Likewise, the percentages of \textit{Disagree}
-(D) responses are significantly larger for questions $(2)$, $(4)$ and $(6)$,
+(D) responses are larger for questions $(2)$, $(4)$ and $(6)$,
 which are related to \textit{Pthreads}. These responses lead to the conclusion
@@
 The percentages of \textit{No Difficulty} $(1)$ and difficulty level $(2)$
-responses are significantly larger for learning \textit{OpenMP}. Likewise, the
+responses are larger for learning \textit{OpenMP}. Likewise, the
 percentages of difficulty level $(4)$ and \textit{Much Difficulty} $(5)$
-responses are significantly larger for learning \textit{Pthreads}.  We derive
+responses are larger for learning \textit{Pthreads}.  We derive
 from these responses the conclusion that \textbf{the students perceived less
@@
 
-The percentage of \textit{OpenMP} selections is significantly larger for all
+The percentage of \textit{OpenMP} selections is larger for all
 questions, except for questions $(7)$ and $(13)$. The responses for questions

### 2017-10-16 data/plot.py

@@ -60,8 +60,9 @@ def plot_bar_stacks(data,
     c_index = 0
 
     for key, value in data.items():
-        rect = ax.bar(indexes, value, width, color = cmap(1. * (c_index / colors)), edgecolor = 'lightgrey', bottom = previous_bottom)
+        rect = ax.bar(indexes, value, width, color = cmap(float(c_index) / float(colors)), edgecolor = 'lightgrey', bottom = previous_bottom)
 
+        print(float(c_index) / float(colors))
         c_index += 1
 
         previous_bottom += value
@@ -286,7 +287,7 @@ if __name__ == '__main__':
                     "",
                     0,
                     100,
-                    "viridis",
+                    "gray",
                     xticks_rotation = '0',
                     align = 'center')
 
@@ -357,7 +358,7 @@ if __name__ == '__main__':
                     "",
                     0,
                     100,
-                    "viridis",
+                    "gray",
                     xticks_rotation = '0',
                     align = 'center')
 
@@ -432,7 +433,7 @@ if __name__ == '__main__':
                     "",
                     0,
                     100,
-                    "viridis",
+                    "gray",
                     xticks_rotation = '0',
                     align = 'center')
 
@@ -483,7 +484,7 @@ if __name__ == '__main__':
                     "",
                     0,
                     100,
-                    "viridis",
+                    "gray",
                     xticks_rotation = '0',
                     align = 'center')
 
@@ -532,7 +533,7 @@ if __name__ == '__main__':
                     "",
                     0,
                     100,
-                    "viridis",
+                    "gray",
                     xticks_rotation = '0',
                     align = 'center')
 

### 2017-10-16 sections/04-research-design.tex
Edits and changes to make the camera ready version.

@@
 \textit{Pthreads} and \textit{OpenMP} to solve a simple problem. Students had
-to implement parallel versions of a sequential code using \textit{Pthreads} and
-\textit{OpenMP}.  We provided a sequential implementation for the calculation
-of the Mandelbrot Set~\cite{douady1984etude}. The sequential code presented an
-embarrassingly parallel problem with two nested loops, and a third loop with
-dependent iterations that could not be parallelized easily.
+to form groups of 2 to 3 members and implement parallel versions of a
+sequential code using \textit{Pthreads} and \textit{OpenMP}.  We provided a
+sequential implementation for the calculation of the Mandelbrot
+Set~\cite{douady1984etude}. The sequential code presented an embarrassingly
+parallel problem with two nested loops, and a third loop with dependent
+iterations that could not be parallelized easily.
 
@@
 
-The questionnaire had two parts. The first part asked questions
-about the students' previous experiences with common programming languages and
-parallel and distributed programming concepts and tools. We were also
-interested in students' perception of the impact of our classes on
-\textit{Pthreads} and \textit{OpenMP} on their learning difficulty.
+After students organized themselves in groups of 2 to 3 students and designed
+and implemented their parallel versions of the code, each individual student
+was asked to answer a questionnaire. The questionnaire had two parts. The first
+part asked questions about the students' previous experiences with common
+programming languages and parallel and distributed programming concepts and
+tools. We were also interested in students' perception of the impact of our
+classes on \textit{Pthreads} and \textit{OpenMP} on their learning difficulty.
 

### 2017-10-16 sections/05-responses.tex
Edits and changes to make the camera ready version.

@@
 
-We asked the students to answer 3 sets of questions regarding their
-experience with \textit{OpenMP} and \textit{Pthreads}, \textit{after}
-completing the assignment.
-The first set of questions used a scale defined by us with 5 levels to measure the relationship
-of the students with the two APIs in the context of the assignment. This kind of scale enables respondents to express their agreement
-or disagreement with specific statements~\cite{likert1932technique}. Respondents choose between five
-options: \textit{Strongly Agree} (SA), \textit{Agree} (A), \textit{Neutral}
-(N), \textit{Disagree} (D) and \textit{Strongly Disagree}.
+We asked the students to answer 3 sets of questions regarding their experience
+with \textit{OpenMP} and \textit{Pthreads}, \textit{after} completing the
+assignment.  The first set of questions used a Likert
+scale~\cite{likert1932technique} to measure the relationship of the students
+with the two APIs in the context of the assignment. The Likert scale enables
+respondents to express their agreement or disagreement with specific
+statements. Respondents choose between five options: \textit{Strongly Agree}
+(SA), \textit{Agree} (A), \textit{Neutral} (N), \textit{Disagree} (D) and
+\textit{Strongly Disagree}.
 
@@
     \end{tabular}
-    \caption{Scale questions for Figure \ref{fig:likert}}
+    \caption{Likert scale questions for Figure \ref{fig:likert}}
     \label{tab:likert}
@@
     \includegraphics[width=0.85\columnwidth]{likert_questions}
-    \caption{Students' responses to the scale questions in Table \ref{tab:likert}}
+    \caption{Students' responses to the Likert scale questions in Table \ref{tab:likert}}
     \label{fig:likert}

### 2017-10-16 sections/07-conclusion.tex
Edits and changes to make the camera ready version.

--- 2017-08-09 sections/07-conclusion.tex
@@
 \textit{OpenMP} made it easier to improve the performance of the sequential
-code we provided.  Our analysis of their submissions revealed that the majority
+code we provided.  Our analysis of their submissions revealed that most
 of the students ($52.2\%$) achieved better performance in the assignment with
@@
 
+Threats to the validity of this study include it being performed in a single
+class of the Concurrent, Parallel and Distributed Computing course.  A
+statistically relevant change to this study would be collecting data on
+multiple instances of the course over a period of multiple semesters. The
+subjectivity of the questionnaire and of the experience of the students is also
+a threat to validity. Additionally, we did not control for the order in which
+students performed their implementations in \textit{Pthreads} and
+\textit{OpenMP}, and we did not had separate groups learning each technology.
+
 Brown \textit{et al.}~\cite{Brown:2010:SPC:1971681.1971689} points that the
